---
title: "Lecture 11"
subtitle: "Association Rule Mining"
format:
  revealjs:
    theme: [default, custom_quarto.css]
    slide-number: true
    progress: true
    hash: true
    navigation-mode: linear
    transition: fade
    width: 1600
    height: 900
    fig-retina: 2
    code-line-numbers: false
    highlight-style: github
    self-contained: true
resources:
  - widgets
execute:
  echo: true
  warning: false
  message: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(arules)
library(arulesViz)
knitr::opts_chunk$set(
  echo = TRUE,
  fig.dim = c(4.8, 4.5),
  fig.retina = 2,
  out.width = "100%",
  comment = "#>"
)
theme_set(theme_minimal(base_size = 14))
```

## What are Association Rules?

Discovering **interesting relationships** among items in large collections of transactions.

:::: {.columns}
::: {.column width="50%" .fragment}
**Classic framing**: market basket analysis

- Supermarket receipts
- Online shopping carts

**But also**:

- Recipe ingredients 🍕
- Medical symptoms → diagnoses
- Web page visits → next click
- Survey question response patterns
:::
::: {.column width="50%" .fragment}
**The question**:

> *"If a recipe uses garlic and olive oil, what other ingredients does it tend to include?"*

An **association rule** has the form:

$$\{garlic, olive\_oil\} \Rightarrow \{parmesan\}$$

- **Antecedent** (left-hand side): condition
- **Consequent** (right-hand side): associated item
:::
::::

---

## Transactions as a Binary Matrix

Each **transaction** is a set of items. We represent the data as a binary matrix:

- Rows = transactions (recipes, shopping trips, patients)
- Columns = items
- Cell = 1 if item present, 0 otherwise

. . .

::: {.callout-tip}
## Key insight
Association rules work on **presence/absence** data — no quantities, no order. Just "was it in the basket or not?"
:::

. . .

Let's build one interactively...

---

## Widget 1: Transaction Builder {.scrollable}

Toggle pizza toppings on and off to build a transaction database. Watch the binary matrix and item frequencies update in real time.

```{ojs}
//| echo: false
import { transactionBuilderWidget } from "./widgets/association-rules/transaction-builder-widget.js"
viewof pizzaToggles = transactionBuilderWidget()
```

---

## Support, Confidence, and Lift

The three key metrics for evaluating association rules:

::: {style="font-size: 85%;"}
:::: {.columns}
::: {.column width="33%" .fragment}
### Support
How **frequently** the itemset appears:

$$\text{supp}(A \Rightarrow B) = \frac{|A \cup B|}{N}$$

*"How common is this combination?"*
:::
::: {.column width="33%" .fragment}
### Confidence
How **reliable** the rule is:

$$\text{conf}(A \Rightarrow B) = \frac{\text{supp}(A \cup B)}{\text{supp}(A)}$$

*"Given A, how often does B also appear?"*

This is $P(B | A)$ — **conditional probability**.
:::
::: {.column width="34%" .fragment}
### Lift
How much **more** likely B is given A than expected by chance:

$$\text{lift}(A \Rightarrow B) = \frac{\text{supp}(A \cup B)}{\text{supp}(A) \cdot \text{supp}(B)}$$

- Lift $> 1$: positive association ✓
- Lift $= 1$: independent (no association)
- Lift $< 1$: negative association
:::
::::
:::

---

## Widget 2: Metric Calculator {.scrollable}

Select items for the antecedent (A) and consequent (B) from the pizza data above, and see support, confidence, and lift computed in real time with a Venn diagram.

```{ojs}
//| echo: false
import { metricCalculatorWidget } from "./widgets/association-rules/metric-calculator-widget.js"
viewof metricInputs = metricCalculatorWidget(pizzaToggles)
```

---

## The Pitfall of High Confidence

::: {.callout-warning}
## High confidence ≠ interesting rule
A rule can have **high confidence** simply because the consequent is extremely common — not because of any real association.
:::

. . .

**Example from our pizza data** — pepperoni appears in ~6–7 out of 8 transactions:

| Rule | Confidence | Lift | Verdict |
|:--|:--|:--|:--|
| {Mushrooms} → {Pepperoni} | ≈ 0.80 | ≈ **1.0** | ❌ Useless — pepperoni is on nearly every pizza anyway |
| {Mushrooms} → {Olives} | ≈ 0.50 | ≈ **2.5** | ✓ Interesting — olives appear *much more* with mushrooms |

. . .

> **Lift corrects for popularity.** It asks: "Is B *more* likely given A than it would be by chance?"
> 
> Lift = 1.0 means A tells us nothing new about B. Lift > 1 means a genuine association.

**Try it** in Widget 2 — select Mushrooms → Pepperoni, then swap to Mushrooms → Olives.

---

## Mining Algorithms: Apriori

The **Apriori algorithm** efficiently discovers association rules by exploiting a key property:

::: {.callout-note}
## The Apriori principle
If an itemset is **infrequent**, all its **supersets** must also be infrequent.

Equivalently: every subset of a frequent itemset must also be frequent.
:::

. . .

**Algorithm outline:**

1. Count support of individual items → keep those above minimum support
2. Generate candidate pairs from frequent singles → count → prune
3. Generate candidate triples from frequent pairs → count → prune
4. Continue until no more frequent itemsets
5. Generate rules from frequent itemsets, filter by minimum confidence

. . .

Let's watch this happen step by step...

---

## Widget 3: Apriori Step-Through

Watch the Apriori algorithm work on the pizza transaction data. Adjust the minimum support threshold and step through each level of candidate generation and pruning.

```{ojs}
//| echo: false
import { aprioriWidget } from "./widgets/association-rules/apriori-widget.js"
viewof aprioriSim = aprioriWidget(pizzaToggles)
```

---

## Apriori vs Eclat

Two popular algorithms — same goal, different data representation:

:::: {.columns}
::: {.column width="50%" .fragment}
### Apriori
- **Horizontal** layout: each row = transaction with item list
- Breadth-first: generates candidates level by level
- Multiple passes over the database
- Easy to understand and implement
- Used in R's `arules::apriori()`
:::
::: {.column width="50%" .fragment}
### Eclat
- **Vertical** layout: each column = item with transaction ID list
- Uses **set intersection** of transaction ID lists to count support
- Often faster — fewer database passes
- Depth-first search through itemset lattice
- Used in R's `arules::eclat()`
:::
::::

. . .

Both produce the same frequent itemsets. Apriori is easier to teach; Eclat is often faster in practice. R's `arules` package supports both.

---

## Real Data: Recipe Ingredient Mining {.smaller}

Let's apply association rules to **real data** — 2,218 recipes from [Allrecipes](https://www.allrecipes.com), categorised by cuisine (TidyTuesday 2025-09-16).

Each recipe is a **transaction**; each ingredient is an **item**.

```{r load-recipes, cache=TRUE}
cuisines_raw <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-09-16/cuisines.csv",
  show_col_types = FALSE
)
glimpse(cuisines_raw)
```

---

## Parsing Ingredients into Transactions

The `ingredients` column is a comma-separated string. We need to extract base ingredient names.

```{r parse-ingredients}
# Words/phrases that are preparation descriptors, not ingredients.
# These appear as standalone tokens after comma-splitting
# (e.g. "garlic, minced" → "garlic" + "minced").
not_ingredients <- c(
  # preparation methods
  "chopped", "minced", "diced", "sliced", "grated", "shredded", "crushed",
  "julienned", "cubed", "halved", "quartered", "peeled", "seeded", "stemmed",
  "trimmed", "cleaned", "rinsed", "drained", "thawed", "softened", "melted",
  "beaten", "divided", "separated", "sifted", "crumbled", "toasted",
  "mashed", "pureed", "blanched", "boiled", "roasted", "juiced",
  "thinly sliced", "finely chopped", "coarsely chopped",
  "finely", "coarsely", "thinly",
  # usage / quantity phrases
  "to taste", "as needed", "or to taste", "or as needed",
  "or more", "or more to taste", "or more as needed",
  "salt to taste", "salt and pepper to taste",
  "salt and ground black pepper to taste",
  "salt and ground black pepper",
  "salt and pepper", "if desired", "for garnish",
  # too generic / ubiquitous to be informative
  "salt", "pepper", "kosher salt", "ground black pepper",
  "freshly ground black pepper",
  # other non-ingredients
  "skinless", "cooking spray", "room temperature", "needed", "taste"
)

# Unit regex: strips leading quantity + measurement word
unit_rx <- regex(
  paste0("^(cups?|tablespoons?|tbsp|teaspoons?|tsp|pounds?|lbs?|ounces?|oz|",
         "cans?|packages?|pkgs?|envelopes?|pinch(es)?|cloves?|slices?|",
         "stalks?|heads?|bunche?s?|pieces?|strips?|sprigs?)\\s+(of\\s+)?"),
  ignore_case = TRUE
)

recipe_items <- cuisines_raw |>
  select(name, country, ingredients) |>
  mutate(recipe_id = row_number()) |>
  separate_longer_delim(ingredients, delim = ", ") |>
  mutate(
    ingredient = ingredients |>
      str_to_lower() |>
      str_trim() |>
      str_remove_all("\\(.*?\\)") |>                          # remove parentheticals
      str_remove("^[0-9½¼¾⅓⅔⅛⅜⅝⅞/\\. ]+") |>               # remove quantities inc. Unicode fractions
      str_remove(unit_rx) |>                                   # remove measurement unit
      str_remove(regex("^(large|medium|small|extra-?large)\\s+", ignore_case = TRUE)) |>
      str_remove(regex("\\s+(to taste|as needed|if desired|or more to taste|or more)$",
                        ignore_case = TRUE)) |>
      str_remove(regex(",\\s*(divided|optional|drained|thawed|softened|melted|beaten)$",
                        ignore_case = TRUE)) |>
      str_trim() |>
      # Normalise variants → canonical form
      str_replace_all("^unsalted butter$", "butter") |>
      str_replace_all("^garlic powder$",   "garlic") |>
      str_replace_all("^minced garlic$",   "garlic") |>
      # Normalise plurals → singular
      str_replace_all("^eggs$",         "egg") |>
      str_replace_all("^onions$",       "onion") |>
      str_replace_all("^potatoes$",     "potato") |>
      str_replace_all("^tomatoes$",     "tomato") |>
      str_replace_all("^carrots$",      "carrot") |>
      str_replace_all("^lemons$",       "lemon") |>
      str_replace_all("^limes$",        "lime") |>
      str_replace_all("^mushrooms$",    "mushroom") |>
      str_replace_all("^bananas$",      "banana") |>
      str_replace_all("^green onions$", "green onion") |>
      str_replace_all("^cloves$",       "clove")
  ) |>
  filter(
    nchar(ingredient) > 2,
    !ingredient %in% not_ingredients
  )

# Keep the most common ingredients for interpretable rules
top_ingredients <- recipe_items |>
  count(ingredient, sort = TRUE) |>
  slice_head(n = 75) |>
  pull(ingredient)

recipe_items_filtered <- recipe_items |>
  filter(ingredient %in% top_ingredients)
```

---

## Converting to Transactions

```{r to-transactions}
# Create transaction format for arules
recipe_trans <- recipe_items_filtered |>
  select(recipe_id, ingredient) |>
  distinct() |>
  mutate(present = TRUE) |>
  pivot_wider(
    names_from = ingredient,
    values_from = present,
    values_fill = FALSE
  ) |>
  select(-recipe_id) |>
  as.matrix() |>
  as("transactions")

recipe_trans
```

---

## Item Frequency Plot

```{r item-freq-plot, fig.dim=c(12, 5)}
itemFrequencyPlot(recipe_trans, topN = 25, type = "relative",
                  col = "#4e79a7", main = "Top 25 Ingredients by Support")
```

---

## Mining Rules

```{r mine-rules}
recipe_rules <- apriori(
  recipe_trans,
  parameter = list(support = 0.05, confidence = 0.4, minlen = 2, maxlen = 4),
  control = list(verbose = FALSE)
)
recipe_rules
```

. . .

```{r top-rules}
# Top rules by lift
inspect(head(sort(recipe_rules, by = "lift"), n = 10))
```

---

## Filtering Rules by Ingredient

```{r filter-rules}
# What goes with garlic?
garlic_rules <- subset(recipe_rules, subset = lhs %in% "garlic" & lift > 1.2)
inspect(head(sort(garlic_rules, by = "confidence"), n = 10))
```

---


```{r rules-to-ojs}
#| echo: false
# Extract rules as a data frame for OJS
rules_df <- DATAFRAME(recipe_rules) |>
  as_tibble() |>
  mutate(
    lhs = as.character(LHS),
    rhs = as.character(RHS)
  ) |>
  select(lhs, rhs, support, confidence, coverage, lift, count) |>
  arrange(desc(lift))

ojs_define(rules_data = rules_df)
```

---

## Widget 4: Rule Explorer {.scrollable}

Explore all mined recipe rules interactively. Brush the scatterplot, adjust thresholds, or search for specific ingredients.

```{ojs}
//| echo: false
import { ruleExplorerWidget } from "./widgets/association-rules/rule-explorer-widget.js"
viewof ruleExplorer = ruleExplorerWidget(transpose(rules_data))
```

---

## Where Does Association Rule Mining Fit?

Association rules share ideas with many techniques you already know:

:::: {.columns}
::: {.column width="50%" .fragment}
### Direct Connections

**Conditional Probability** (STAT courses)

$$\text{confidence}(A \Rightarrow B) = P(B \mid A)$$

Confidence *is* conditional probability, estimated from data.

. . .

**Bayesian Reasoning**

$$\text{lift}(A \Rightarrow B) = \frac{P(B \mid A)}{P(B)}$$

Lift measures how much A changes our belief about B — like a likelihood ratio.
:::
::: {.column width="50%" .fragment}
### Unsupervised Discovery Family

**Clustering** (Week 10)

- Both discover structure without labels
- Clustering groups **observations** (rows)
- Association rules group **items** (columns)

**Decision Trees** (Week 9)

- Both produce interpretable IF→THEN rules
- Trees *partition* the feature space
- Association rules find *co-occurrence* patterns
:::
::::

---

## Collaborative Filtering: The Famous Application

> *"Customers who bought X also bought Y"*

This is **literally** association rule mining applied to recommendation systems.

:::: {.columns}
::: {.column width="50%" .fragment}
### Market Basket → Recommendations

| Market Basket | Recommendation |
|:--|:--|
| Transaction = shopping cart | Transaction = user purchase history |
| Item = product | Item = product |
| Rule: {A} → {B} | Recommendation: "You might also like B" |
| High lift = genuine association | High lift = personalised recommendation |
:::
::: {.column width="50%" .fragment}
### Key difference

Association rules find rules from **all transactions** together.

Modern recommender systems (Netflix, Spotify) use more sophisticated collaborative filtering (matrix factorisation, deep learning), but the **core intuition** is the same:

> Find patterns of co-occurrence across many users/transactions.
:::
::::

---

## Widget 5: Technique Comparison {.scrollable}

Click any cell to expand the explanation. See how association rules compare with other methods you've learned.

```{ojs}
//| echo: false
import { techniqueComparisonWidget } from "./widgets/association-rules/technique-comparison-widget.js"
viewof techComparison = techniqueComparisonWidget()
```

---

## Practical Tips

:::: {.columns}
::: {.column width="50%"}
### When to Use Association Rules

✅ Transactional data (baskets, carts, records)

✅ Binary/categorical variables

✅ Exploratory discovery — "what goes with what?"

✅ Large number of items with sparse transactions

✅ Business applications: cross-selling, layout optimisation, recommendations
:::
::: {.column width="50%"}
### When NOT to Use

❌ Small datasets (< 100 transactions) — rules will be unreliable

❌ Continuous data without discretisation

❌ When you need **causal** claims (rules show co-occurrence, not causation)

❌ When you have a specific target variable (use classification instead)

❌ Very dense data (every item in every transaction)
:::
::::

. . .

### Choosing Thresholds

| Parameter | Too low | Too high |
|:--|:--|:--|
| **min support** | Millions of rules, most rare & unreliable | Only obvious, uninteresting patterns |
| **min confidence** | Many weak associations | Miss rules where consequent is rare |

> **Start with** support ≥ 0.01, confidence ≥ 0.5, then adjust based on domain knowledge and the number of rules produced.

---

## Summary

:::: {.columns}
::: {.column width="60%"}
### Key Takeaways

1. **Association rules** discover co-occurrence patterns in transactional data
2. **Support** = how common, **Confidence** = how reliable, **Lift** = how interesting
3. High confidence alone is misleading — always check **lift**
4. **Apriori** prunes the search space using the antimonotone property
5. Association rules are the foundation of **recommendation systems**
6. Related to conditional probability, Bayes, clustering, and decision trees
:::
::: {.column width="40%"}
### Coming Up: Lab 11

In the lab you'll practice with:

- The built-in `Groceries` dataset (9,835 transactions)
- The `bank-rules.csv` dataset (banking service cross-selling)
- Mining, inspecting, and filtering rules
- Using `inspectDT()` for interactive exploration

### Study Guide

Chapter 8 of the course notes covers the same concepts with different examples and more mathematical detail.
:::
::::
