---
title: "Lecture 11"
subtitle: "Cluster Analysis"
author: "Data Mining"
format:
  revealjs:
    theme: [default, lecture11.scss]
    slide-number: true
    progress: true
    chalkboard: false
    incremental: false
    smaller: false
    transition: fade
    controls: true
    navigation-mode: linear
    width: 1600
    height: 900
execute:
  echo: true
  warning: false
  message: false
resources:
  - widgets
---

```{r}
#| label: setup
#| include: false
project_lib <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), "..", ".Rlib"),
  winslash = "/",
  mustWork = FALSE
)
dir.create(project_lib, recursive = TRUE, showWarnings = FALSE)
.libPaths(c(project_lib, .libPaths()))

suppressPackageStartupMessages(suppressWarnings(library(tidyverse)))
library(cluster)
library(broom)

theme_set(theme_minimal(base_size = 18))
options(pillar.sigfig = 5)
knitr::opts_chunk$set(fig.width = 8.4, fig.height = 5.4, fig.retina = 2, out.width = "100%", comment = "#>")

doc_dir <- dirname(knitr::current_input(dir = TRUE))
data_dir <- normalizePath(file.path(doc_dir, "..", "data"), winslash = "/", mustWork = FALSE)
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

read_tidytuesday_backup <- function(url, backup_file) {
  backup_path <- file.path(data_dir, backup_file)
  online <- tryCatch(
    readr::read_csv(url, show_col_types = FALSE),
    error = function(e) NULL
  )

  if (!is.null(online)) {
    if (requireNamespace("arrow", quietly = TRUE)) {
      arrow::write_parquet(online, backup_path)
    }
    return(online)
  }

  if (!requireNamespace("arrow", quietly = TRUE)) {
    stop("Package 'arrow' is required to read the local parquet backup at ", backup_path, call. = FALSE)
  }
  if (!file.exists(backup_path)) {
    stop("Could not download ", url, " and no local backup was found at ", backup_path, call. = FALSE)
  }

  arrow::read_parquet(backup_path) |> as_tibble()
}

make_synthetic_points <- function(seed = 161324) {
  set.seed(seed)
  blob_a <- tibble(
    x = rnorm(22, -2.7, 0.42),
    y = rnorm(22, -1.6, 0.42),
    shape = "blob_a"
  )
  ellipse <- tibble(a = rnorm(22, 0, 0.85), b = rnorm(22, 0, 0.18)) |>
    transmute(
      x = -0.2 + a - 0.35 * b,
      y = 2.3 + 0.35 * a + b,
      shape = "ellipse"
    )
  blob_b <- tibble(
    x = rnorm(22, 2.7, 0.38),
    y = rnorm(22, -1.8, 0.4),
    shape = "blob_b"
  )
  crescent <- tibble(
    theta = runif(20, 0.1 * pi, 1.02 * pi),
    radius = rnorm(20, 1.15, 0.08)
  ) |>
    transmute(
      x = 2.3 + radius * cos(theta),
      y = 1.8 + 0.78 * radius * sin(theta),
      shape = "crescent"
    )
  ring <- tibble(
    theta = runif(20, 0, 2 * pi),
    radius = rnorm(20, 0.86, 0.07)
  ) |>
    transmute(
      x = -3.1 + radius * cos(theta),
      y = 2.2 + radius * sin(theta),
      shape = "ring"
    )
  noise <- tibble(
    x = runif(12, -4, 4),
    y = runif(12, -3, 3),
    shape = "noise"
  )

  bind_rows(blob_a, ellipse, blob_b, crescent, ring, noise) |>
    mutate(id = row_number(), label = paste0("P", id)) |>
    select(id, label, x, y, shape)
}

monster_url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-05-27/monsters.csv"
qatar_url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-12-09/qatarcars.csv"

synthetic_points <- make_synthetic_points()
hclust_widget_points <- synthetic_points |>
  filter(shape %in% c("blob_a", "ellipse", "blob_b")) |>
  slice(1:14)
hclust_widget_dist <- dist(hclust_widget_points |> select(x, y))
hclust_single <- hclust(hclust_widget_dist, method = "single")
hclust_complete <- hclust(hclust_widget_dist, method = "complete")
hclust_average <- hclust(hclust_widget_dist, method = "average")
monsters <- read_tidytuesday_backup(monster_url, "monsters.parquet")
qatarcars <- read_tidytuesday_backup(qatar_url, "qatarcars.parquet")

numeric_cols <- c("str", "dex", "con", "int", "wis", "cha", "hp_number", "cr", "speed_base_number")
monster_scaled <- monsters |>
  mutate(across(all_of(numeric_cols), scale)) |>
  mutate(across(all_of(numeric_cols), as.numeric))

monster_dist <- monster_scaled |>
  select(all_of(numeric_cols)) |>
  dist(method = "euclidean")

monster_hc <- hclust(monster_dist, method = "complete")
monster_groups <- cutree(monster_hc, k = 5)
monster_kmeans_models <- set_names(
  map(1:8, function(k) {
    set.seed(161324 + k)
    kmeans(monster_scaled |> select(all_of(numeric_cols)), centers = k, nstart = 20)
  }),
  1:8
)
monster_kmeans <- monster_kmeans_models[["5"]]
monster_pam_models <- set_names(
  map(2:8, function(k) {
    pam(monster_scaled |> select(all_of(numeric_cols)), k = k)
  }),
  2:8
)
monster_pam <- monster_pam_models[["5"]]
monster_pca <- prcomp(monster_scaled |> select(all_of(numeric_cols)), scale. = FALSE)
monster_umap <- if (requireNamespace("uwot", quietly = TRUE)) {
  set.seed(161324)
  uwot::umap(
    monster_scaled |> select(all_of(numeric_cols)),
    n_neighbors = 15,
    min_dist = 0.15,
    metric = "euclidean",
    verbose = FALSE
  ) |>
    as_tibble(.name_repair = ~ c("umap1", "umap2"))
} else {
  NULL
}

monster_display_points <- if (!is.null(monster_umap)) {
  monster_umap |>
    transmute(x = umap1, y = umap2)
} else {
  tibble(x = monster_pca$x[, 1], y = monster_pca$x[, 2])
}
monster_display_labels <- if (!is.null(monster_umap)) c("UMAP 1", "UMAP 2") else c("PC1", "PC2")
monster_compare_points <- monsters |>
  transmute(id = row_number(), name, type) |>
  bind_cols(monster_display_points)

monster_kmeans_points <- monsters |>
  transmute(
    name,
    type,
    pc1 = monster_pca$x[, 1],
    pc2 = monster_pca$x[, 2]
  ) |>
  bind_cols(
    map_dfc(2:8, function(k) {
      tibble(!!paste0("k", k) := monster_kmeans_models[[as.character(k)]]$cluster)
    })
  )

monster_ranges <- monsters |>
  summarise(across(all_of(numeric_cols), list(min = min, max = max))) |>
  pivot_longer(everything(), names_to = c("variable", ".value"), names_pattern = "^(.*)_(min|max)$")

monster_elbow <- tibble(k = 1:8) |>
  mutate(
    model = map(k, ~ monster_kmeans_models[[as.character(.x)]]),
    glance = map(model, broom::glance)
  ) |>
  unnest(glance)

dbscan_base <- function(data, eps = 0.9, min_pts = 5) {
  d <- as.matrix(dist(data))
  n <- nrow(d)
  neighbors <- lapply(seq_len(n), function(i) which(d[i, ] <= eps))
  core <- lengths(neighbors) >= min_pts
  visited <- rep(FALSE, n)
  cluster <- rep(0L, n)
  cluster_id <- 0L

  for (i in seq_len(n)) {
    if (visited[i]) next
    visited[i] <- TRUE
    if (!core[i]) next
    cluster_id <- cluster_id + 1L
    cluster[i] <- cluster_id
    queue <- setdiff(neighbors[[i]], i)

    while (length(queue) > 0) {
      j <- queue[1]
      queue <- queue[-1]
      if (!visited[j]) {
        visited[j] <- TRUE
        if (core[j]) {
          queue <- union(queue, neighbors[[j]])
        }
      }
      if (cluster[j] == 0L) {
        cluster[j] <- cluster_id
      }
    }
  }

  tibble(
    cluster_id = if_else(cluster == 0L, -1L, cluster),
    cluster = if_else(cluster == 0L, "noise", paste0("cluster_", cluster)),
    core = core
  )
}

monster_compare_assignments <- bind_rows(
  map_dfr(c("single", "complete", "average"), function(linkage_value) {
    tibble(
      id = seq_len(nrow(monsters)),
      method = "hierarchical",
      linkage = linkage_value,
      k = 5L,
      eps = NA_real_,
      min_pts = NA_integer_,
      cluster_id = cutree(hclust(monster_dist, method = linkage_value), k = 5)
    )
  }),
  map_dfr(2:8, function(k_value) {
    tibble(
      id = seq_len(nrow(monsters)),
      method = "kmeans",
      linkage = NA_character_,
      k = k_value,
      eps = NA_real_,
      min_pts = NA_integer_,
      cluster_id = monster_kmeans_models[[as.character(k_value)]]$cluster
    )
  }),
  map_dfr(2:8, function(k_value) {
    tibble(
      id = seq_len(nrow(monsters)),
      method = "kmedoids",
      linkage = NA_character_,
      k = k_value,
      eps = NA_real_,
      min_pts = NA_integer_,
      cluster_id = unname(monster_pam_models[[as.character(k_value)]]$clustering)
    )
  }),
  expand_grid(eps = seq(0.6, 1.4, by = 0.1), min_pts = 3:8) |>
    mutate(
      result = map2(
        eps,
        min_pts,
        ~ dbscan_base(monster_scaled |> select(all_of(numeric_cols)), eps = .x, min_pts = .y)
      )
    ) |>
    mutate(
      rows = pmap(
        list(eps, min_pts, result),
        \(eps, min_pts, result) {
          tibble(
            id = seq_len(nrow(monsters)),
            method = "dbscan",
            linkage = NA_character_,
            k = NA_integer_,
            eps = eps,
            min_pts = min_pts,
            cluster_id = result$cluster_id
          )
        }
      )
    ) |>
    pull(rows) |>
    bind_rows()
)

ari <- function(labels_a, labels_b) {
  tab <- table(labels_a, labels_b)
  choose2 <- function(x) x * (x - 1) / 2
  nij <- sum(choose2(tab))
  ai <- sum(choose2(rowSums(tab)))
  bj <- sum(choose2(colSums(tab)))
  n <- sum(tab)
  expected <- ai * bj / choose2(n)
  max_index <- 0.5 * (ai + bj)
  if (max_index == expected) return(1)
  (nij - expected) / (max_index - expected)
}

bootstrap_ari <- function(data, baseline, reps = 30) {
  out <- numeric(reps)
  for (i in seq_len(reps)) {
    idx <- sample.int(nrow(data), replace = TRUE)
    unique_idx <- sort(unique(idx))
    if (length(unique_idx) < 5) next
    boot_data <- data[unique_idx, ]
    boot_dist <- dist(boot_data)
    boot_groups <- cutree(hclust(boot_dist, method = "complete"), k = 5)
    out[i] <- ari(baseline[unique_idx], boot_groups)
  }
  out
}

monster_bootstrap <- bootstrap_ari(monster_scaled |> select(all_of(numeric_cols)), monster_groups, reps = 35)

dbscan_data <- synthetic_points |>
  filter(shape %in% c("crescent", "ring", "noise"))

dbscan_clusters <- dbscan_base(dbscan_data |> select(x, y), eps = 0.92, min_pts = 5)
dbscan_plot_data <- bind_cols(dbscan_data, dbscan_clusters)

doc_dir <- dirname(knitr::current_input(dir = TRUE))
csv_out <- file.path(doc_dir, "widgets", "clustering", "synthetic_points.csv")
write_csv(synthetic_points, csv_out)

if (requireNamespace("arrow", quietly = TRUE)) {
  arrow::write_parquet(synthetic_points, file.path(doc_dir, "..", "data", "synthetic-clusters.parquet"))
  arrow::write_parquet(qatarcars, file.path(doc_dir, "..", "data", "qatarcars.parquet"))
}
```

# Cluster Analysis

## What changes in an unsupervised problem?

- Classification starts with known labels. Clustering starts with structure and asks whether groups exist at all.
- We need a notion of similarity before we can talk about a cluster.
- Different clustering algorithms make different promises.
  - Hierarchical clustering gives a nested tree.
  - K-means gives compact partitions.
  - K-medoids swaps means for representative observations.
  - DBSCAN keeps dense shapes and leaves sparse points as noise.

---

## Lecture roadmap

1. Build dissimilarities for multivariate data.
2. Cut a dendrogram and check whether the resulting groups are stable.
3. Compare centroid and medoid based partitioning.
4. Use density clustering when convex clusters fail.
5. Finish with a modern workflow: parquet, HDBSCAN, UMAP, and the Qatar Cars exercise.

---

## Three datasets for today
#todo place link to actual tidytuesday datasets here

- `synthetic_points`: 118 two-dimensional observations with blobs, an ellipse, a crescent, a ring, and noise.
- `monsters`: the real TidyTuesday D&D Monsters data.
- `qatarcars`: the TidyTuesday Qatar Cars data.

```{r}
bind_rows(
  tibble(dataset = "Synthetic 2D", rows = nrow(synthetic_points), columns = ncol(synthetic_points)),
  tibble(dataset = "D&D Monsters", rows = nrow(monsters), columns = ncol(monsters)),
  tibble(dataset = "Qatar Cars", rows = nrow(qatarcars), columns = ncol(qatarcars))
)
```

---

## Synthetic data for the widgets

```{r}
ggplot(synthetic_points, aes(x = x, y = y, colour = shape)) +
  geom_point(alpha = 0.8, size = 2.1) +
  guides(colour = guide_legend(nrow = 2)) +
  labs(x = NULL, y = NULL, colour = NULL)
```

::: {.widget-note}
The widgets use the same broad geometry, but they subsample points so each step remains legible on a slide.
:::

---

## Dissimilarity matrices

<!-- TODO: put in gower dissimilarity metric from notes -->

- Euclidean distance:
  $$
  d_{st} = \sqrt{\sum_{j=1}^p (x_{sj} - x_{tj})^2}
  $$
- Manhattan distance:
  $$
  d_{st} = \sum_{j=1}^p |x_{sj} - x_{tj}|
  $$
- Gower dissimilarity mixes numeric and categorical variables by rescaling each numeric field to its range and then combining numeric and factor mismatches.

---

## A distance matrix is just pairwise bookkeeping

```{r}
round(as.matrix(dist(synthetic_points |> slice(1:6) |> select(x, y))), 2)
```

- Diagonal entries are zero.
- The matrix is symmetric.
- Everything that follows depends on how informative this matrix is for the data at hand.

---

## Scaling matters before distance-based clustering

<!-- TODO: this needs cleaned up. The point is that the numeric features of the monsters are on different scales, so we need to standardise them before using Euclidean distance. The code below shows the ranges of the original features, which motivates the need for scaling. -->

::: {.left-code}
```{r}
monster_ranges
```
:::

::: {.right-plot}
```{r}
monster_ranges |>
  pivot_longer(c(min, max), names_to = "bound", values_to = "value") |>
  ggplot(aes(x = variable, y = value, fill = bound)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = NULL)
```
:::

---

## D&D Monsters: why this is a good clustering demo

<!-- TODO: remove this if it stays duplicative. The numeric features are on different scales, so scaling is necessary before clustering; this may already be covered by the previous slide. -->

- Ability scores live roughly on the `1` to `25` scale.
- `hp_number` is much larger.
- `type` gives us a rough truth signal after clustering.
- The data are wide enough that 2D plots are incomplete, but still interpretable.

```{r}
monsters |>
  count(type, sort = TRUE)
```

# Hierarchical Clustering

---

## Agglomerative hierarchical clustering

1. Start with each observation in its own cluster.
2. Find the closest pair of clusters.
3. Merge them.
4. Repeat until only one cluster remains.

The only extra decision is how to define the distance between clusters.

- Single linkage uses the closest pair.
- Complete linkage uses the furthest pair.
- Average linkage uses the mean pairwise distance.

---

## Hierarchical clustering on monsters

```{r}
plot(monster_hc, labels = FALSE, hang = -1, main = "Complete-linkage dendrogram for D&D Monsters")
```

---

## Cutting the tree gives an actual partition

<!-- TODO: cleanup this slide. The point is that we can cut the tree at different heights to get different numbers of clusters, then compare those clusters to the known `type` variable to see how well the clustering did. The table shows the cross-tabulation of `type` and the cluster assignments, and the plot shows how those clusters look in terms of `hp_number` and `cr`, coloured by cluster. -->

::: {.left-code}
```{r}
table(type = monsters$type, cluster = monster_groups)
```
:::

::: {.right-plot}
```{r}
monsters |>
  mutate(cluster = factor(monster_groups)) |>
  ggplot(aes(x = hp_number, y = cr, colour = cluster)) +
  geom_point(alpha = 0.85, size = 2.1) +
  labs(x = "Hit points", y = "Challenge rating")
```
:::

---

## Interactive build: hierarchical clustering

<!-- TODO: the point is that the order in which clusters merge depends on the linkage method, and this can lead to different clusterings. The widget should make that tradeoff visible. Also consider making it physically larger on the slide because it can take the whole slide. -->

::: {.widget-cell}
```{ojs}
//| echo: false
import { hierarchicalWidget } from "./widgets/clustering/hierarchical-widget.js"
hierarchicalWidget()
```
:::

---

## Linkage is not a minor detail

::: {.widget-cell}
```{ojs}
//| echo: false
import { hierarchicalComparison } from "./widgets/clustering/hierarchical-widget.js"
hierarchicalComparison()
```
:::

- Single linkage is sensitive to chains of points.
- Complete linkage prefers tight, compact groups.
- Average linkage usually sits between them.
- The widget makes the tradeoff visible because the same points merge in a different order as you change linkage.
---


## Bootstrap stability: do the same clusters return?

- Resample observations.
- Refit the clustering on each resample.
- Compare the resulting partition with the original.
- A stable cluster should survive small perturbations to the data.

---

## Bootstrap stability on monsters

::: {.widget-note}
The adjusted Rand index compares every pair of monsters across two partitions: did that pair stay together, or stay apart, in both clusterings? It then subtracts the agreement expected by chance,
$$
\mathrm{ARI} = \frac{\sum_{ij}\binom{n_{ij}}{2} - E}{\frac{1}{2}\left[\sum_i \binom{a_i}{2} + \sum_j \binom{b_j}{2}\right] - E},
$$
so `1` means a perfect match to the baseline clustering, `0` means about as much agreement as random assignment, and negative values mean worse than random agreement.
:::

::: {.left-code}
```{r}
#| echo: false
tibble(bootstrap_ari = monster_bootstrap)
```
:::

::: {.right-plot}
```{r}
#| echo: false
tibble(bootstrap_ari = monster_bootstrap) |>
  ggplot(aes(x = bootstrap_ari)) +
  geom_histogram(binwidth = 0.05, fill = "#197278", colour = "white") +
  labs(x = "Adjusted Rand index against baseline", y = "Resamples")
```
:::

---

## Interactive resampling view

::: {.widget-cell}
```{ojs}
//| echo: false
import { bootstrapStabilityWidget } from "./widgets/clustering/bootstrap-stability-widget.js"
bootstrapStabilityWidget()
```
:::

# K-means and K-medoids

---

## K-means minimises within-cluster variation

For Euclidean distance the objective is
$$
\sum_{k=1}^K \sum_{C(i)=k} \lVert x_i - \mu_k \rVert^2.
$$

Algorithm:

<!-- TODO: match the algorithm wording in the /notes. -->

1. Initialise centroids.
2. Assign each point to the nearest centroid.
3. Recompute centroids.
4. Repeat until the assignments stop changing.

---

## K-means on monsters

```{r}
#| echo: false
ojs_define(
  monster_kmeans_ojs_points = monster_kmeans_points,
  monster_elbow_ojs = monster_elbow |> select(k, tot.withinss)
)
```

::: {.widget-cell}
```{ojs}
//| echo: false
import { monsterKmeansWidget } from "./widgets/clustering/monster-kmeans-widget.js"
monsterKmeansWidget(monster_kmeans_ojs_points, monster_elbow_ojs)
```
:::

---

## Interactive centroid updates

::: {.widget-cell}
```{ojs}
//| echo: false
import { kmeansKmedoidsWidget } from "./widgets/clustering/kmeans-kmedoids-widget.js"
kmeansKmedoidsWidget()
```
:::

---

## K-medoids / PAM

- Replace the centroid with a real observation, the medoid.
- This makes the method less sensitive to outliers.
- It also lets us use any dissimilarity, not just Euclidean distance.

```{r}
monster_pam |> pluck("id.med")
```

---

## Silhouette widths for PAM

```{r}
plot(silhouette(monster_pam), border = NA, do.clus.stat = FALSE, do.n.k = FALSE, main = "PAM silhouettes")
```

---

## When K-means and PAM disagree

- K-means is faster and works well for compact numeric clusters.
- PAM is more robust when outliers are likely or when the cluster center needs to be interpretable.
- Both still force every observation into one of `K` groups.

# Density Clustering

---

## DBSCAN definitions

- `eps`: neighborhood radius.
- `MinPts`: minimum neighborhood size.
- Core point: at least `MinPts` observations in its `eps`-neighborhood.
- Border point: not core, but within the neighborhood of a core point.
- Noise point: neither core nor border.

---

## Why DBSCAN is different

- It does not require `K`.
- It can recover non-convex shapes.
- It can leave points unclustered.
- It is sensitive to scale and parameter choice.

---

## Custom DBSCAN on the synthetic crescent/ring data

```{r}
ggplot(dbscan_plot_data, aes(x = x, y = y, colour = cluster, shape = core)) +
  geom_point(size = 2.2, alpha = 0.88) +
  labs(x = NULL, y = NULL, colour = NULL, shape = "Core point")
```

---

## Parameter choice: k-distance idea

<!-- TODO: cleanup this slide. The point is that we can use the distance to the `k`th nearest neighbor to help choose `eps`. We sort the points by this distance and look for an elbow in the plot, which indicates a natural separation between dense regions and noise. -->

- Here `MinPts = 5`, so a point becomes core when it has `4` other points inside its `eps` neighborhood.
- That makes the distance to the `4`th nearest other point the smallest `eps` that would make that observation core.
- Sort those distances and look for an elbow: dense-cluster points stay low, while border/noise points rise sharply.
- Choosing `eps` near that bend usually keeps dense regions together without flooding sparse noise.

```{r}
#| echo: true
min_pts <- 5
k <- min_pts - 1  # 4 other neighbors, because the point itself is counted in MinPts
dmat <- as.matrix(dist(dbscan_data |> select(x, y)))
knn_distance <- apply(dmat, 1, function(row) sort(row)[k + 1])  # +1 because the self-distance is 0
tibble(rank = seq_along(knn_distance), kth_distance = sort(knn_distance)) |>
  ggplot(aes(x = rank, y = kth_distance)) +
  geom_line(colour = "#197278", size = 1.1) +
  labs(
    x = "Ordered points",
    y = paste0("Distance to ", k, "th nearest other point (MinPts = ", min_pts, ")")
  )
```

---

## Interactive DBSCAN expansion

::: {.widget-cell}
```{ojs}
//| echo: false
import { dbscanWidget } from "./widgets/clustering/dbscan-widget.js"
dbscanWidget()
```
:::

::: {.widget-note}
DBSCAN repeats: inspect an unvisited point, test whether it is core, expand through its neighbors if it is core, and otherwise leave it as provisional noise until a later core point reaches it.
:::

---

## HDBSCAN: the 2026 default when density varies

<!-- TODO: remove or rewrite this slide. The point is that DBSCAN can struggle when clusters have different densities, because a single `eps` value may not capture all the clusters well. HDBSCAN extends DBSCAN by building a hierarchy over varying density thresholds and then extracting stable clusters. It currently feels tacked on without an example or code-heavy payoff; maybe link to the study guide instead. -->

- HDBSCAN extends DBSCAN into a hierarchy and avoids fixing a single `eps`.
- It is a better default when dense and sparse clusters coexist.
- In R, the standard entry point is `dbscan::hdbscan()`.

```{r}
#| eval: false
library(dbscan)
hdbscan(monster_scaled |> select(all_of(numeric_cols)), minPts = 8)
```

---

## UMAP in one slide

<!-- TODO: this may duplicate the later point that high-dimensional data are hard to visualise directly, and that UMAP provides a 2D view while preserving useful structure. Decide whether this theory slide and the later practice framing should be merged. -->

- UMAP starts from local neighbor relationships in the original feature space and builds weighted affinities `w_{ij}`.
- It then finds low-dimensional coordinates `y_i` whose low-dimensional affinities `v_{ij}` preserve those neighborhoods as well as possible.

$$
v_{ij} = \left(1 + a\lVert y_i - y_j \rVert^{2b}\right)^{-1}
$$

$$
\mathcal{L} =
\sum_{i \ne j}
\left[
w_{ij}\log\frac{w_{ij}}{v_{ij}}
  + (1-w_{ij})\log\frac{1-w_{ij}}{1-v_{ij}}
\right]
$$

- So `UMAP 1` and `UMAP 2` are not original variables. They are coordinates chosen to keep nearby monsters nearby in the map.

---

## UMAP view of monsters

```{r}
#| echo: false
ojs_define(
  monster_umap_ojs_points = monster_compare_points,
  monster_umap_axis_x = monster_display_labels[[1]],
  monster_umap_axis_y = monster_display_labels[[2]]
)
```

::: {.widget-cell}
```{ojs}
//| echo: false
import { monsterUmapWidget } from "./widgets/clustering/monster-umap-widget.js"
monsterUmapWidget(
  monster_umap_ojs_points,
  monster_umap_axis_x,
  monster_umap_axis_y
)
```
:::

::: {.widget-note}
The next comparison slide uses these UMAP axes only for display. Each clustering method is still fit on the full scaled monster feature set.
:::

---

## Method comparison on the same geometry

```{r}
#| echo: false
ojs_define(
  monster_compare_ojs_points = monster_compare_points,
  monster_compare_ojs_assignments = monster_compare_assignments,
  monster_compare_axis_x = monster_display_labels[[1]],
  monster_compare_axis_y = monster_display_labels[[2]]
)
```

::: {.widget-cell}
```{ojs}
//| echo: false
import { clusterComparisonWidget } from "./widgets/clustering/cluster-comparison-widget.js"
clusterComparisonWidget(
  monster_compare_ojs_points,
  monster_compare_ojs_assignments,
  monster_compare_axis_x,
  monster_compare_axis_y
)
```
:::

---

## Practical comparison

| Method | Strength | Risk | Needs scaling? |
|:--|:--|:--|:--|
| Hierarchical | Full tree, no need to pre-fix `K` | Linkage choice matters | Usually yes |
| K-means | Fast, familiar, easy to explain | Splits non-convex shapes | Yes |
| K-medoids | Robust, medoids are real points | Slower than K-means | Usually yes |
| DBSCAN | Handles noise and shape | Parameter tuning can be awkward | Yes |
| HDBSCAN | Better with varying density | Extra package, less introductory intuition | Yes |

# Modern Workflow Notes

<!-- TODO: this whole section is funky. The point is that modern data science practice often uses parquet for data exchange, UMAP for high-dimensional cluster displays, and tidyclust for tidyverse/tidymodels workflows. The Qatar Cars dataset is the lab focus where students apply these clustering techniques to a real dataset. -->

---

## Parquet for lab and project handoff

<!-- TODO: the files should be saved before this slide, and the point is that parquet is a modern file format that is more efficient than CSV and preserves data types. The code below checks whether `arrow` is available and writes the monsters, Qatar Cars, and synthetic points datasets to `data/`. Rework the framing because students will not have a lab directory. -->

```{r}
if (requireNamespace("arrow", quietly = TRUE)) {
  tibble(status = "Monsters, Qatar Cars, and synthetic cluster parquet files are available under data/.")
} else {
  tibble(status = "arrow is not installed here; keep the code pattern and run it in the teaching environment.")
}
```

```{r}
#| eval: false
library(arrow)
write_parquet(monsters, "../data/monsters.parquet")
write_parquet(qatarcars, "../data/qatarcars.parquet")
write_parquet(synthetic_points, "../data/synthetic-clusters.parquet")
```

---

## tidyclust keeps clustering inside tidymodels

```{r}
#| eval: false
library(tidyclust)
library(parsnip)

k_means(num_clusters = 5) |>
  fit(~ ., data = monster_scaled |> select(all_of(numeric_cols)))
```

- Use this when you want clustering to sit inside a recipes and workflows pipeline.

---

## Student exercise preview: Qatar Cars

::: {.left-code-wide}
```{r}
qatarcars |>
  select(type, enginetype, length, width, height, mass, horsepower, price, performance, economy, trunk, seating) |>
  summary()
```
:::

::: {.right-plot-narrow}
```{r}
qatarcars |>
  ggplot(aes(x = price, y = horsepower, colour = type)) +
  geom_point(alpha = 0.85, size = 2.1) +
  labs(x = "Price", y = "Horsepower", colour = "Type")
```
:::

---

## Questions for the lab

1. Which variables should define similarity for the cars?
2. How much does scaling change the result?
3. Does a fixed `K` make sense, or do the data want a density-based view?
4. Which truth signal is more informative for validation: `type` or `enginetype`?

---

## Suggested workflow for the Qatar Cars lab

1. Start with `summary()` and a few scatterplots.
2. Scale the numeric features before using Euclidean distance.
3. Compare complete-linkage hierarchical clustering with K-means.
4. Use PAM silhouettes to decide whether your chosen `K` is defensible.
5. Try DBSCAN last on a small feature subset where the geometry is visible.

# Summary

---

## Cluster analysis summary

- Distances create the problem definition, so scaling and variable choice come first.
- Hierarchical clustering is good for structure discovery and explanation.
- K-means and PAM are explicit partitioning tools once a candidate `K` is on the table.
- DBSCAN is the right move when you care about shape or noise.
- Modern practice adds parquet for exchange, HDBSCAN for density variation, and UMAP for visualization.

---

## Code references for later

```{r}
tribble(
  ~task, ~code,
  "Distance matrix", "dist(x, method = 'euclidean')",
  "Hierarchical clustering", "hclust(d, method = 'complete')",
  "Cut dendrogram", "cutree(hc, k = 5)",
  "K-means", "kmeans(x, centers = 5, nstart = 20)",
  "K-medoids", "cluster::pam(x, k = 5)",
  "DBSCAN", "dbscan::dbscan(x, eps = ..., minPts = ...)",
  "HDBSCAN", "dbscan::hdbscan(x, minPts = ...)"
)
```
