---
title: "Lecture 10"
subtitle: "Ensemble Learning"
author: "Data Mining"
format:
  revealjs:
    theme: [default, custom_quarto.css]
    slide-number: true
    progress: true
    chalkboard: false
    incremental: false
    smaller: false
    transition: fade
    controls: true
    navigation-mode: linear
    width: 1600
    height: 900
execute:
  echo: true
  warning: false
  message: false
resources:
  - widgets
---

```{r}
#| label: setup
#| include: false
project_lib <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), "..", ".Rlib"),
  winslash = "/",
  mustWork = FALSE
)
dir.create(project_lib, recursive = TRUE, showWarnings = FALSE)
.libPaths(c(project_lib, .libPaths()))

suppressPackageStartupMessages(suppressWarnings(library(tidyverse)))
library(tidymodels)
library(baguette)
library(ranger)
tidymodels_prefer()

theme_set(theme_minimal(base_size = 18))
options(pillar.sigfig = 5)
knitr::opts_chunk$set(fig.width = 8.4, fig.height = 5.4, fig.retina = 2, out.width = "100%", comment = "#>")

doc_dir <- dirname(knitr::current_input(dir = TRUE))
data_dir <- normalizePath(file.path(doc_dir, "..", "data"), winslash = "/", mustWork = FALSE)
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

source("../common/longbeach.R")
longbeach <- read_csv("../data/longbeach.csv", show_col_types = FALSE)
adoption_data <- prepare_longbeach_adoption_data(longbeach)

set.seed(2024)
adopt_split  <- initial_split(adoption_data, prop = 0.75, strata = is_adopted)
adopt_train  <- training(adopt_split)
adopt_test   <- testing(adopt_split)
adopt_folds  <- vfold_cv(adopt_train, v = 5, strata = is_adopted)

adopt_recipe <- recipe(is_adopted ~ ., data = adopt_train) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors())

# Load AutoML data for widget
automl_json <- tryCatch(
  jsonlite::read_json("widgets/ensembles/data/longbeach-automl.json"),
  error = function(e) NULL
)
automl_data <- if (!is.null(automl_json)) automl_json else list(leaderboard = list())
```

## Why Combine Models?

- Single models can be unstable (high variance)
- Different models make different errors
- Combining predictions reduces overall error

### The variance reduction argument

If we have $B$ learners $\hat{f}_1, \ldots, \hat{f}_B$, each with variance $\sigma^2$, and pairwise correlation $\rho$:

$$\operatorname{Var}\!\left(\frac{1}{B}\sum_{b=1}^B \hat{f}_b(x)\right) = \rho\sigma^2 + \frac{1-\rho}{B}\,\sigma^2$$

- The second term shrinks as $B$ increases
- Averaging helps most when learners are **weakly correlated** (low $\rho$)
- Building diverse learners is central to ensemble design

---

## Bagging: Bootstrap Aggregating

- Draw $B$ bootstrap samples from training data
- Fit a learner to each bootstrap sample
- Average predictions from all $B$ learners

### Key insight

Each bootstrap sample excludes ~36.8% of observations (out-of-bag)
→ We can use OOB observations to estimate test error without a validation set

### Interactive: Bagging Widget

```{ojs}
import { baggingWidget } from "./widgets/ensembles/bagging-widget.js"
baggingWidget()
```

---

## Bagging in R with tidymodels

```{r}
#| eval: false
# Single decision tree
single_tree_spec <- decision_tree(mode = "classification") |>
  set_engine("rpart")

# Bagged trees (50 bootstrap samples)
bag_spec <- bag_tree(mode = "classification") |>
  set_engine("rpart", times = 50)

wf <- workflow() |> add_recipe(adopt_recipe)

# Compare performance
single_tree_res <- wf |>
  add_model(single_tree_spec) |>
  fit_resamples(adopt_folds, metrics = metric_set(accuracy, roc_auc))

bag_res <- wf |>
  add_model(bag_spec) |>
  fit_resamples(adopt_folds, metrics = metric_set(accuracy, roc_auc))

bind_rows(
  collect_metrics(single_tree_res) |> mutate(model = "single tree"),
  collect_metrics(bag_res)         |> mutate(model = "bagged trees")
) |> select(model, .metric, mean)
```

---

## Random Forests

- Bagging + one important modification
- At each split, only consider a random subset of $m$ predictors
- Common defaults: $m = \sqrt{p}$ (classification), $m = p/3$ (regression)

### Why it works

- Reduces correlation between trees (lower $\rho$ in variance formula)
- Each tree sees only a subset of predictors
- More diverse ensemble → better predictions

### Variable Importance

Two main approaches:

1. **Impurity-based**: Sum of Gini/variance reduction from splits on each variable
2. **Permutation**: Increase in OOB error when variable values are randomly shuffled

---

## Random Forest in R

```{r}
#| eval: false
rf_spec <- rand_forest(mode = "classification", trees = 500, mtry = tune()) |>
  set_engine("ranger", importance = "impurity")

rf_res <- wf |>
  add_model(rf_spec) |>
  tune_grid(adopt_folds, grid = 5, metrics = metric_set(roc_auc))

show_best(rf_res, metric = "roc_auc")

# Fit final model
final_rf <- wf |>
  add_model(rf_spec) |>
  add_formula(is_adopted ~ .) |>
  last_fit(adopt_split)

collect_metrics(final_rf)

# Variable importance
final_rf |> extract_fit_engine() |> vip()
```

---

## Boosting: Sequential Error Correction

- Build learners **sequentially**, not independently
- Each new learner targets the **residuals** (errors) of the previous
- Learning rate controls how much each tree contributes

### How it works

1. Start with a simple model (e.g., mean prediction)
2. Fit a tree to the residuals
3. Add this tree to the ensemble (weighted by learning rate)
4. Repeat

### Interactive: Boosting Widget

```{ojs}
import { boostingWidget } from "./widgets/ensembles/boosting-widget.js"
boostingWidget()
```

---

## Gradient Boosting vs XGBoost

### Gradient Boosting

- General framework for minimizing a loss function
- At each step, fit a tree to the **negative gradient** of the loss
- Loss function determines what we're optimizing (MSE, logloss, etc.)

### XGBoost (eXtreme Gradient Boosting)

- Optimized implementation with regularization
- Adds L2 penalty on leaf weights: $\lambda \sum_j w_j^2$
- Adds tree complexity penalty: $\gamma \times \text{number of leaves}$
- Much faster, better regularization than classic gradient boosting

---

## XGBoost Hyperparameters

| Parameter | Effect |
|-----------|--------|
| `eta` (learning rate) | Lower = more trees, less overfitting |
| `max_depth` | Deeper = more complex, more overfitting |
| `subsample` | Fraction of data per tree (reduces overfitting) |
| `colsample_bytree` | Fraction of features per tree |
| `min_child_weight` | Minimum sum of instance weight in leaf |
| `lambda` (reg_lambda) | L2 regularization on leaf weights |
| `alpha` (reg_alpha) | L1 regularization on leaf weights |

### Interactive: XGBoost Widget

```{ojs}
import { xgboostWidget } from "./widgets/ensembles/xgboost-widget.js"
xgboostWidget()
```

---

## Stacking: Combining Different Model Families

- Use different model families as **base learners**
- Train a **meta-learner** on their predictions
- Base learner predictions become features for the meta-learner

### Key idea

Each model family has different inductive biases → different errors
→ a well-chosen meta-learner can exploit complementary strengths

### Interactive: Stacking Widget

```{ojs}
import { stackingWidget } from "./widgets/ensembles/stacking-widget.js"
stackingWidget()
```

---

## Stacking in R

```{r}
#| eval: false
# Define base models
tree_spec <- decision_tree() |> set_mode("classification")
rf_spec <- rand_forest(trees = 100) |> set_mode("classification")
xgb_spec <- boost_tree(trees = 100) |> set_mode("classification")
log_spec <- logistic_reg() |> set_mode("classification")

# Create a model stack
library(stacks)
stack_spec <- stacks() |>
  add_candidates(tree_res) |>
  add_candidates(rf_res) |>
  add_candidates(xgb_res) |>
  add_candidates(log_res)

# Fit the meta-learner
stack_fit <- stack_spec |>
  blend_predictions() |>
  fit_members()

# Predictions use all members + meta-learner
stack_pred <- adopt_test |>
  bind_cols(predict(stack_fit, adopt_test))
```

---

## AutoML: Automated Model Selection

- Automatically searches over many model families and hyperparameters
- Popular tools: H2O AutoML, auto-sklearn, TPOT, mljar

### Trade-offs

**Pros:**
- Fast baseline performance
- Less manual tuning required
- Good for initial exploration

**Cons:**
- Can be a "black box"
- Risk of data leakage in validation
- May miss domain-specific insights
- Often overkill for simple problems

### Interactive: AutoML Widget

```{ojs}
import { automlWidget } from "./widgets/ensembles/automl-widget.js"
automlWidget(automl_data)
```

---

## Summary: Ensemble Methods

| Method | Key Idea | When to Use |
|--------|----------|-------------|
| **Bagging** | Average bootstrap trees | Reduce variance of unstable models |
| **Random Forests** | Bagging + feature subsampling | Default for tabular data |
| **Boosting** | Sequential error correction | When you need high accuracy |
| **XGBoost** | Regularized boosting | Industry standard for competitions |
| **Stacking** | Combine different model families | When you have diverse models |
| **AutoML** | Automated search | Fast baseline, exploration |

### Key Takeaways

- Ensembles typically outperform single models
- More trees = more stable (but diminishing returns)
- Balance complexity with interpretability
- Always validate properly (don't overfit to validation!)

---

## For More Details

- **Notes Chapter 9**: Full theoretical treatment with mathematical derivations
- **Lab 9**: Hands-on practice with ensemble methods
- **Tidymodels documentation**: https://www.tidymodels.org/
- **XGBoost documentation**: https://xgboost.readthedocs.io/
