---
title: "Workshop 10: Ensemble Learning with tidymodels"
output: html_document
---

In this workshop we build an ensemble workflow on the **Long Beach Animal
Shelter** TidyTuesday data. The lecture introduced the concepts; here we
will do the modelling step by step in `tidymodels`.

Our main goals are to:

- define a clean binary target for adoption
- build one shared preprocessing recipe
- compare a single tree, a bagged tree, and a boosted tree
- evaluate the models with cross-validation
- optionally combine tuned models with `stacks`

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(baguette)
library(stacks)
library(bonsai)
source(file.path(dirname(knitr::current_input(dir = TRUE)), "..", "common", "tidytuesday.R"))
source(file.path(dirname(knitr::current_input(dir = TRUE)), "..", "common", "longbeach.R"))

theme_set(theme_minimal())
tidymodels_prefer()

doc_dir <- dirname(knitr::current_input(dir = TRUE))
data_dir <- resolve_data_dir(doc_dir)
longbeach_url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-04/longbeach.csv"

longbeach <- read_tidytuesday_cached(
  longbeach_url,
  backup_file = "longbeach.parquet",
  data_dir = data_dir,
  csv_file = "longbeach.csv"
)
if (requireNamespace("arrow", quietly = TRUE)) {
  write_tidytuesday_parquet(longbeach, "longbeach.parquet", data_dir)
}

adoption_data <- prepare_longbeach_adoption_data(longbeach)
```

## 1. Inspect the modelling data

We are predicting whether a cat or dog ends up in the `Adopted` class,
using intake-time variables only.

```{r}
adoption_data |> glimpse()

adoption_data |>
  count(is_adopted) |>
  mutate(prop = n / sum(n))
```

### Try yourself

1. Which predictors are numeric, and which are categorical?
2. Why is `accuracy` not enough on its own for this task?
3. Which variables do you think might be most predictive of adoption?

## 2. Training/test split and resampling

```{r}
set.seed(161324)
adoption_split <- initial_split(adoption_data, prop = 0.8, strata = is_adopted)
adoption_train <- training(adoption_split)
adoption_test <- testing(adoption_split)

adoption_folds <- vfold_cv(adoption_train, v = 3, strata = is_adopted)
adoption_folds
```

We will compare models using:

- `roc_auc`
- `pr_auc`
- `accuracy`

```{r}
ensemble_metrics <- metric_set(roc_auc, pr_auc, accuracy)
```

## 3. One recipe for all workflows

We will use one recipe across all models so the comparison is fair.

```{r}
adoption_recipe <- recipe(is_adopted ~ ., data = adoption_train) |>
  step_other(primary_color, intake_subtype, threshold = 0.03) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())

adoption_recipe
```

### Try yourself

1. Why is `step_other()` useful here?
2. Why is `step_dummy()` needed before `xgboost`?
3. Which parts of the recipe are protecting us against new categories in test data?

## 4. Define model specifications

We will compare:

- a single decision tree
- a bagged tree
- a boosted tree with `xgboost`

```{r}
tree_spec <- decision_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")

bag_spec <- bag_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart", times = 10) |>
  set_mode("classification")

xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")
```

Now attach the recipe to each model:

```{r}
tree_wf <- workflow() |>
  add_recipe(adoption_recipe) |>
  add_model(tree_spec)

bag_wf <- workflow() |>
  add_recipe(adoption_recipe) |>
  add_model(bag_spec)

xgb_wf <- workflow() |>
  add_recipe(adoption_recipe) |>
  add_model(xgb_spec)
```

## 5. Tune the models

Use intentionally small grids so the workflow stays fast and inspectable in a workshop setting.

```{r}
tree_grid <- grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 3)
bag_grid <- grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 2)
xgb_grid <- grid_latin_hypercube(
  trees(),
  tree_depth(),
  learn_rate(),
  min_n(),
  loss_reduction(),
  sample_prop(),
  size = 6
)
```

### Try yourself

Tune each workflow on the same folds:

```{r}
set.seed(161324)
tree_res <- tune_grid(
  tree_wf,
  resamples = adoption_folds,
  grid = tree_grid,
  metrics = ensemble_metrics,
  control = control_stack_grid()
)

set.seed(161324)
bag_res <- tune_grid(
  bag_wf,
  resamples = adoption_folds,
  grid = bag_grid,
  metrics = ensemble_metrics,
  control = control_stack_grid()
)

set.seed(161324)
xgb_res <- tune_grid(
  xgb_wf,
  resamples = adoption_folds,
  grid = xgb_grid,
  metrics = ensemble_metrics,
  control = control_stack_grid()
)
```

Compare the best results:

```{r}
bind_rows(
  show_best(tree_res, metric = "roc_auc", n = 3) |> mutate(model = "tree"),
  show_best(bag_res, metric = "roc_auc", n = 3) |> mutate(model = "bagged"),
  show_best(xgb_res, metric = "roc_auc", n = 3) |> mutate(model = "xgboost")
) |>
  select(model, .metric, mean, everything())
```

## 6. Finalise the best settings

Choose the best tuning parameters for each model and evaluate them on
the held-out test set.

```{r}
best_tree <- select_best(tree_res, metric = "roc_auc")
best_bag <- select_best(bag_res, metric = "roc_auc")
best_xgb <- select_best(xgb_res, metric = "roc_auc")

final_tree <- finalize_workflow(tree_wf, best_tree)
final_bag <- finalize_workflow(bag_wf, best_bag)
final_xgb <- finalize_workflow(xgb_wf, best_xgb)
```

### Try yourself

1. Fit each final workflow to `adoption_train`.
2. Use `augment()` or `predict()` on `adoption_test`.
3. Compare the final test metrics.
4. Which model would you deploy if:
   - you only care about predictive performance?
   - you want a more explainable model?

## 7. Optional capstone: stack the tuned models

If you have time, use the tuning results to build a stacked ensemble.

```{r}
model_stack <- stacks() |>
  add_candidates(tree_res) |>
  add_candidates(bag_res) |>
  add_candidates(xgb_res) |>
  blend_predictions(metric = metric_set(roc_auc)) |>
  fit_members()
```

### Try yourself

1. Compare the stack against the best single model on the test set.
2. Which member models get non-zero weight?
3. Does stacking buy enough extra performance to justify the extra complexity?
