---
title: "Workshop 12: Association Rule Mining"
output: html_document
---

In this workshop we build association rules from a TidyTuesday recipe
dataset, then finish with a shorter extension on the local bank-service
data. The main goals are to:

- turn text data into transactions
- inspect item frequencies
- mine rules with `apriori()`
- interpret support, confidence, and lift
- filter rules using `subset()`

We will use the `tidytuesdayR` package to pull the recipe data directly
from TidyTuesday. If RStudio prompts you to install `tidytuesdayR`, let
it do so before you continue.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(arules)

theme_set(theme_minimal())

load_tidytuesday_dataset <- function(tt_date, dataset_name, parquet_url) {
  if (requireNamespace("tidytuesdayR", quietly = TRUE)) {
    tt_data <- tryCatch(
      suppressMessages(tidytuesdayR::tt_load(tt_date)),
      error = function(e) NULL
    )

    if (!is.null(tt_data) &&
        dataset_name %in% names(tt_data) &&
        !is.null(tt_data[[dataset_name]])) {
      return(tt_data[[dataset_name]] |> tibble::as_tibble())
    }
  }

  if (!requireNamespace("arrow", quietly = TRUE)) {
    stop(
      "Install the 'tidytuesdayR' package (RStudio will usually prompt you), ",
      "or install 'arrow' to use the backup data file.",
      call. = FALSE
    )
  }

  parquet_file <- tempfile(fileext = ".parquet")
  on.exit(unlink(parquet_file), add = TRUE)
  download_status <- tryCatch(
    utils::download.file(parquet_url, parquet_file, mode = "wb", quiet = TRUE),
    error = function(e) 1L
  )

  if (!identical(download_status, 0L) || !file.exists(parquet_file)) {
    stop(
      "Could not load the dataset from TidyTuesday or the course backup file.",
      call. = FALSE
    )
  }

  arrow::read_parquet(parquet_file) |> tibble::as_tibble()
}

compose_ggplot <- function(plot, ...) {
  layers <- list(...)

  for (i in seq_along(layers)) {
    plot <- ggplot2:::add_ggplot(plot, layers[[i]], paste0("layer_", i))
  }

  plot
}
```

## 1. Recipe ingredients as transactions

The lecture used a TidyTuesday dataset of recipes grouped by cuisine.
Each recipe will be treated as one transaction, and each ingredient as
an item.

```{r}
cuisines_raw <- load_tidytuesday_dataset(
  tt_date = "2025-09-16",
  dataset_name = "cuisines",
  parquet_url = "https://knowlton.co.nz/161324/data/cuisines.parquet"
)

cuisines_raw |> glimpse()
```

The `ingredients` column is a comma-separated string, so we need to turn
it into one ingredient per row before creating the transaction matrix.

```{r}
not_ingredients <- c(
  "chopped", "minced", "diced", "sliced", "grated", "shredded", "crushed",
  "julienned", "cubed", "halved", "quartered", "peeled", "seeded", "stemmed",
  "trimmed", "cleaned", "rinsed", "drained", "thawed", "softened", "melted",
  "beaten", "divided", "separated", "sifted", "crumbled", "toasted",
  "mashed", "pureed", "blanched", "boiled", "roasted", "juiced",
  "thinly sliced", "finely chopped", "coarsely chopped",
  "finely", "coarsely", "thinly",
  "to taste", "as needed", "or to taste", "or as needed",
  "or more", "or more to taste", "or more as needed",
  "salt to taste", "salt and pepper to taste",
  "salt and ground black pepper to taste",
  "salt and ground black pepper",
  "salt and pepper", "if desired", "for garnish",
  "salt", "pepper", "kosher salt", "ground black pepper",
  "freshly ground black pepper",
  "skinless", "cooking spray", "room temperature", "needed", "taste"
)

unit_rx <- regex(
  paste0(
    "^(cups?|tablespoons?|tbsp|teaspoons?|tsp|pounds?|lbs?|ounces?|oz|",
    "cans?|packages?|pkgs?|envelopes?|pinch(es)?|cloves?|slices?|",
    "stalks?|heads?|bunche?s?|pieces?|strips?|sprigs?)\\s+(of\\s+)?"
  ),
  ignore_case = TRUE
)

recipe_items <- cuisines_raw |>
  select(name, country, ingredients) |>
  mutate(recipe_id = row_number()) |>
  separate_longer_delim(ingredients, delim = ", ") |>
  mutate(
    ingredient = ingredients |>
      str_to_lower() |>
      str_trim() |>
      str_remove_all("\\(.*?\\)") |>
      str_remove("^[0-9\\u00bd\\u00bc\\u00be\\u2153\\u2154\\u215b\\u215c\\u215d\\u215e/\\. ]+") |>
      str_remove(unit_rx) |>
      str_remove(regex("^(large|medium|small|extra-?large)\\s+", ignore_case = TRUE)) |>
      str_remove(regex("\\s+(to taste|as needed|if desired|or more to taste|or more)$",
                       ignore_case = TRUE)) |>
      str_remove(regex(",\\s*(divided|optional|drained|thawed|softened|melted|beaten)$",
                       ignore_case = TRUE)) |>
      str_trim() |>
      str_replace_all("^unsalted butter$", "butter") |>
      str_replace_all("^garlic powder$", "garlic") |>
      str_replace_all("^minced garlic$", "garlic") |>
      str_replace_all("^eggs$", "egg") |>
      str_replace_all("^onions$", "onion") |>
      str_replace_all("^potatoes$", "potato") |>
      str_replace_all("^tomatoes$", "tomato") |>
      str_replace_all("^carrots$", "carrot") |>
      str_replace_all("^lemons$", "lemon") |>
      str_replace_all("^limes$", "lime") |>
      str_replace_all("^mushrooms$", "mushroom") |>
      str_replace_all("^bananas$", "banana") |>
      str_replace_all("^green onions$", "green onion") |>
      str_replace_all("^cloves$", "clove")
  ) |>
  filter(
    nchar(ingredient) > 2,
    !ingredient %in% not_ingredients
  )
```

### Try yourself

1. Count the most common ingredients in `recipe_items`.

2. Do the common ingredients seem globally useful, or are some likely to
   inflate confidence without being especially informative?

3. How might that affect your later interpretation of lift?

```{r}
recipe_items |>
  count(ingredient, sort = TRUE) |>
  slice_head(n = 20)
```

## 2. Converting to a transaction database

To keep the rules interpretable, we restrict attention to the 75 most
common ingredients.

```{r}
top_ingredients <- recipe_items |>
  count(ingredient, sort = TRUE) |>
  slice_head(n = 75) |>
  pull(ingredient)

recipe_items_filtered <- recipe_items |>
  filter(ingredient %in% top_ingredients)

recipe_trans <- recipe_items_filtered |>
  select(recipe_id, ingredient) |>
  distinct() |>
  mutate(present = TRUE) |>
  pivot_wider(
    names_from = ingredient,
    values_from = present,
    values_fill = FALSE
  ) |>
  select(-recipe_id) |>
  as.matrix() |>
  as("transactions")

recipe_trans
```

### Try yourself

1. Inspect the first ten transactions using `inspect()`.

2. Compare `inspect()` with `DATAFRAME()`. Which would you use for
   exploratory work?

```{r}
recipe_trans |>
  head(n = 10) |>
  inspect()
```

## 3. Item frequencies

```{r}
itemFrequencyPlot(recipe_trans, topN = 25, type = "relative")
```

### Try yourself

1. Recreate the item-frequency plot using `ggplot2`.

2. Which ingredients are common enough that a high-confidence rule with
   that ingredient on the right-hand side might still be uninteresting?

```{r}
recipe_freq <- recipe_trans |>
  itemFrequency() |>
  enframe(name = "ingredient", value = "support") |>
  arrange(desc(support)) |>
  slice_head(n = 25)

compose_ggplot(
  ggplot(recipe_freq, aes(x = support, y = fct_reorder(ingredient, support))),
  geom_col()
)
```

## 4. Mining rules with Apriori

We now mine rules with moderate support and confidence thresholds.

```{r}
recipe_rules <- apriori(
  recipe_trans,
  parameter = list(support = 0.05, confidence = 0.40, minlen = 2, maxlen = 4),
  control = list(verbose = FALSE)
)

recipe_rules
summary(recipe_rules)
```

### Try yourself

1. How many rules were found?

2. Inspect the top ten rules by support, by confidence, and by lift.

3. Which ranking would you use first if the goal is to find genuinely
   informative associations rather than merely common ones?

```{r}
recipe_rules |>
  sort(by = "support") |>
  head(n = 10) |>
  inspect()

recipe_rules |>
  sort(by = "confidence") |>
  head(n = 10) |>
  inspect()

recipe_rules |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```

## 5. Misleading high-confidence rules

High confidence on its own can be misleading when the consequent is very
common. The code below looks for rules with high confidence but lift
close to one.

```{r}
recipe_rules |>
  subset(confidence >= 0.75 & lift <= 1.05) |>
  sort(by = "confidence") |>
  head(n = 10) |>
  inspect()
```

### Try yourself

1. Pick one rule from the output above and explain why it is not very
   interesting despite its high confidence.

2. Compare it with a rule that has slightly lower confidence but much
   higher lift.

## 6. Filtering rules

Use `subset()` together with `%in%`, `%ain%`, and `%pin%` to focus on a
particular ingredient or collection of ingredients.

```{r}
garlic_rules <- recipe_rules |>
  subset(lhs %in% "garlic" & lift > 1.1)

garlic_rules |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```

### Try yourself

1. Find rules with `olive oil` on the left-hand side.

2. Find rules where both `garlic` and `onion` appear on the left-hand
   side.

3. Find one ingredient whose rules look cuisine-specific rather than
   globally common.

## 7. Optional interactive inspection

If you have `arulesViz` available, interactive inspection is useful once
you have reduced the rule set to something manageable.

```{r, eval=FALSE}
library(arulesViz)

recipe_rules |>
  sort(by = "lift") |>
  head(n = 100) |>
  inspectDT()
```

## 8. Extension: bank-service rules

The bank dataset is already stored locally in `../data/bank-rules.csv`.
Each row is one account-service combination.

```{r}
bank <- read_csv("../data/bank-rules.csv", show_col_types = FALSE) |>
  transactions(format = "long", cols = c("ACCOUNT", "SERVICE"))

bank_rules <- apriori(
  bank,
  parameter = list(support = 0.005, confidence = 0.4),
  control = list(verbose = FALSE)
)
```

### Try yourself

1. Find rules with `CCRD` on the right-hand side. Which services look
   useful for cross-selling a credit card?

2. Find rules that contain both `MTG` and `HMEQLC`.

3. Compare the bank rules with the recipe rules. Which dataset gives
   more obviously actionable associations?

```{r}
bank_rules |>
  subset(rhs %in% "CCRD") |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```

### Final questions

1. In the recipe data, what kinds of items produce high confidence but
   uninformative rules?

2. Which filtering tools in `arules` did you find most useful?

3. If you had to recommend one metric for ranking candidate rules in an
   exploratory analysis, would you choose support, confidence, or lift?
