---
title: "Workshop 10 Guide: Clustering"
output: html_document
---

In this workshop we follow the lecture workflow for clustering on **two datasets**.
For each task, the **2020 NZ Election** results are shown as a fully worked
example. Your job is to adapt the same code to the **D&D Monsters** data — the
variable names are similar, so the code changes are small.

- build a dissimilarity matrix and check the effect of scaling
- use PCA and UMAP to visualise the data in two dimensions
- compare hierarchical clustering, $K$-means, and PAM
- use silhouette widths to defend a choice of $K$
- profile clusters
- apply density-based clustering with DBSCAN

Use the `tidytuesdayR` package to pull the monsters data from TidyTuesday. If
RStudio prompts you to install `tidytuesdayR`, let it do so before continuing.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(cluster)
library(uwot)
library(dbscan)
library(plotly)
library(tidytuesdayR)
library(scales)


theme_set(theme_minimal())
```

## The two datasets

### 2020 New Zealand General Election

In the 2020 general election each voter casts two votes: a **party vote** (for a
political party) and a **candidate vote** (for a local MP). We cluster the 72
electorates by party vote shares to see how regions group together politically.

```{r import-election-data}
party <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nzelect/2020_party.csv")

party_mat <- party |>
  column_to_rownames("Electorate")

party_scaled <- party_mat |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

party |> summarise(across(-Electorate, mean))
```

### D&D Monsters

Each row is a monster with a `type` label plus combat features like strength,
dexterity, HP, challenge rating, and speed. We cluster on the numeric features
and use `type` afterward to interpret the clusters.

```{r load-monsters}
monsters <- tidytuesdayR::tt_load("2025-05-27")$monsters |>
  tibble::as_tibble()

monster_vars <- c(
  "str", "dex", "con", "int", "wis", "cha",
  "hp_number", "cr", "speed_base_number"
)

monster_rows <- monsters |>
  mutate(.row = row_number()) |>
  select(.row, name, type, all_of(monster_vars)) |>
  mutate(
    name = if_else(
      is.na(name) | name == "",
      paste0("monster_", .row),
      name
    ),
    name = make.unique(name)
  ) |>
  drop_na(type, all_of(monster_vars))

monster_data <- monster_rows |>
  select(type, all_of(monster_vars))

monster_scaled <- monster_rows |>
  select(all_of(monster_vars)) |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))
```

# 1. Distances and scaling

## Voting — worked example

Party vote shares are percentages on a common 0–100 scale. Parties with larger
vote shares (like Labour) naturally have larger variance. We compare raw
Euclidean distance with z-scored distance for the first six electorates:

```{r}
round(as.matrix(dist(party_mat |> slice(1:6))), 2)
round(as.matrix(dist(party_scaled |> slice(1:6))), 2)
```

## D&D — your turn

Compute the raw and z-scored Euclidean distance matrices for the first six
monsters. Which variables appear to dominate the unscaled distances?

```{r}
# Raw distances
monster_mat <- monster_rows |>
  select(name, all_of(monster_vars)) |>
  slice(1:6) |>
  column_to_rownames("name")

round(as.matrix(dist(monster_mat)), 2)

# Z-scored distances
monster_scaled_first6 <- monster_scaled |>
  slice(1:6) |>
  as.data.frame()
rownames(monster_scaled_first6) <- monster_rows$name[1:6]

round(as.matrix(dist(monster_scaled_first6)), 2)
```

The raw distances are mostly driven by `hp_number`, `cr`, and
`speed_base_number`, because they vary on much larger raw scales than the
ability scores.

# 2. Dimension reduction: PCA and UMAP

## Voting — worked example

We build PCA (linear) and UMAP (nonlinear) 2D projections so we can inspect the
electorate structure before clustering.

```{r}

party_pca <- prcomp(party_scaled, center = TRUE, scale. = TRUE)
party_pca_points <- as_tibble(party_pca$x) |>
  select(pc1 = PC1, pc2 = PC2) |>
  (
    function(scores) {
      bind_cols(party |> transmute(Electorate), scores)
    }
  )()
set.seed(8675309)
party_umap_points <- uwot::umap(
  party_scaled,
  n_neighbors = 15,
  min_dist = 0.1,
  n_components = 2,
  metric = "euclidean"
) |>
  as_tibble() |>
  set_names(c("umap1", "umap2")) |>
  (
    function(embedding) {
      bind_cols(party |> transmute(Electorate), embedding)
    }
  )()

p_pca <- ggplot(
  party_pca_points,
  aes(x = pc1, y = pc2, label = Electorate, text = Electorate)
) +
  geom_point() +
  labs(x = "PC1", y = "PC2", title = "Voting: PCA")

p_umap <- ggplot(
  party_umap_points,
  aes(x = umap1, y = umap2, label = Electorate, text = Electorate)
) +
  geom_point() +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Voting: UMAP")

ggplotly(p_pca, tooltip = "text")
ggplotly(p_umap, tooltip = "text")

```

Try to change the seed and see how UMAP changes. 

## D&D — your turn

Fit PCA and UMAP on `monster_scaled`. Use `set.seed(161324)` for UMAP. The code
follows the same pattern as the voting example — swap the data and variable
names. Use `name` as the point identifier.

```{r}
# PCA
monster_pca <- prcomp(monster_scaled, center = TRUE, scale. = TRUE)
monster_pca_points <- as_tibble(monster_pca$x) |>
  select(pc1 = PC1, pc2 = PC2) |>
  (
    function(scores) {
      bind_cols(monster_rows |> transmute(name), scores)
    }
  )()

# UMAP
set.seed(161324)
monster_umap_points <- uwot::umap(
  monster_scaled,
  n_neighbors = 15,
  min_dist = 0.1,
  n_components = 2,
  metric = "euclidean"
) |>
  as_tibble() |>
  set_names(c("umap1", "umap2")) |>
  (
    function(embedding) {
      bind_cols(monster_rows |> transmute(name), embedding)
    }
  )()
```

Plot both projections side-by-side with linked highlighting (`key = name`).

```{r}
# Side-by-side PCA vs UMAP plot
p_monster_pca <- ggplot(
  monster_pca_points,
  aes(x = pc1, y = pc2, label = name, text = name, key = name)
) +
  geom_point() +
  labs(x = "PC1", y = "PC2", title = "Monsters: PCA")

p_monster_umap <- ggplot(
  monster_umap_points,
  aes(x = umap1, y = umap2, label = name, text = name, key = name)
) +
  geom_point() +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monsters: UMAP")

subplot(
  ggplotly(p_monster_pca, tooltip = "text"),
  ggplotly(p_monster_umap, tooltip = "text"),
  nrows = 1
)
```

The two views tell a similar story: there is a broad central mass of monsters,
plus a smaller set of more extreme high-power cases that separate more clearly.

# 3. Hierarchical clustering

## Voting — worked example

Average-linkage hierarchical clustering on scaled party data:

```{r}
party_dist <- dist(party_scaled, method = "euclidean")
party_hc <- hclust(party_dist, method = "average")

plot(party_hc, labels = FALSE, hang = -1,
     main = "Average-linkage dendrogram for NZ 2020 party vote")
```

Try single, complete, and average linkage side by side:

```{r}
party_hc_single   <- hclust(party_dist, method = "single")
party_hc_complete <- hclust(party_dist, method = "complete")
party_hc_average  <- hclust(party_dist, method = "average")

plot(party_hc_single,   labels = FALSE, hang = -1, main = "Single linkage")
plot(party_hc_complete, labels = FALSE, hang = -1, main = "Complete linkage")
plot(party_hc_average,  labels = FALSE, hang = -1, main = "Average linkage")
```

Use the silhouette measure to choose $K$, then cut the tree:

```{r}
library(factoextra)

party_scaled |>
  fviz_nbclust(FUNcluster = factoextra::hcut, hc_method = "average", k.max = 10,
               method = "silhouette") |>
  ggplotly()

party_groups <- cutree(party_hc, k = 2)
```

Cluster means:

```{r}
party_clustered <- party |>
  mutate(Cluster = factor(party_groups))

party_clustered |>
  group_by(Cluster) |>
  summarise(across(-Electorate, mean), n_electorates = n()) |>
  arrange(desc(n_electorates))
```

Which electorates are in each cluster?

```{r}
party_clustered |>
  select(Electorate, Cluster) |>
  arrange(Cluster, Electorate) |>
  print(n = 72)
```

## D&D — your turn

Fit hierarchical clustering on `monster_scaled`. Try single, complete, and
average linkage. Which gives the most balanced cluster sizes?

```{r}
# Distance matrix
monster_dist <- dist(monster_scaled, method = "euclidean")


# Three linkage methods
monster_hc_single   <- hclust(monster_dist, method = "single")
monster_hc_complete <- hclust(monster_dist, method = "complete")
monster_hc_average  <- hclust(monster_dist, method = "average")

plot(monster_hc_single,   labels = FALSE, hang = -1, main = "Monsters: Single linkage")
plot(monster_hc_complete, labels = FALSE, hang = -1, main = "Monsters: Complete linkage")
plot(monster_hc_average,  labels = FALSE, hang = -1, main = "Monsters: Average linkage")
```

Complete linkage is the least chain-like here, so it gives the cleanest working
partition.

```{r}
# Silhouette to choose K
monster_scaled |>
  fviz_nbclust(FUNcluster = factoextra::hcut, hc_method = "average", k.max = 10,
               method = "silhouette") |>
  ggplotly()


# Cut the tree at your chosen K
monster_hc <- monster_hc_complete
monster_hc_groups <- cutree(monster_hc, k = 2)
```

The silhouette peak is at $k = 2$, so this guide uses a broad two-cluster cut.

```{r}
# Cluster means by monster type (hint: join back to monster_data)
monster_clustered <- monster_rows |>
  mutate(Cluster = factor(monster_hc_groups))

monster_clustered |>
  group_by(Cluster) |>
  summarise(across(all_of(monster_vars), mean), n = n()) |>
  arrange(desc(n))

monster_clustered |>
  count(Cluster, type, sort = TRUE) |>
  head(25)
```

Cluster 1 picks up many of the larger dragons and fiends, while Cluster 2 holds
the broader middle of the dataset.

# 4. $K$-means and the elbow heuristic

## Voting — worked example

```{r}
set.seed(67)
party_elbow <- tibble(k = 1:10) |>
  mutate(
    model = map(k, ~ kmeans(party_scaled, centers = .x, nstart = 20)),
    fit = map(model, glance)
  ) |>
  unnest(fit)

party_elbow
```

```{r}
p<-party_elbow |>
  ggplot(aes(x = k, y = tot.withinss)) +
  geom_line() +
  geom_point(aes(text = paste0("k = ", k, "<br>Total WSS = ", round(tot.withinss))), size = 2)

ggplotly(p, tooltip = "text")
```

Fit $K$-means with $k = 6$:

```{r}
set.seed(67)
party_km6 <- kmeans(party_scaled, centers = 6, nstart = 25)

party |>
  mutate(Cluster = factor(party_km6$cluster)) |>
  group_by(Cluster) |>
  summarise(across(-Electorate, mean), n = n()) |>
  arrange(desc(n))
```

Let's plot this out on the PCA projection:
```{r}
# no legend, just colour by cluster
party_km6_points <- party_pca_points |>
  mutate(Cluster = factor(party_km6$cluster))

p_km_pca <- party_km6_points |>
  ggplot(aes(x = pc1, y = pc2, colour = Cluster,
             label = Electorate, key = Electorate)) +
  geom_point(aes(text = Electorate), alpha = 0.8, size
             = 2) +
  labs(x = "PC1", y = "PC2", title = "Voting:
K-means clusters on PCA")

ggplotly(p_km_pca, tooltip = "text")

# lets add a UMAP cluster
party_umap_km6_points <- party_umap_points |>
  mutate(Cluster = factor(party_km6$cluster))


p_km_umap <- party_umap_km6_points |>
  ggplot(aes(x = umap1, y = umap2, colour = Cluster,
             label = Electorate, key = Electorate)) +
  geom_point(aes(text = Electorate), alpha = 0.8, size
             = 2) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Voting:
K-means")

ggplotly(p_km_umap, tooltip = "text")


```


## D&D — your turn

Fit $K$-means for several values of $k$, produce an elbow plot, then fit a final
model with your chosen $k$.

```{r}
# Elbow tibble
set.seed(161324)
monster_elbow <- tibble(k = 1:15) |>
  mutate(
    model = map(k, ~ kmeans(monster_scaled, centers = .x, nstart = 20)),
    fit = map(model, glance)
  ) |>
  unnest(fit)

monster_elbow

# Elbow plot
monster_elbow |>
  ggplot(aes(x = k, y = tot.withinss)) +
  geom_line() +
  geom_point(aes(text = paste0("k = ", k, "<br>Total WSS = ", round(tot.withinss))), size = 2)
```

The elbow is gradual, but $k = 10$ is a reasonable compromise between detail and
parsimony.

```{r}
# Final K-means fit and cluster means
set.seed(161324)
monster_km10 <- kmeans(monster_scaled, centers = 10, nstart = 25)

monster_rows |>
  mutate(Cluster = factor(monster_km10$cluster)) |>
  group_by(Cluster) |>
  summarise(across(all_of(monster_vars), mean), n = n()) |>
  arrange(desc(n))

# PCA plot coloured by K-means cluster
monster_km10_points <- monster_pca_points |>
  mutate(Cluster = factor(monster_km10$cluster))

p_monster_km_pca <- monster_km10_points |>
  ggplot(aes(x = pc1, y = pc2, colour = Cluster,
             label = name, key = name)) +
  geom_point(aes(text = name), alpha = 0.8, size = 2) +
  labs(x = "PC1", y = "PC2", title = "Monsters: K-means clusters on PCA")

ggplotly(p_monster_km_pca, tooltip = "text")
```

# 5. PAM and silhouette widths

## Voting — worked example

PAM fits $K$-medoids for $K = 2$ through $K = 8$. We compare average silhouette
widths:

```{r}
pam_sil_summary <- map_dfr(2:8, function(k_value) {
  pam_fit <- pam(party_scaled, k = k_value)
  sil <- silhouette(pam_fit)
  widths <- sil[, "sil_width"]
  tibble(
    k = k_value,
    avg_silhouette = mean(widths)
  )
})

pam_sil_summary |>
  arrange(desc(avg_silhouette))
```

Fit PAM with $k = 4$ and produce the silhouette plot:

```{r}
party_pam4 <- pam(party_scaled, k = 4)
plot(silhouette(party_pam4), border = NA, do.clus.stat = FALSE, do.n.k = FALSE)
```

Per-cluster average silhouette widths and boxplots:

```{r}
party_pam_sil <- party_pam4 |>
  pluck("silinfo", "widths") |>
  as_tibble()

party_pam_sil |>
  group_by(cluster) |>
  summarise(avg_sil = mean(sil_width), n_electorates = n())

pps<- party_pam_sil |>
  ggplot(aes(x = factor(cluster), y = sil_width)) +
  geom_boxplot() +
  labs(x = "Cluster", y = "Silhouette width")
  
ggplotly(pps, tooltip = "y")
```

Which electorates are the medoids?

```{r}
medoid_rows <- party_pam4$id.med
party_mat |>
  slice(medoid_rows) |>
  rownames_to_column("Electorate") |>
  mutate(Cluster = 1:4) |>
  select(Cluster, Electorate, everything())

party_medoid_ids <- rownames(party_mat)[medoid_rows]

pca_medoid_plot <- party_pca_points |>
  mutate(
    is_medoid = Electorate %in% party_medoid_ids,
    hover = paste0(
      Electorate,
      "<br>Medoid: ", if_else(is_medoid, "Yes", "No")
    )
  ) |>
  ggplot(aes(x = pc1, y = pc2)) +
  geom_point(
    data = ~ filter(.x, !is_medoid),
    aes(text = hover),
    alpha = 0.5,
    size = 2,
    colour = "grey50"
  ) +
  geom_point(
    data = ~ filter(.x, is_medoid),
    aes(text = hover),
    alpha = 1,
    size = 4,
    shape = 21,
    stroke = 1.2,
    fill = "#2C7FB8",
    colour = "black"
  ) +
  labs(
    x = "PC1",
    y = "PC2",
    title = "PCA: PAM medoids highlighted"
  )

umap_medoid_plot <- party_umap_points |>
  mutate(
    is_medoid = Electorate %in% party_medoid_ids,
    hover = paste0(
      Electorate,
      "<br>Medoid: ", if_else(is_medoid, "Yes", "No")
    )
  ) |>
  ggplot(aes(x = umap1, y = umap2)) +
  geom_point(
    data = ~ filter(.x, !is_medoid),
    aes(text = hover),
    alpha = 0.5,
    size = 2,
    colour = "grey50"
  ) +
  geom_point(
    data = ~ filter(.x, is_medoid),
    aes(text = hover),
    alpha = 1,
    size = 4,
    shape = 21,
    stroke = 1.2,
    fill = "#2C7FB8",
    colour = "black"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    title = "PCA/Umap: PAM medoids highlighted"
  )

subplot(
  ggplotly(pca_medoid_plot, tooltip = "text"),
  ggplotly(umap_medoid_plot, tooltip = "text"),
  nrows = 1
)
```

## D&D — your turn

Fit PAM for $K = 2$ through $K = 8$, identify the best $K$ by average silhouette
width, produce the silhouette plot, boxplots, and extract the medoids.

```{r}
# PAM silhouette summary for K = 2 to 8
monster_pam_sil_summary <- map_dfr(2:12, function(k_value) {
  pam_fit <- pam(monster_scaled, k = k_value)
  sil <- silhouette(pam_fit)
  widths <- sil[, "sil_width"]
  tibble(
    k = k_value,
    avg_silhouette = mean(widths)
  )
})

monster_pam_sil_summary |>
  arrange(desc(avg_silhouette))
```

Average silhouette is highest at $k = 2$, so PAM also prefers a broad
two-cluster split here.

```{r}
# PAM with best K, silhouette plot
best_monster_pam_k <- monster_pam_sil_summary |>
  slice_max(avg_silhouette, n = 1, with_ties = FALSE) |>
  pull(k)

monster_pam_best <- pam(monster_scaled, k = best_monster_pam_k)

best_monster_pam_k
monster_pam_best$silinfo$avg.width

plot(silhouette(monster_pam_best), border = NA, do.clus.stat = FALSE, do.n.k = FALSE)
```

```{r}
# Per-cluster silhouettes and boxplots
monster_pam_sil <- monster_pam_best |>
  pluck("silinfo", "widths") |>
  as_tibble()

monster_pam_sil |>
  group_by(cluster) |>
  summarise(avg_sil = mean(sil_width), n_monsters = n())

monster_pam_sil |>
  ggplot(aes(x = factor(cluster), y = sil_width)) +
  geom_boxplot() +
  labs(x = "Cluster", y = "Silhouette width")
```

```{r}
# Medoid monsters
medoid_rows <- monster_pam_best$id.med
monster_rows |>
  slice(medoid_rows) |>
  mutate(Cluster = seq_along(medoid_rows)) |>
  select(Cluster, name, type, all_of(monster_vars))
```

```{r}

# Plot the mediods as above highlighted on teh PCA and UMAP projections

monster_medoid_ids <- monster_rows$name[medoid_rows]

pca_medoid_data <- monster_pca_points |>
  mutate(
    is_medoid = name %in% monster_medoid_ids,
    hover = paste0(
      name,
      "<br>Medoid: ", if_else(is_medoid, "Yes", "No")
    )
  )

umap_medoid_data <- monster_umap_points |>
  mutate(
    is_medoid = name %in% monster_medoid_ids,
    hover = paste0(
      name,
      "<br>Medoid: ", if_else(is_medoid, "Yes", "No")
    )
  )

p_pca <- plot_ly() |>
  add_markers(
    data = pca_medoid_data |> filter(!is_medoid),
    x = ~pc1,
    y = ~pc2,
    text = ~hover,
    hoverinfo = "text",
    marker = list(
      size = 6,
      opacity = 0.5,
      color = "grey50"
    ),
    name = "Non-medoid"
  ) |>
  add_markers(
    data = pca_medoid_data |> filter(is_medoid),
    x = ~pc1,
    y = ~pc2,
    text = ~hover,
    hoverinfo = "text",
    marker = list(
      size = 12,
      opacity = 1,
      color = "#2C7FB8",
      line = list(
        color = "black",
        width = 1.2
      )
    ),
    name = "Medoid"
  ) |>
  layout(
    title = "PCA: PAM medoids highlighted",
    xaxis = list(title = "PC1"),
    yaxis = list(title = "PC2")
  )

p_umap <- plot_ly() |>
  add_markers(
    data = umap_medoid_data |> filter(!is_medoid),
    x = ~umap1,
    y = ~umap2,
    text = ~hover,
    hoverinfo = "text",
    marker = list(
      size = 6,
      opacity = 0.5,
      color = "grey50"
    ),
    name = "Non-medoid"
  ) |>
  add_markers(
    data = umap_medoid_data |> filter(is_medoid),
    x = ~umap1,
    y = ~umap2,
    text = ~hover,
    hoverinfo = "text",
    marker = list(
      size = 12,
      opacity = 1,
      color = "#2C7FB8",
      line = list(
        color = "black",
        width = 1.2
      )
    ),
    name = "Medoid"
  ) |>
  layout(
    title = "UMAP: PAM medoids highlighted",
    xaxis = list(title = "UMAP 1"),
    yaxis = list(title = "UMAP 2")
  )

subplot(
  p_pca,
  p_umap,
  nrows = 1,
  margin = 0.05,
  shareX = FALSE,
  shareY = FALSE
)
```
# 6. Cluster profiles

## Voting — worked example

We rescale each party column to $[0,1]$ within each party, then average within
each cluster. This profile highlights which parties are unusually high or low
for each group:

```{r}
party_radar <- party_clustered |>
  mutate(across(Act:Other, rescale)) |>
  group_by(Cluster) |>
  summarise(across(-Electorate, mean),
            n = n(), .groups = "drop") |>
  mutate(Cluster = paste0("Cluster ", Cluster, " (n=", n, ")")) |>
  select(-n)

pr <- party_radar |>
  pivot_longer(-Cluster, names_to = "Party", values_to = "ScaledVote") |>
  ggplot(aes(x = Party, y = ScaledVote, group = Cluster)) +
  geom_col() +
  facet_wrap(~ Cluster) +
  coord_flip() +
  labs(y = "Vote share rescaled within party", x = NULL)

ggplotly(pr, tooltip = "y")
```

## D&D — your turn

Rescale each monster variable within itself (using `scales::rescale`), average
within each cluster, and produce a faceted bar chart. Use your hierarchical
cluster labels from Section 3.

```{r}
# Join cluster labels to monster_data
monster_clustered <- monster_rows |>
  mutate(Cluster = factor(monster_hc_groups))


# Rescale and summarise
monster_radar <- monster_clustered |>
  mutate(across(all_of(monster_vars), rescale)) |>
  group_by(Cluster) |>
  summarise(across(all_of(monster_vars), mean),
            n = n(), .groups = "drop") |>
  mutate(Cluster = paste0("Cluster ", Cluster, " (n=", n, ")")) |>
  select(-n)
```

```{r}
# Faceted bar chart of cluster profiles
monster_radar |>
  pivot_longer(-Cluster, names_to = "Variable", values_to = "ScaledValue") |>
  ggplot(aes(x = Variable, y = ScaledValue, group = Cluster)) +
  geom_col() +
  facet_wrap(~ Cluster) +
  coord_flip() +
  labs(y = "Value rescaled within variable", x = NULL)
```

The profile bars make the split easy to explain: one cluster contains the
higher-HP, higher-CR monsters, while the other covers the broader mid-range.

# 7. DBSCAN: density-based clustering

DBSCAN groups dense points and labels isolated points as **noise**. It has two
parameters: `eps` (neighbourhood radius) and `minPts` (minimum points for a core
point). Unlike $K$-means and PAM, DBSCAN does not need $K$ in advance — it
discovers the number of clusters from the data.

## Voting — worked example

Choose `eps` with a kNN distance plot:

```{r}
dbscan::kNNdistplot(as.matrix(party_scaled), k = 3)
abline(h = 1.7, lty = 2, col = 2)
```
**Note: If you were going to fit minPts to 5 then you would do a kNNdistplot with k = 5 and choose eps accordingly.**

Fit DBSCAN:

```{r}
party_db <- dbscan::dbscan(as.matrix(party_scaled), eps = 1.7, minPts = 3)
table(party_db$cluster)
```

Colour the PCA projection by DBSCAN cluster:

```{r}
party_db_points <- party_pca_points |>
  mutate(cluster = factor(party_db$cluster))

p_db <- party_db_points |>
  ggplot(aes(x = pc1, y = pc2, colour = cluster, label = Electorate)) +
  geom_point(aes(text = paste0(Electorate, "<br>Cluster: ", cluster)),
             alpha = 0.8, size = 2) +
  labs(x = "PC1", y = "PC2", title = "Voting: DBSCAN clusters on PCA") +
  scale_colour_viridis_d()

ggplotly(p_db, tooltip = "text")

```

What happens with different `eps` and `minPts`?

```{r}
try_params <- function(eps_val, minpts_val) {
  db <- dbscan::dbscan(as.matrix(party_scaled), eps = eps_val, minPts = minpts_val)
  labs <- db$cluster
  tibble(
    eps = eps_val,
    minPts = minpts_val,
    n_clusters = length(setdiff(unique(labs), 0L)),
    n_noise = sum(labs == 0L),
    n_total = length(labs)
  )
}

bind_rows(
  try_params(1.5, 5),
  try_params(1, 3),
  try_params(2, 4),
  try_params(1.8, 3),
  try_params(2, 3)
)
```

## D&D — your turn

Apply DBSCAN to the monster data. Use `kNNdistplot` to choose `eps`, fit the
model, tabulate clusters against monster `type`, and experiment with different
parameter values.

```{r}
# kNN distance plot to choose eps
dbscan::kNNdistplot(as.matrix(monster_scaled), k = 4)
abline(h = 1.8, lty = 2, col = 2)
```

```{r}
# Fit DBSCAN and tabulate cluster counts
monster_db <- dbscan::dbscan(as.matrix(monster_scaled), eps = 1.8, minPts = 4)
table(monster_db$cluster)
```

With these settings, most monsters fall into one dense cluster and a smaller set
are treated as noise.

```{r}
# Tabulate DBSCAN clusters against monster type
monster_db_results <- monster_rows |>
  mutate(
    cluster = factor(monster_db$cluster),
    status = if_else(cluster == "0", "noise", "clustered")
  )

monster_db_results |>
  count(cluster, type, sort = TRUE) 

monster_db_results |>
  count(status)
```

```{r}
# Experiment with different eps and minPts values
try_params <- function(eps_val, minpts_val) {
  db <- dbscan::dbscan(as.matrix(monster_scaled), eps = eps_val, minPts = minpts_val)
  labs <- db$cluster
  tibble(
    eps = eps_val,
    minPts = minpts_val,
    n_clusters = length(setdiff(unique(labs), 0L)),
    n_noise = sum(labs == 0L),
    n_total = length(labs)
  )
}

bind_rows(
  try_params(1.2, 3),
  try_params(1.3, 5),
  try_params(1.4, 5),
  try_params(1.25, 4),
  try_params(1.25, 5)
)
```

# 8. Wrap-up

### Final questions

1. Which clustering method gave the most interpretable segmentation of
   the election data? Of the D&D monsters?

2. How important was scaling for the monsters data? Would your
   conclusions change if you used the raw values instead of z-scores?

3. How do party-vote clusters differ from the political narrative you
   expected? Which electorates surprised you?

4. For the monsters, did DBSCAN find a different number of clusters
   than your chosen $K$ from the other methods? What does that tell you
   about the data structure?

5. If you had to explain your final clustering of D&D monsters to
   someone who has never played D&D, which output would you show: a
   dendrogram, cluster means, medoids, silhouette widths, the profile
   bars, or the DBSCAN plot?

### Guide notes

1. For the election data, the cluster summaries and profiles are the easiest way
   to explain the groupings. For the monsters, the $K$-means summaries and the
   profile bars give the quickest overview, while PAM contributes concrete
   example monsters.

2. Scaling matters a lot for the monsters. Without z-scores, `hp_number`, `cr`,
   and `speed_base_number` dominate the distances and the ability scores matter
   much less. Go back and take a look at this lab with the unscaled data to see how much the clusters change.

3. The election data still has a broad mainstream block, but some electorates
   stand out because of distinctive Green or Māori Party support. Those are the
   places most worth checking against your prior expectations.

4. Yes. DBSCAN mostly finds one dense core plus some noise points instead of
   reproducing the same $K$ as the partitioning methods. That suggests the
   monster data has a strong central mass with a tail of more unusual cases. 
   Does this match your intuition from the PCA and UMAP plots? If there we large clusters what would that mean for fair
   game design?

5. If I had to choose one output for a non-player, I would show the silhouette boxplots. 
   They are easy to understand and they show that the two-cluster solution is a much better 
   fit than the alternatives. The profile bars are also good for showing how the clusters differ, 
   but they are more complex to interpret.
