---
title: "Workshop 10: Clustering"
output: html_document
---

In this workshop we follow the lecture workflow for clustering on **two datasets**.
For each task, the **2020 NZ Election** results are shown as a fully worked
example. Your job is to adapt the same code to the **D&D Monsters** data — the
variable names are similar, so the code changes are small.

- build a dissimilarity matrix and check the effect of scaling
- use PCA and UMAP to visualise the data in two dimensions
- compare hierarchical clustering, $K$-means, and PAM
- use silhouette widths to defend a choice of $K$
- profile clusters
- apply density-based clustering with DBSCAN

Use the `tidytuesdayR` package to pull the monsters data from TidyTuesday. If
RStudio prompts you to install `tidytuesdayR`, let it do so before continuing.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(cluster)
library(uwot)
library(dbscan)
library(plotly)
library(tidytuesdayR)
library(scales)


theme_set(theme_minimal())
```

## The two datasets

### 2020 New Zealand General Election

In the 2020 general election each voter casts two votes: a **party vote** (for a
political party) and a **candidate vote** (for a local MP). We cluster the 72
electorates by party vote shares to see how regions group together politically.

```{r import-election-data}
party <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nzelect/2020_party.csv")

party_mat <- party |>
  column_to_rownames("Electorate")

party_scaled <- party_mat |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

party |> summarise(across(-Electorate, mean))
```

### D&D Monsters

Each row is a monster with a `type` label plus combat features like strength,
dexterity, HP, challenge rating, and speed. We cluster on the numeric features
and use `type` afterward to interpret the clusters.

```{r load-monsters}
monsters <- tidytuesdayR::tt_load("2025-05-27")$monsters |>
  tibble::as_tibble()

monster_vars <- c(
  "str", "dex", "con", "int", "wis", "cha",
  "hp_number", "cr", "speed_base_number"
)

monster_data <- monsters |>
  select(type, all_of(monster_vars)) |>
  drop_na()

monster_scaled <- monster_data |>
  select(all_of(monster_vars)) |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))
```

# 1. Distances and scaling

## Voting — worked example

Party vote shares are percentages on a common 0–100 scale. Parties with larger
vote shares (like Labour) naturally have larger variance. We compare raw
Euclidean distance with z-scored distance for the first six electorates:

```{r}
round(as.matrix(dist(party_mat |> slice(1:6))), 2)
round(as.matrix(dist(party_scaled |> slice(1:6))), 2)
```

## D&D — your turn

Compute the raw and z-scored Euclidean distance matrices for the first six
monsters. Which variables appear to dominate the unscaled distances?

```{r}
# Raw distances


# Z-scored distances

```

# 2. Dimension reduction: PCA and UMAP

## Voting — worked example

We build PCA (linear) and UMAP (nonlinear) 2D projections so we can inspect the
electorate structure before clustering.

```{r}

party_pca <- prcomp(party_scaled, center = TRUE, scale. = TRUE)
party_pca_points <- as_tibble(party_pca$x) |>
  select(pc1 = PC1, pc2 = PC2) |>
  (
    function(scores) {
      bind_cols(party |> transmute(Electorate), scores)
    }
  )()
set.seed(8675309)
party_umap_points <- uwot::umap(
  party_scaled,
  n_neighbors = 15,
  min_dist = 0.1,
  n_components = 2,
  metric = "euclidean"
) |>
  as_tibble() |>
  set_names(c("umap1", "umap2")) |>
  (
    function(embedding) {
      bind_cols(party |> transmute(Electorate), embedding)
    }
  )()

p_pca <- ggplot(
  party_pca_points,
  aes(x = pc1, y = pc2, label = Electorate, text = Electorate)
) +
  geom_point() +
  labs(x = "PC1", y = "PC2", title = "Voting: PCA")

p_umap <- ggplot(
  party_umap_points,
  aes(x = umap1, y = umap2, label = Electorate, text = Electorate)
) +
  geom_point() +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Voting: UMAP")

ggplotly(p_pca, tooltip = "text")
ggplotly(p_umap, tooltip = "text")

```

Try to change the seed and see how UMAP changes. 

## D&D — your turn

Fit PCA and UMAP on `monster_scaled`. Use `set.seed(161324)` for UMAP. The code
follows the same pattern as the voting example — swap the data and variable
names. Use `name` as the point identifier.

```{r}
# PCA


# UMAP

```

Plot both projections side-by-side with linked highlighting (`key = name`).

```{r}
# Linked PCA vs UMAP plot

```

# 3. Hierarchical clustering

## Voting — worked example

Average-linkage hierarchical clustering on scaled party data:

```{r}
party_dist <- dist(party_scaled, method = "euclidean")
party_hc <- hclust(party_dist, method = "average")

plot(party_hc, labels = FALSE, hang = -1,
     main = "Average-linkage dendrogram for NZ 2020 party vote")
```

Try single, complete, and average linkage side by side:

```{r}
party_hc_single   <- hclust(party_dist, method = "single")
party_hc_complete <- hclust(party_dist, method = "complete")
party_hc_average  <- hclust(party_dist, method = "average")

plot(party_hc_single,   labels = FALSE, hang = -1, main = "Single linkage")
plot(party_hc_complete, labels = FALSE, hang = -1, main = "Complete linkage")
plot(party_hc_average,  labels = FALSE, hang = -1, main = "Average linkage")
```

Use the silhouette measure to choose $K$, then cut the tree:

```{r}
library(factoextra)

party_scaled |>
  fviz_nbclust(FUNcluster = factoextra::hcut, hc_method = "average", k.max = 10,
               method = "silhouette") |>
  ggplotly()

party_groups <- cutree(party_hc, k = 2)
```

Cluster means:

```{r}
party_clustered <- party |>
  mutate(Cluster = factor(party_groups))

party_clustered |>
  group_by(Cluster) |>
  summarise(across(-Electorate, mean), n_electorates = n()) |>
  arrange(desc(n_electorates))
```

Which electorates are in each cluster?

```{r}
party_clustered |>
  select(Electorate, Cluster) |>
  arrange(Cluster, Electorate) |>
  print(n = 72)
```

## D&D — your turn

Fit hierarchical clustering on `monster_scaled`. Try single, complete, and
average linkage. Which gives the most balanced cluster sizes?

```{r}
# Distance matrix


# Three linkage methods

```

```{r}
# Silhouette to choose K


# Cut the tree at your chosen K

```

```{r}
# Cluster means by monster type (hint: join back to monster_data)

```

# 4. $K$-means and the elbow heuristic

## Voting — worked example

```{r}
set.seed(67)
party_elbow <- tibble(k = 1:10) |>
  mutate(
    model = map(k, ~ kmeans(party_scaled, centers = .x, nstart = 20)),
    fit = map(model, glance)
  ) |>
  unnest(fit)

party_elbow
```

```{r}
p<-party_elbow |>
  ggplot(aes(x = k, y = tot.withinss)) +
  geom_line() +
  geom_point(aes(text = paste0("k = ", k, "<br>Total WSS = ", round(tot.withinss))), size = 2)

ggplotly(p, tooltip = "text")
```

Fit $K$-means with $k = 6$:

```{r}
set.seed(67)
party_km6 <- kmeans(party_scaled, centers = 6, nstart = 25)

party |>
  mutate(Cluster = factor(party_km6$cluster)) |>
  group_by(Cluster) |>
  summarise(across(-Electorate, mean), n = n()) |>
  arrange(desc(n))
```

Let's plot this out on the PCA projection:
```{r}
# no legend, just colour by cluster
party_km6_points <- party_pca_points |>
  mutate(Cluster = factor(party_km6$cluster))

p_km_pca <- party_km6_points |>
  ggplot(aes(x = pc1, y = pc2, colour = Cluster,
             label = Electorate, key = Electorate)) +
  geom_point(aes(text = Electorate), alpha = 0.8, size
             = 2) +
  labs(x = "PC1", y = "PC2", title = "Voting:
K-means clusters on PCA")

ggplotly(p_km_pca, tooltip = "text")

# lets add a UMAP cluster
party_umap_km6_points <- party_umap_points |>
  mutate(Cluster = factor(party_km6$cluster))


p_km_umap <- party_umap_km6_points |>
  ggplot(aes(x = umap1, y = umap2, colour = Cluster,
             label = Electorate, key = Electorate)) +
  geom_point(aes(text = Electorate), alpha = 0.8, size
             = 2) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Voting:
K-means")

ggplotly(p_km_umap, tooltip = "text")


```


## D&D — your turn

Fit $K$-means for several values of $k$, produce an elbow plot, then fit a final
model with your chosen $k$.

```{r}
# Elbow tibble


# Elbow plot

```

```{r}
# Final K-means fit and cluster means

# PCA plot coloured by K-means cluster

```

# 5. PAM and silhouette widths

## Voting — worked example

PAM fits $K$-medoids for $K = 2$ through $K = 8$. We compare average silhouette
widths:

```{r}
pam_sil_summary <- map_dfr(2:8, function(k_value) {
  pam_fit <- pam(party_scaled, k = k_value)
  sil <- silhouette(pam_fit)
  widths <- sil[, "sil_width"]
  tibble(
    k = k_value,
    avg_silhouette = mean(widths)
  )
})

pam_sil_summary |>
  arrange(desc(avg_silhouette))
```

Fit PAM with $k = 4$ and produce the silhouette plot:

```{r}
party_pam4 <- pam(party_scaled, k = 4)
plot(silhouette(party_pam4), border = NA, do.clus.stat = FALSE, do.n.k = FALSE)
```

Per-cluster average silhouette widths and boxplots:

```{r}
party_pam_sil <- party_pam4 |>
  pluck("silinfo", "widths") |>
  as_tibble()

party_pam_sil |>
  group_by(cluster) |>
  summarise(avg_sil = mean(sil_width), n_electorates = n())

pps<- party_pam_sil |>
  ggplot(aes(x = factor(cluster), y = sil_width)) +
  geom_boxplot() +
  labs(x = "Cluster", y = "Silhouette width")
  
ggplotly(pps, tooltip = "y")
```

Which electorates are the medoids?

```{r}
medoid_rows <- party_pam4$id.med
party_mat |>
  slice(medoid_rows) |>
  rownames_to_column("Electorate") |>
  mutate(Cluster = 1:4) |>
  select(Cluster, Electorate, everything())

party_medoid_ids <- rownames(party_mat)[medoid_rows]

pca_medoid_plot <- party_pca_points |>
  mutate(
    is_medoid = Electorate %in% party_medoid_ids,
    hover = paste0(
      Electorate,
      "<br>Medoid: ", if_else(is_medoid, "Yes", "No")
    )
  ) |>
  ggplot(aes(x = pc1, y = pc2)) +
  geom_point(
    data = ~ filter(.x, !is_medoid),
    aes(text = hover),
    alpha = 0.5,
    size = 2,
    colour = "grey50"
  ) +
  geom_point(
    data = ~ filter(.x, is_medoid),
    aes(text = hover),
    alpha = 1,
    size = 4,
    shape = 21,
    stroke = 1.2,
    fill = "#2C7FB8",
    colour = "black"
  ) +
  labs(
    x = "PC1",
    y = "PC2",
    title = "PCA: PAM medoids highlighted"
  )

umap_medoid_plot <- party_umap_points |>
  mutate(
    is_medoid = Electorate %in% party_medoid_ids,
    hover = paste0(
      Electorate,
      "<br>Medoid: ", if_else(is_medoid, "Yes", "No")
    )
  ) |>
  ggplot(aes(x = umap1, y = umap2)) +
  geom_point(
    data = ~ filter(.x, !is_medoid),
    aes(text = hover),
    alpha = 0.5,
    size = 2,
    colour = "grey50"
  ) +
  geom_point(
    data = ~ filter(.x, is_medoid),
    aes(text = hover),
    alpha = 1,
    size = 4,
    shape = 21,
    stroke = 1.2,
    fill = "#2C7FB8",
    colour = "black"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    title = "UMAP: PAM medoids highlighted"
  )

subplot(
  ggplotly(pca_medoid_plot, tooltip = "text"),
  ggplotly(umap_medoid_plot, tooltip = "text"),
  nrows = 1
)
```

## D&D — your turn

Fit PAM for $K = 2$ through $K = 8$, identify the best $K$ by average silhouette
width, produce the silhouette plot, boxplots, and extract the medoids.

```{r}
# PAM silhouette summary for K = 2 to 8

```

```{r}
# PAM with best K, silhouette plot

```

```{r}
# Per-cluster silhouettes and boxplots

```

```{r}
# Medoid monsters

```

# 6. Cluster profiles

## Voting — worked example

We rescale each party column to $[0,1]$ within each party, then average within
each cluster. This profile highlights which parties are unusually high or low
for each group:

```{r}
party_radar <- party_clustered |>
  mutate(across(Act:Other, rescale)) |>
  group_by(Cluster) |>
  summarise(across(-Electorate, mean),
            n = n(), .groups = "drop") |>
  mutate(Cluster = paste0("Cluster ", Cluster, " (n=", n, ")")) |>
  select(-n)

pr <- party_radar |>
  pivot_longer(-Cluster, names_to = "Party", values_to = "ScaledVote") |>
  ggplot(aes(x = Party, y = ScaledVote, group = Cluster)) +
  geom_col() +
  facet_wrap(~ Cluster) +
  coord_flip() +
  labs(y = "Vote share rescaled within party", x = NULL)

ggplotly(pr, tooltip = "y")
```

## D&D — your turn

Rescale each monster variable within itself (using `scales::rescale`), average
within each cluster, and produce a faceted bar chart. Use your hierarchical
cluster labels from Section 3.

```{r}
# Join cluster labels to monster_data


# Rescale and summarise

```

```{r}
# Faceted bar chart of cluster profiles

```

# 7. DBSCAN: density-based clustering

DBSCAN groups dense points and labels isolated points as **noise**. It has two
parameters: `eps` (neighbourhood radius) and `minPts` (minimum points for a core
point). Unlike $K$-means and PAM, DBSCAN does not need $K$ in advance — it
discovers the number of clusters from the data.

## Voting — worked example

Choose `eps` with a kNN distance plot:

```{r}
dbscan::kNNdistplot(as.matrix(party_scaled), k = 3)
abline(h = 1.7, lty = 2, col = 2)
```
**Note: If you were going to fit minPts to 5 then you would do a kNNdistplot with k = 5 and choose eps accordingly.**

Fit DBSCAN:

```{r}
party_db <- dbscan::dbscan(as.matrix(party_scaled), eps = 1.7, minPts = 3)
table(party_db$cluster)
```

Colour the PCA projection by DBSCAN cluster:

```{r}
party_db_points <- party_pca_points |>
  mutate(cluster = factor(party_db$cluster))

p_db <- party_db_points |>
  ggplot(aes(x = pc1, y = pc2, colour = cluster, label = Electorate)) +
  geom_point(aes(text = paste0(Electorate, "<br>Cluster: ", cluster)),
             alpha = 0.8, size = 2) +
  labs(x = "PC1", y = "PC2", title = "Voting: DBSCAN clusters on PCA") +
  scale_colour_viridis_d()

ggplotly(p_db, tooltip = "text")

```

What happens with different `eps` and `minPts`?

```{r}
try_params <- function(eps_val, minpts_val) {
  db <- dbscan::dbscan(as.matrix(party_scaled), eps = eps_val, minPts = minpts_val)
  labs <- db$cluster
  tibble(
    eps = eps_val,
    minPts = minpts_val,
    n_clusters = length(setdiff(unique(labs), 0L)),
    n_noise = sum(labs == 0L),
    n_total = length(labs)
  )
}

bind_rows(
  try_params(1.5, 5),
  try_params(1, 3),
  try_params(2, 4),
  try_params(1.8, 3),
  try_params(2, 3)
)
```

## D&D — your turn

Apply DBSCAN to the monster data. Use `kNNdistplot` to choose `eps`, fit the
model, tabulate clusters against monster `type`, and experiment with different
parameter values.

```{r}
# kNN distance plot to choose eps

```

```{r}
# Fit DBSCAN and tabulate cluster counts

```

```{r}
# Tabulate DBSCAN clusters against monster type

```

```{r}
# Experiment with different eps and minPts values

```

# 8. Wrap-up

### Final questions

1. Which clustering method gave the most interpretable segmentation of
   the election data? Of the D&D monsters?

2. How important was scaling for the monsters data? Would your
   conclusions change if you used the raw values instead of z-scores?

3. How do party-vote clusters differ from the political narrative you
   expected? Which electorates surprised you?

4. For the monsters, did DBSCAN find a different number of clusters
   than your chosen $K$ from the other methods? What does that tell you
   about the data structure?

5. If you had to explain your final clustering of D&D monsters to
   someone who has never played D&D, which output would you show: a
   dendrogram, cluster means, medoids, silhouette widths, the profile
   bars, or the DBSCAN plot?