---
title: "Workshop 11: Association Rule Mining"
output: html_document
---

In this workshop we build association rules from a Recipe
dataset, then finish with a shorter extension on the local bank-service
data. The main goals are to:

- turn text data into transactions
- inspect item frequencies
- mine rules with `apriori()`
- interpret support, confidence, and lift
- filter rules using `subset()`

We will use the same recipe dataset as in the lecture, but we will do more cleaning to make the rules more interpretable. 
The code for that cleaning is a bit long, but it is mostly just a series of string manipulations. 
You can skim it if you like, or even skip. The main thing to notice is that we are trying to remove any words that are not really part of the ingredient name, 
such as quantities, units, and preparation instructions. We also standardize some common ingredients to avoid having multiple variants of the same item.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(arules)

theme_set(theme_minimal())
```

## 1. Recipe ingredients as transactions

The lecture used a TidyTuesday dataset of recipes grouped by cuisine.
Each recipe will be treated as one transaction, and each ingredient as
an item.

```{r}
cuisines_raw <- read_csv("https://knowlton.co.nz/161324/data/cuisines.csv.gz",
                          show_col_types = FALSE)

cuisines_raw |> glimpse()
```

The `ingredients` column is a comma-separated string, so we need to turn
it into one ingredient per row before creating the transaction matrix.

```{r}
not_ingredients <- c(
  "chopped", "minced", "diced", "sliced", "grated", "shredded", "crushed",
  "julienned", "cubed", "halved", "quartered", "peeled", "seeded", "stemmed",
  "trimmed", "cleaned", "rinsed", "drained", "thawed", "softened", "melted",
  "beaten", "divided", "separated", "sifted", "crumbled", "toasted",
  "mashed", "pureed", "blanched", "boiled", "roasted", "juiced",
  "thinly sliced", "finely chopped", "coarsely chopped",
  "finely", "coarsely", "thinly",
  "to taste", "as needed", "or to taste", "or as needed",
  "or more", "or more to taste", "or more as needed",
  "if desired", "for garnish",
  "skinless", "cooking spray", "room temperature", "needed", "taste"
)

unit_rx <- regex(
  paste0(
    "^(cups?|tablespoons?|tbsp|teaspoons?|tsp|pounds?|lbs?|ounces?|oz|",
    "cans?|packages?|pkgs?|envelopes?|pinch(es)?|cloves?|slices?|",
    "stalks?|heads?|bunche?s?|pieces?|strips?|sprigs?)\\s+(of\\s+)?"
  ),
  ignore_case = TRUE
)

recipe_items <- cuisines_raw |>
  select(name, country, ingredients) |>
  mutate(recipe_id = row_number()) |>
  separate_longer_delim(ingredients, delim = ", ") |>
  mutate(
    ingredient = ingredients |>
      str_to_lower() |>
      str_trim() |>
      str_remove_all("\\(.*?\\)") |>
      str_remove("^[0-9\\u00bd\\u00bc\\u00be\\u2153\\u2154\\u215b\\u215c\\u215d\\u215e/\\. ]+") |>
      str_remove(unit_rx) |>
      str_remove(regex("^(large|medium|small|extra-?large)\\s+", ignore_case = TRUE)) |>
      str_remove(regex("\\s+(to taste|as needed|if desired|or more to taste|or more)$",
                       ignore_case = TRUE)) |>
      str_remove(regex(",\\s*(divided|optional|drained|thawed|softened|melted|beaten)$",
                       ignore_case = TRUE)) |>
      str_trim() |>
      str_replace_all("^salt$", "salt") |>
      str_replace_all("^unsalted butter$", "butter") |>
      str_replace_all("^garlic powder$", "garlic") |>
      str_replace_all("^minced garlic$", "garlic") |>
      str_replace_all("^eggs$", "egg") |>
      str_replace_all("^onions$", "onion") |>
      str_replace_all("^potatoes$", "potato") |>
      str_replace_all("^tomatoes$", "tomato") |>
      str_replace_all("^carrots$", "carrot") |>
      str_replace_all("^lemons$", "lemon") |>
      str_replace_all("^limes$", "lime") |>
      str_replace_all("^mushrooms$", "mushroom") |>
      str_replace_all("^bananas$", "banana") |>
      str_replace_all("^green onions$", "green onion") |>
      str_replace_all("^cloves$", "clove")
  ) |>
  filter(
    nchar(ingredient) > 2,
    !ingredient %in% not_ingredients
  )
```

### Try yourself

1. Count the most common ingredients in `recipe_items`.

2. Do the common ingredients seem globally useful, or are some likely to
   inflate confidence without being especially informative?

3. How might that affect your later interpretation of lift?

```{r}
recipe_items |>
  count(ingredient, sort = TRUE) |>
  slice_head(n = 20)
```

## 2. Converting to a transaction database

To keep the rules interpretable, we restrict attention to the 75 most
common ingredients.

```{r}
top_ingredients <- recipe_items |>
  count(ingredient, sort = TRUE) |>
  slice_head(n = 75) |>
  pull(ingredient)

recipe_items_filtered <- recipe_items |>
  filter(ingredient %in% top_ingredients)

recipe_trans <- recipe_items |>
  filter(ingredient %in% top_ingredients) |>
  
  # Remove duplicate ingredient entries within the same recipe.
  # This ensures each recipe contributes each ingredient at most once.
  distinct(recipe_id, ingredient) |>
  
  # For each recipe, collapse its ingredients into a list-column.
  # Each row now represents one recipe and contains a vector of its ingredients.
  summarise(items = list(ingredient), .by = recipe_id) |>
  
  # Extract the list-column, giving a plain list where each element is
  # the ingredient vector for one recipe.
  pull(items) |>
  
  # Convert the list of ingredient vectors into an arules transactions object.
  as("transactions")

recipe_trans
```

### Try yourself

1. Inspect the first ten transactions using `inspect()`.

2. Compare `inspect()` with `DATAFRAME()`. Which would you use for
   exploratory work?

## 3. Item frequencies

```{r}
itemFrequencyPlot(recipe_trans, topN = 25, type = "relative")
```

### Try yourself

1. Recreate the item-frequency plot using `ggplot2`.

2. Which ingredients are common enough that a high-confidence rule with
   that ingredient on the right-hand side might still be uninteresting?

## 4. Mining rules with Apriori

We now mine rules with moderate support and confidence thresholds.

```{r}
recipe_rules <- apriori(
  recipe_trans,
  parameter = list(support = 0.02, confidence = 0.20, minlen = 1, maxlen = 4),
  control = list(verbose = FALSE)
)

```

### Try yourself

1. How many rules were found?

2. Inspect the top ten rules by support, by confidence, and by lift.

3. Which ranking would you use first if the goal is to find genuinely
   informative associations rather than merely common ones?


## 5. Misleading high-confidence rules

High confidence, which we define as more than 50% of recipes, on its own can be misleading when the consequent is very
common. The code below looks for rules with high confidence but lift close to one.

```{r}
recipe_rules |>
  subset(confidence >= 0.5 & lift < 1.2) |>
  sort(by = "confidence") |>
  inspect()
```

These rules are not necessarily wrong, but they are not especially informative. In the table above only `salt` appears on the RHS of the rules, and it is present in >40% of recipes. So any rule that has `salt` on the RHS is likely to have high confidence, but that does not necessarily indicate a strong association between the LHS and `salt`. The lift values close to 1 suggest that the presence of the LHS ingredients does not significantly increase the likelihood of `salt` being present compared to its overall frequency in the dataset.

## 6. Filtering rules

Use `subset()` together with `%in%`, `%ain%`, and `%pin%` to focus on a
particular ingredient or collection of ingredients.

```{r}
garlic_rules <- recipe_rules |>
  subset(lhs %in% "garlic" & lift > 1.1)

garlic_rules |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```
Use the '%pin%' operator to find rules that contain a particular ingredient as
part of a longer item name. For example, the code below finds rules that contain 'pepper' as part of an ingredient name, which could include 'black pepper', 'red pepper', 'green pepper', etc. This allows us to capture a broader range of rules that involve any type of pepper, rather than just those that match 'pepper' exactly.

```{r}

pepper_rules <- recipe_rules |>
  subset(lhs %pin% "pepper" & lift > 1.1)

pepper_rules |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()

```
Use the '%ain%' operator to find rules that contain all of a specified set of ingredients on the left-hand side. For example, the code below finds rules where both 'garlic' and 'onion' appear on the left-hand side, regardless of what other ingredients may also be present. This allows us to focus on associations that involve the combination of garlic and onion together.

```{r}
garlic_onion_rules <- recipe_rules |>
  subset(lhs %ain% c("garlic", "olive oil") & lift > 1.1)

garlic_onion_rules |>
  sort(by = "lift") |>
  inspect()
```

### Try yourself

1. Find rules with `salt` on the left-hand side.

2. Find rules where both `garlic` and `onion` appear on the left-hand
   side.

3. Find one ingredient whose rules look cuisine-specific rather than
   globally common.

## 7. Optional interactive inspection

Install `arulesViz` so that you can use an interactive inspection of your rule set.

```{r, eval=FALSE}
library(arulesViz)

recipe_rules |>
  sort(by = "lift") |>
  head(n = 100) |>
  inspectDT()
```

## 8. More Practice: bank-service rules

The following bank dataset contains a set of abbreviated service labels that indicate the types of financial products or services associated with each customer. The label `ATM` refers to ATM access or an ATM card service, while `AUTO` indicates an auto loan. The label `CCRD` represents a credit card, and `CD` refers to a certificate of deposit, also known in some banking systems as a term deposit or fixed deposit. The label `CKCRD` most plausibly refers to a cheque card or debit card linked to a chequing account, while `CKING` indicates a chequing account. The label `HMEQLC` refers to a home equity line of credit, and `IRA` indicates an individual retirement account. The label `MMDA` represents a money market deposit account, while `MTG` refers to a mortgage. The label `PLOAN` indicates a personal loan, `SVG` refers to a savings account, and `TRUST` represents a trust account or trust-related banking service.

These labels can be interpreted as indicators of customer engagement with different banking products. Deposit-related products include chequing accounts, savings accounts, money market deposit accounts, certificates of deposit, and retirement accounts. Lending products include auto loans, mortgages, home equity lines of credit, personal loans, and credit cards. Card and transaction access services include ATM access and cheque or debit cards. Trust services are distinct from ordinary deposit and lending products because they generally involve fiduciary, estate, or asset-management arrangements rather than standard retail banking activity.

```{r}
bank <- read_csv("https://knowlton.co.nz/161324/data/bank-rules.csv.gz", show_col_types = FALSE) |>
  transactions(format = "long", cols = c("ACCOUNT", "SERVICE"))

bank_rules <- apriori(
  bank,
  parameter = list(support = 0.005, confidence = 0.4),
  control = list(verbose = FALSE)
)
```

### Try yourself

1. Find rules with `CCRD` on the right-hand side. Which services look
   useful for cross-selling a credit card?

2. Find rules that contain both `MTG` and `HMEQLC`.

3. Compare the bank rules with the recipe rules. Which dataset gives
   more obviously actionable associations?


### Final questions

1. In the recipe data, what kinds of items produce high confidence but
   uninformative rules?

2. Which filtering tools in `arules` did you find most useful?

3. If you had to recommend one metric for ranking candidate rules in an
   exploratory analysis, would you choose support, confidence, or lift?
