---
title: "Workshop 11 Guide: Clustering"
output: html_document
---

This guide works through one complete clustering analysis using the same
flow as `lab10.Rmd`. There is still room for judgment, especially when
choosing the number of clusters, but the code below gives one sensible
solution path.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(cluster)
source(file.path(dirname(knitr::current_input(dir = TRUE)), "..", "common", "tidytuesday.R"))
source(file.path(dirname(knitr::current_input(dir = TRUE)), "..", "common", "ggplot_utils.R"))

theme_set(theme_minimal())
doc_dir <- dirname(knitr::current_input(dir = TRUE))
data_dir <- resolve_data_dir(doc_dir)

qatar_url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-12-09/qatarcars.csv"
monster_url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-05-27/monsters.csv"

qatarcars <- read_tidytuesday_cached(qatar_url, "qatarcars.parquet", data_dir = data_dir, csv_file = "qatarcars.csv")
monsters <- read_tidytuesday_cached(monster_url, "monsters.parquet", data_dir = data_dir, csv_file = "monsters.csv")
```

## 1. Warm-up: scaling matters

```{r}
car_vars <- c(
  "length", "width", "height", "mass", "horsepower",
  "price", "performance", "economy", "trunk", "seating"
)

cars <- qatarcars |>
  select(type, enginetype, all_of(car_vars)) |>
  drop_na() |>
  mutate(car_id = row_number(), .before = 1)

cars_raw <- cars |>
  select(all_of(car_vars))

cars_scaled <- cars_raw |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

summary(cars_raw)
```

The raw summaries show that variables such as `price`, `mass`, and
`horsepower` are on much larger scales than `seating` or `height`. If
we use Euclidean distance directly, those larger-scale variables will
dominate the dissimilarities.

```{r}
round(as.matrix(dist(cars_raw |> slice(1:6))), 2)
round(as.matrix(dist(cars_scaled |> slice(1:6))), 2)
```

The scaled distance matrix is the more appropriate input for the
clustering methods used below.

## 2. Hierarchical clustering

```{r}
cars_dist <- dist(cars_scaled, method = "euclidean")
cars_hc <- hclust(cars_dist, method = "complete")

plot(cars_hc, labels = FALSE, hang = -1, main = "Complete-linkage dendrogram for Qatar Cars")
```

```{r}
cars_hc4 <- cars |>
  mutate(cluster = factor(cutree(cars_hc, k = 4)))

cars_hc5 <- cars |>
  mutate(cluster = factor(cutree(cars_hc, k = 5)))

cars_hc4 |>
  count(cluster, type, sort = TRUE)

cars_hc4 |>
  count(cluster, enginetype, sort = TRUE)
```

There is no unique correct cut, but `k = 4` is a reasonable working
choice because it keeps the interpretation manageable while still
separating noticeably different car groups.

## 3. $K$-means and the elbow plot

```{r}
set.seed(161324)
car_elbow <- tibble(k = 1:8) |>
  mutate(
    model = map(k, ~ kmeans(cars_scaled, centers = .x, nstart = 20)),
    fit = map(model, glance)
  ) |>
  unnest(fit)

car_elbow
```

```{r}
compose_ggplot(
  ggplot(car_elbow, aes(x = k, y = tot.withinss)),
  geom_line(),
  geom_point(size = 2)
)
```

One reasonable reading of the elbow plot is that the main gains occur by
about `k = 4`, so we continue with that choice.

```{r}
set.seed(161324)
cars_km4 <- kmeans(cars_scaled, centers = 4, nstart = 25)

cars |>
  mutate(cluster = factor(cars_km4$cluster)) |>
  count(cluster, type, sort = TRUE)
```

```{r}
cars |>
  mutate(cluster = factor(cars_km4$cluster)) |>
  group_by(cluster) |>
  summarise(across(all_of(car_vars), mean), n = n())
```

The cluster summaries on the original scale are usually easier to
interpret than the scaled cluster centres.

## 4. PAM and silhouette widths

```{r}
cars_pam4 <- pam(cars_scaled, k = 4)
plot(silhouette(cars_pam4), border = NA, do.clus.stat = FALSE, do.n.k = FALSE)
```

```{r}
cars_pam_sil <- cars_pam4 |>
  pluck("silinfo", "widths") |>
  as_tibble()

cars_pam_sil |>
  group_by(cluster) |>
  summarise(avg_sil = mean(sil_width), n = n())
```

High average silhouette widths indicate that the cluster is reasonably
well separated from the others. Low or negative widths would suggest a
poor fit or an artificial split.

## 5. Comparison with D&D Monsters

```{r}
monster_vars <- c(
  "str", "dex", "con", "int", "wis", "cha",
  "hp_number", "cr", "speed_base_number"
)

monster_data <- monsters |>
  select(type, all_of(monster_vars)) |>
  drop_na()

monster_scaled <- monster_data |>
  select(all_of(monster_vars)) |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

monster_hc <- monster_scaled |>
  dist(method = "euclidean") |>
  hclust(method = "complete")

monster_groups <- cutree(monster_hc, k = 5)

monster_data |>
  mutate(cluster = factor(monster_groups)) |>
  count(cluster, type, sort = TRUE)
```

This remains an unsupervised problem. The known `type` values are used
only after fitting, to help interpret whether the clustering has picked
up recognizable structure.

## 6. Optional DBSCAN extension

```{r, eval=FALSE}
library(dbscan)

cars_2d <- cars |>
  select(price, horsepower) |>
  drop_na() |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

dbscan::kNNdistplot(as.matrix(cars_2d), k = 4)
abline(h = 0.6, lty = 2, col = 2)

cars_db <- dbscan::dbscan(as.matrix(cars_2d), eps = 0.6, minPts = 5)

bind_cols(cars_2d, cluster = factor(cars_db$cluster)) |>
  (\(data) compose_ggplot(
    ggplot(data, aes(x = price, y = horsepower, colour = cluster)),
    geom_point(alpha = 0.8, size = 2)
  ))()
```

The DBSCAN extension is mainly about parameter choice and recognising
that density-based clustering answers a different question from
hierarchical clustering or $K$-means.

## Wrap-up

One defensible conclusion is:

- scaling is essential before Euclidean clustering on the Qatar Cars data
- `k = 4` gives a practical working partition across several methods
- hierarchical clustering is useful for exploration
- $K$-means and PAM give clearer fixed partitions once a candidate `K`
  is on the table
