---
title: "Workshop 10: Clustering"
output: html_document
---

In this workshop we follow the lecture workflow for clustering:

- build a dissimilarity matrix
- check the effect of scaling
- compare hierarchical clustering, $K$-means, and PAM
- use silhouette widths to defend a choice of $K$
- compare the final results with a second dataset

The main dataset is the Qatar Cars TidyTuesday data. The D&D Monsters
data are used later as a shorter comparison exercise. Use the
`tidytuesdayR` package to pull both datasets directly from TidyTuesday.
If RStudio prompts you to install `tidytuesdayR`, let it do so before
continuing.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(cluster)

theme_set(theme_minimal())

load_tidytuesday_dataset <- function(tt_date, dataset_name, parquet_url) {
  if (requireNamespace("tidytuesdayR", quietly = TRUE)) {
    tt_data <- tryCatch(
      suppressMessages(tidytuesdayR::tt_load(tt_date)),
      error = function(e) NULL
    )

    if (!is.null(tt_data) &&
        dataset_name %in% names(tt_data) &&
        !is.null(tt_data[[dataset_name]])) {
      return(tt_data[[dataset_name]] |> tibble::as_tibble())
    }
  }

  if (!requireNamespace("arrow", quietly = TRUE)) {
    stop(
      "Install the 'tidytuesdayR' package (RStudio will usually prompt you), ",
      "or install 'arrow' to use the backup data file.",
      call. = FALSE
    )
  }

  parquet_file <- tempfile(fileext = ".parquet")
  on.exit(unlink(parquet_file), add = TRUE)
  download_status <- tryCatch(
    utils::download.file(parquet_url, parquet_file, mode = "wb", quiet = TRUE),
    error = function(e) 1L
  )

  if (!identical(download_status, 0L) || !file.exists(parquet_file)) {
    stop(
      "Could not load the dataset from TidyTuesday or the course backup file.",
      call. = FALSE
    )
  }

  arrow::read_parquet(parquet_file) |> tibble::as_tibble()
}

compose_ggplot <- function(plot, ...) {
  layers <- list(...)

  for (i in seq_along(layers)) {
    plot <- ggplot2:::add_ggplot(plot, layers[[i]], paste0("layer_", i))
  }

  plot
}

qatarcars <- load_tidytuesday_dataset(
  tt_date = "2025-12-09",
  dataset_name = "qatarcars",
  parquet_url = "https://knowlton.co.nz/161324/data/qatarcars.parquet"
)
monsters <- load_tidytuesday_dataset(
  tt_date = "2025-05-27",
  dataset_name = "monsters",
  parquet_url = "https://knowlton.co.nz/161324/data/monsters.parquet"
)
```

## 1. Warm-up: distances and scaling

For the clustering analysis we will use the following numeric variables:

```{r}
car_vars <- c(
  "length", "width", "height", "mass", "horsepower",
  "price", "performance", "economy", "trunk", "seating"
)

cars <- qatarcars |>
  select(type, enginetype, all_of(car_vars)) |>
  drop_na() |>
  mutate(car_id = row_number(), .before = 1)

cars_raw <- cars |>
  select(all_of(car_vars))

cars_scaled <- cars_raw |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

cars |> glimpse()
```

### Try yourself

1. Use `summary()` on `cars_raw`. Which variables appear to be on very
   different scales?

2. Compute the Euclidean distance matrix for the first six cars before
   scaling and after scaling.

3. Which variables do you expect to dominate the unscaled distances?

```{r}
round(as.matrix(dist(cars_raw |> slice(1:6))), 2)
round(as.matrix(dist(cars_scaled |> slice(1:6))), 2)
```

## 2. Hierarchical clustering on Qatar Cars

We start with complete-linkage hierarchical clustering on the scaled
data.

```{r}
cars_dist <- dist(cars_scaled, method = "euclidean")
cars_hc <- hclust(cars_dist, method = "complete")

plot(cars_hc, labels = FALSE, hang = -1, main = "Complete-linkage dendrogram for Qatar Cars")
```

### Try yourself

1. Cut the dendrogram into `k = 4` and `k = 5` clusters.

2. For each choice of `k`, compare the resulting clusters with `type`
   and `enginetype`.

3. Which value of `k` looks more interpretable? There is no unique
   answer here, but for the rest of the worksheet continue with
   `k = 4` unless you have a strong case for another value.

```{r}
cars_hc4 <- cars |>
  mutate(cluster = factor(cutree(cars_hc, k = 4)))

cars_hc5 <- cars |>
  mutate(cluster = factor(cutree(cars_hc, k = 5)))

cars_hc4 |>
  count(cluster, type, sort = TRUE)

cars_hc5 |>
  count(cluster, type, sort = TRUE)
```

## 3. $K$-means and the elbow heuristic

Next fit $K$-means for several values of $K$ and inspect the total
within-cluster sum of squares.

```{r}
set.seed(161324)
car_elbow <- tibble(k = 1:8) |>
  mutate(
    model = map(k, ~ kmeans(cars_scaled, centers = .x, nstart = 20)),
    fit = map(model, glance)
  ) |>
  unnest(fit)

car_elbow
```

### Try yourself

1. Produce an elbow plot of `tot.withinss` against `k`.

2. Where does the main drop occur?

3. Fit a final $K$-means model with `K = 4` and compare the clusters
   with `type` and `enginetype`.

```{r}
compose_ggplot(
  ggplot(car_elbow, aes(x = k, y = tot.withinss)),
  geom_line(),
  geom_point(size = 2)
)
```

```{r}
set.seed(161324)
cars_km4 <- kmeans(cars_scaled, centers = 4, nstart = 25)

cars |>
  mutate(cluster = factor(cars_km4$cluster)) |>
  count(cluster, type, sort = TRUE)
```

### Try yourself

1. Summarise the original car variables within each $K$-means cluster.

2. Which variables seem to separate the clusters most clearly?

3. Do the $K$-means clusters tell a similar story to the hierarchical
   clusters?

```{r}
cars |>
  mutate(cluster = factor(cars_km4$cluster)) |>
  group_by(cluster) |>
  summarise(across(all_of(car_vars), mean), n = n())
```

## 4. PAM and silhouette widths

PAM lets us keep the clustering on the same scaled variables, but with a
medoid rather than a centroid.

```{r}
cars_pam4 <- pam(cars_scaled, k = 4)
plot(silhouette(cars_pam4), border = NA, do.clus.stat = FALSE, do.n.k = FALSE)
```

### Try yourself

1. Extract the silhouette widths from `cars_pam4`.

2. Compute the average silhouette width in each cluster.

3. Compare the overall clustering quality from PAM with your earlier
   interpretation of the hierarchical and $K$-means fits.

```{r}
cars_pam_sil <- cars_pam4 |>
  pluck("silinfo", "widths") |>
  as_tibble()

cars_pam_sil |>
  group_by(cluster) |>
  summarise(avg_sil = mean(sil_width), n = n())
```

## 5. Short comparison: D&D Monsters

The lecture used the D&D Monsters data as a second example because the
known `type` variable gives us a rough truth signal after clustering.

```{r}
monster_vars <- c(
  "str", "dex", "con", "int", "wis", "cha",
  "hp_number", "cr", "speed_base_number"
)

monster_data <- monsters |>
  select(type, all_of(monster_vars)) |>
  drop_na()

monster_scaled <- monster_data |>
  select(all_of(monster_vars)) |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

monster_hc <- monster_scaled |>
  dist(method = "euclidean") |>
  hclust(method = "complete")

monster_groups <- cutree(monster_hc, k = 5)
```

### Try yourself

1. Tabulate the 5-cluster monster solution against `type`.

2. Which monster types are concentrated in one cluster, and which are
   spread across several clusters?

3. Why is this still an unsupervised task even though we are using
   `type` for interpretation afterwards?

```{r}
monster_data |>
  mutate(cluster = factor(monster_groups)) |>
  count(cluster, type, sort = TRUE)
```

## 6. Optional challenge: DBSCAN on a two-variable view

If you have the `dbscan` package available, try a density-based
clustering on two scaled variables. This is only a low-dimensional
projection, so do not expect it to match the earlier results exactly.

```{r, eval=FALSE}
library(dbscan)

cars_2d <- cars |>
  select(price, horsepower) |>
  drop_na() |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

dbscan::kNNdistplot(as.matrix(cars_2d), k = 4)
abline(h = 0.6, lty = 2, col = 2)

cars_db <- dbscan::dbscan(as.matrix(cars_2d), eps = 0.6, minPts = 5)

cars_db_plot <- bind_cols(cars_2d, cluster = factor(cars_db$cluster))

compose_ggplot(
  ggplot(cars_db_plot, aes(x = price, y = horsepower, colour = cluster)),
  geom_point(alpha = 0.8, size = 2)
)
```

### Final questions

1. Which clustering method gave the most interpretable segmentation of
   the Qatar Cars data?

2. How important was scaling for this problem?

3. If you had to explain your final clustering to a non-technical
   audience, which output would you show: a dendrogram, cluster means,
   medoids, or silhouette widths?
