---
title: "Workshop 10: Clustering"
output: html_document
---

In this workshop we follow the lecture workflow for clustering on two datasets.
For each task, the 2020 NZ Election results are shown as a fully worked example.
Your job is to adapt the same code to the D&D Monsters data. The aim is not to write the shortest possible R code. The aim is to make each analysis step visible.

By the end of the workshop you should be able to:

- build a distance matrix and explain why scaling matters
- use PCA and UMAP to visualise high-dimensional data in two dimensions
- compare hierarchical clustering, K-means, PAM, and DBSCAN
- use silhouette widths and elbow plots to defend a choice of cluster number
- describe clusters using their means and representative examples

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(cluster)
library(uwot)
library(dbscan)
library(plotly)
library(factoextra)
library(scales)
library(DT)

theme_set(theme_minimal())
```

# The two datasets

## 2020 New Zealand General Election

In the 2020 general election each voter casts two votes: a party vote and a candidate vote. We cluster the 72 electorates by party vote shares to see how regions group together politically.

```{r import-election-data}
party_vars <- c(
  "Act",
  "Green",
  "Labour",
  "Maori",
  "National",
  "Other"
)

party_data <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nzelect/2020_party.csv") |>
  select(Electorate, all_of(party_vars))

party_scaled <- party_data |>
  select(all_of(party_vars)) |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

party_means <- party_data |>
  summarise(across(all_of(party_vars), mean))

party_means
```

## D&D Monsters

Each row is a monster with a type label plus combat features such as strength, dexterity, HP, challenge rating, and speed. We cluster on the numeric features and use type afterward to interpret the clusters.

```{r load-monsters}
monsters <- read_csv("http://knowlton.co.nz/161324/data/monsters.csv.gz")

monster_vars <- c(
  "str",
  "dex",
  "con",
  "int",
  "wis",
  "cha",
  "hp_number",
  "cr",
  "speed_base_number"
)

monster_data <- monsters |>
  select(name, type, all_of(monster_vars)) |>
  drop_na()

monster_data$name <- make.unique(monster_data$name)

monster_scaled <- monster_data |>
  select(all_of(monster_vars)) |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))

monster_data
monster_scaled
```

# 1. Distances and scaling

## Voting worked example

Party vote shares are percentages on a common 0 to 100 scale. Parties with larger vote shares, such as Labour, naturally have larger variance. We compare raw Euclidean distance with z-scored distance for the first six electorates.

```{r voting-distance}
# Need to use the original unscaled data for comparison.
party_data |>
  slice_head(n = 6) |>
  column_to_rownames("Electorate") |>
  dist() |>
  round(2)

party_data |>
  select(Electorate) |>
  bind_cols(party_scaled) |>
  slice_head(n = 6) |>
  column_to_rownames("Electorate") |>
  dist() |>
  round(2)

```

## D&D your turn

Compute the raw and z-scored Euclidean distance matrices for the first six monsters. Which variables appear to dominate the unscaled distances?

```{r monster-distance}
monster_data |>
  slice_head(n = 6) |>
  select(name, all_of(monster_vars)) |>
  column_to_rownames("name") |>
  dist() |>
  round(2)

monster_data |>
  select(name) |>
  bind_cols(monster_scaled) |>
  slice_head(n = 6) |>
  column_to_rownames("name") |>
  dist() |>
  round(2)

```

The raw distances are mostly driven by `hp_number`, `cr`, and `speed_base_number`, because those variables use larger numerical scales than the six ability scores. After z-scoring, every variable contributes on a comparable scale.

# 2. Dimension reduction: PCA and UMAP

## Voting worked example

PCA gives a linear two-dimensional view. UMAP gives a nonlinear two-dimensional view. These plots help us inspect the data structure before we cluster.

```{r voting-pca-umap}
# PCA
party_pca <- prcomp(
  party_scaled,
  center = TRUE,
  scale. = TRUE
)

party_pca_points <- party_pca$x |>
  as_tibble() |>
  transmute(
    pc1 = PC1,
    pc2 = PC2,
    Electorate = party_data$Electorate
  )

party_pca_plot <- party_pca_points |>
  ggplot(aes(x = pc1, y = pc2, text = Electorate)) +
  geom_point() +
  labs(
    x = "PC1",
    y = "PC2",
    title = "Voting: PCA"
  )

ggplotly(party_pca_plot, tooltip = "text")


# UMAP
set.seed(584752)

party_umap_points <- uwot::umap(
  party_scaled,
  n_neighbors = 15,
  min_dist = 0.1,
  n_components = 2,
  metric = "euclidean"
) |>
  as_tibble(.name_repair = ~ c("umap1", "umap2")) |>
  mutate(Electorate = party_data$Electorate)

party_umap_plot <- party_umap_points |>
  ggplot(aes(x = umap1, y = umap2, text = Electorate)) +
  geom_point() +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    title = "Voting: UMAP"
  )

ggplotly(party_umap_plot, tooltip = "text")
```

Try changing the seed and re-running the UMAP code. UMAP can move points around between runs, especially when the seed changes, but the broad structure should be similar.

## D&D your turn

Fit PCA and UMAP on `monster_scaled`. Use `set.seed(161324)` for UMAP. Use `name` as the point identifier.

```{r monster-pca-umap}
# PCA
monster_pca <- prcomp(
  monster_scaled,
  center = TRUE,
  scale. = TRUE
)

monster_pca_points <- monster_pca$x |>
  as_tibble() |>
  transmute(
    pc1 = PC1,
    pc2 = PC2,
    name = monster_data$name
  )

monster_pca_plot <- monster_pca_points |>
  ggplot(aes(x = pc1, y = pc2, text = name, key = name)) +
  geom_point() +
  labs(
    x = "PC1",
    y = "PC2",
    title = "Monsters: PCA"
  )

ggplotly(monster_pca_plot, tooltip = "text")


# UMAP
set.seed(161324)

monster_umap_points <- uwot::umap(
  monster_scaled,
  n_neighbors = 15,
  min_dist = 0.1,
  n_components = 2,
  metric = "euclidean"
) |>
  as_tibble(.name_repair = ~ c("umap1", "umap2")) |>
  mutate(name = monster_data$name)

monster_umap_plot <- monster_umap_points |>
  ggplot(aes(x = umap1, y = umap2, text = name, key = name)) +
  geom_point() +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    title = "Monsters: UMAP"
  )

ggplotly(monster_umap_plot, tooltip = "text")
```

The two projections usually show a broad central mass of monsters plus a smaller set of high-power cases that separate more clearly.

# 3. Hierarchical clustering

## Voting worked example

Average-linkage hierarchical clustering on scaled party data.

```{r voting-hclust}
party_dist <- dist(party_scaled, method = "euclidean")
party_hc <- hclust(party_dist, method = "average")

plot(
  party_hc,
  labels = FALSE,
  hang = -1,
  main = "Average-linkage dendrogram for NZ 2020 party vote"
)
```

Try single, complete, and average linkage.

```{r voting-linkage}
party_hc_single <- hclust(party_dist, method = "single")
party_hc_complete <- hclust(party_dist, method = "complete")
party_hc_average <- hclust(party_dist, method = "average")

plot(party_hc_single, labels = FALSE, hang = -1, main = "Single linkage")
plot(party_hc_complete, labels = FALSE, hang = -1, main = "Complete linkage")
plot(party_hc_average, labels = FALSE, hang = -1, main = "Average linkage")
```

Use the silhouette measure to choose K, then cut the tree.

```{r voting-hclust-silhouette}
fviz_nbclust(
  party_scaled,
  FUNcluster = factoextra::hcut,
  hc_method = "average",
  k.max = 10,
  method = "silhouette"
)

party_groups <- cutree(party_hc, k = 2)
```

Cluster means.

```{r voting-hclust-means}
party_clustered <- party_data
party_clustered$Cluster <- factor(party_groups)

party_cluster_means <- party_clustered |>
  group_by(Cluster) |>
  summarise(across(all_of(party_vars), mean), n_electorates = n()) |>
  arrange(desc(n_electorates))

party_cluster_means
```

Which electorates are in each cluster?

```{r voting-hclust-members}
party_clustered |>
  select(Electorate, Cluster) |>
  arrange(Cluster, Electorate) |>
  datatable()
```

## D&D your turn

Fit hierarchical clustering on `monster_scaled`. Try single, complete, and average linkage. Which gives the most balanced cluster sizes?

```{r monster-hclust}
monster_dist <- dist(monster_scaled, method = "euclidean")

monster_hc_single <- hclust(monster_dist, method = "single")
monster_hc_complete <- hclust(monster_dist, method = "complete")
monster_hc_average <- hclust(monster_dist, method = "average")

plot(monster_hc_single, labels = FALSE, hang = -1, main = "Monsters: single linkage")
plot(monster_hc_complete, labels = FALSE, hang = -1, main = "Monsters: complete linkage")
plot(monster_hc_average, labels = FALSE, hang = -1, main = "Monsters: average linkage")

single_groups <- cutree(monster_hc_single, k = 2)
complete_groups <- cutree(monster_hc_complete, k = 2)
average_groups <- cutree(monster_hc_average, k = 2)

table(single_groups)
table(complete_groups)
table(average_groups)
```

Complete linkage is usually the least chain-like option here. It tends to give a cleaner working partition for this dataset.

```{r monster-hclust-cut}
fviz_nbclust(
  monster_scaled,
  FUNcluster = factoextra::hcut,
  hc_method = "complete",
  k.max = 10,
  method = "silhouette"
)

monster_hc_groups <- cutree(monster_hc_complete, k = 2)

monster_clustered <- monster_data
monster_clustered$Cluster <- factor(monster_hc_groups)

monster_hc_means <- monster_clustered |>
  group_by(Cluster) |>
  summarise(across(all_of(monster_vars), mean), n = n()) |>
  arrange(desc(n))

monster_hc_means

monster_clustered |>
  count(Cluster, type, sort = TRUE) |>
  head(25)
```

The two-cluster solution is a broad split between ordinary and high-power monsters. It is interpretable, but it is also coarse.

# 4. K-means and the elbow heuristic

## Voting worked example

We fit K-means for K values from 1 to 10 and record the total within-cluster sum of squares. This is the elbow diagnostic.

```{r voting-kmeans-elbow}
set.seed(67)

party_elbow <- tibble(
  k = 1:10,
  tot_withinss = NA_real_
)

for (i in 1:nrow(party_elbow)) {
  k_value <- party_elbow$k[i]
  km_fit <- kmeans(party_scaled, centers = k_value, nstart = 20)
  party_elbow$tot_withinss[i] <- km_fit$tot.withinss
}

party_elbow

party_elbow_plot <- ggplot(party_elbow) +
  aes(x = k, y = tot_withinss, group = 1, text = paste0("K = ", k, "<br>Total WSS = ", round(tot_withinss))) +
  geom_line() +
  geom_point() +
  labs(x = "Number of clusters", y = "Total within-cluster sum of squares")

ggplotly(party_elbow_plot, tooltip = "text")
```

As an additional check, we can also use `fviz_nbclust()` to compare average silhouette width across K values for K-means.

```{r voting-kmeans-silhouette}
set.seed(67)

fviz_nbclust(
  party_scaled,
  FUNcluster = kmeans,
  method = "silhouette",
  k.max = 10,
  nstart = 20
)
```

Fit K-means with K = 6.

```{r voting-kmeans-final}
set.seed(67)
party_km6 <- kmeans(party_scaled, centers = 6, nstart = 25)

party_km_clustered <- party_data |>
  mutate(Cluster = factor(party_km6$cluster))

party_km_means <- party_km_clustered |>
  group_by(Cluster) |>
  summarise(across(all_of(party_vars), mean), n = n()) |>
  arrange(desc(n))

party_km_means
```

Plot the K-means clusters on the PCA and UMAP projections.

```{r voting-kmeans-plots}
party_km_pca_points <- party_pca_points |>
  mutate(Cluster = factor(party_km6$cluster))

party_km_pca_plot <- ggplot(party_km_pca_points) +
  aes(x = pc1, y = pc2, colour = Cluster, text = Electorate, key = Electorate) +
  geom_point(alpha = 0.8, size = 2) +
  labs(x = "PC1", y = "PC2", title = "Voting: K-means clusters on PCA")

ggplotly(party_km_pca_plot, tooltip = "text")

party_km_umap_points <- party_umap_points |>
  mutate(Cluster = factor(party_km6$cluster))

party_km_umap_plot <- ggplot(party_km_umap_points) +
  aes(x = umap1, y = umap2, colour = Cluster, text = Electorate, key = Electorate) +
  geom_point(alpha = 0.8, size = 2) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Voting: K-means clusters on UMAP")

ggplotly(party_km_umap_plot, tooltip = "text")
```

## D&D your turn

Fit K-means for several values of K, produce an elbow plot, then fit a final model with your chosen K.

```{r monster-kmeans-elbow}
set.seed(161324)

monster_elbow <- tibble(
  k = 1:15,
  tot_withinss = NA_real_
)

for (i in 1:nrow(monster_elbow)) {
  k_value <- monster_elbow$k[i]
  km_fit <- kmeans(monster_scaled, centers = k_value, nstart = 20)
  monster_elbow$tot_withinss[i] <- km_fit$tot.withinss
}

monster_elbow

monster_elbow_plot <- ggplot(monster_elbow) +
  aes(x = k, y = tot_withinss, group = 1, text = paste0("K = ", k, "<br>Total WSS = ", round(tot_withinss))) +
  geom_line() +
  geom_point() +
  labs(x = "Number of clusters", y = "Total within-cluster sum of squares")

ggplotly(monster_elbow_plot, tooltip = "text")
```

The elbow is gradual. K = 10 is a reasonable compromise between detail and interpretability.

Here is the same K-means choice shown with a silhouette plot.

```{r monster-kmeans-silhouette}
set.seed(161324)

fviz_nbclust(
  monster_scaled,
  FUNcluster = kmeans,
  method = "silhouette",
  k.max = 12,
  nstart = 20
)
```

```{r monster-kmeans-final}
set.seed(161324)
monster_km10 <- kmeans(monster_scaled, centers = 10, nstart = 25)

monster_km_clustered <- monster_data |>
  mutate(Cluster = factor(monster_km10$cluster))

monster_km_means <- monster_km_clustered |>
  group_by(Cluster) |>
  summarise(across(all_of(monster_vars), mean), n = n()) |>
  arrange(desc(n))

monster_km_means

monster_km_pca_points <- monster_pca_points |>
  mutate(Cluster = factor(monster_km10$cluster))

monster_km_pca_plot <- ggplot(monster_km_pca_points) +
  aes(x = pc1, y = pc2, colour = Cluster, text = name, key = name) +
  geom_point(alpha = 0.8, size = 2) +
  labs(x = "PC1", y = "PC2", title = "Monsters: K-means clusters on PCA")

ggplotly(monster_km_pca_plot, tooltip = "text")
```

# 5. PAM and silhouette widths

PAM is similar to K-means, but it chooses real observations as cluster centres. These representative observations are called medoids.

## Voting worked example

```{r voting-pam-silhouette}
party_pam_sil <- tibble(
  k = 2:8,
  avg_sil_width = NA_real_
)

for (i in 1:nrow(party_pam_sil)) {
  k_value <- party_pam_sil$k[i]
  pam_fit <- pam(party_scaled, k = k_value)
  party_pam_sil$avg_sil_width[i] <- pam_fit$silinfo$avg.width
}

party_pam_sil

party_pam_sil_plot <- ggplot(party_pam_sil) +
  aes(x = k, y = avg_sil_width) +
  geom_line() +
  geom_point() +
  labs(x = "Number of clusters", y = "Average silhouette width")

party_pam_sil_plot
```

`fviz_nbclust()` gives the same silhouette idea in one line, so it is a handy quick check for PAM too.

```{r voting-pam-fviz}
fviz_nbclust(
  party_scaled,
  FUNcluster = cluster::pam,
  method = "silhouette",
  k.max = 8
)
```

Fit the best PAM model from this range.

```{r voting-pam-final}
best_party_row <- which.max(party_pam_sil$avg_sil_width)
best_party_k <- party_pam_sil$k[best_party_row]

party_pam_best <- pam(party_scaled, k = best_party_k)

party_pam_clustered <- party_data
party_pam_clustered$Cluster <- factor(party_pam_best$clustering)

party_pam_clustered |>
  group_by(Cluster) |>
  summarise(across(all_of(party_vars), mean), n = n()) |>
  arrange(desc(n))

party_medoid_rows <- party_pam_best$id.med
party_data |>
  slice(party_medoid_rows)
```

## D&D your turn

Use PAM to find representative monsters. Compare K values from 2 to 12 using average silhouette width.

```{r monster-pam-silhouette}
monster_pam_sil <- tibble(
  k = 2:12,
  avg_sil_width = NA_real_
)

for (i in 1:nrow(monster_pam_sil)) {
  k_value <- monster_pam_sil$k[i]
  pam_fit <- pam(monster_scaled, k = k_value)
  monster_pam_sil$avg_sil_width[i] <- pam_fit$silinfo$avg.width
}

monster_pam_sil

monster_pam_sil_plot <- ggplot(monster_pam_sil) +
  aes(x = k, y = avg_sil_width) +
  geom_line() +
  geom_point() +
  labs(x = "Number of clusters", y = "Average silhouette width")

monster_pam_sil_plot
```

The same shortcut works for the monster data.

```{r monster-pam-fviz}
fviz_nbclust(
  monster_scaled,
  FUNcluster = cluster::pam,
  method = "silhouette",
  k.max = 12
)
```

Fit the best PAM model and inspect its medoids.

```{r monster-pam-final}
best_monster_row <- which.max(monster_pam_sil$avg_sil_width)
best_monster_k <- monster_pam_sil$k[best_monster_row]

monster_pam_best <- pam(monster_scaled, k = best_monster_k)

monster_pam_clustered <- monster_data
monster_pam_clustered$Cluster <- factor(monster_pam_best$clustering)

monster_pam_clustered |>
  group_by(Cluster) |>
  summarise(across(all_of(monster_vars), mean), n = n()) |>
  arrange(desc(n))

monster_medoid_rows <- monster_pam_best$id.med
monster_data |>
  slice(monster_medoid_rows)
```

Plot PAM medoids on the PCA and UMAP projections.

```{r monster-pam-medoid-plots}
monster_medoid_names <- monster_data$name[monster_medoid_rows]

monster_pca_medoid_points <- monster_pca_points |>
  mutate(
    is_medoid = name %in% monster_medoid_names,
    Medoid = if_else(is_medoid, "Medoid", "Other"),
    Medoid = factor(Medoid, levels = c("Other", "Medoid"))
  ) |>
  arrange(is_medoid)

monster_umap_medoid_points <- monster_umap_points |>
  mutate(
    is_medoid = name %in% monster_medoid_names,
    Medoid = if_else(is_medoid, "Medoid", "Other"),
    Medoid = factor(Medoid, levels = c("Other", "Medoid"))
  ) |>
  arrange(is_medoid)

monster_medoid_pca_plot <- ggplot(monster_pca_medoid_points) +
  aes(
    x = pc1,
    y = pc2,
    colour = Medoid,
    text = name,
    key = name
  ) +
  geom_point(alpha = 0.8, size = 2) +
  labs(
    x = "PC1",
    y = "PC2",
    title = "PCA: PAM medoids highlighted"
  )

monster_medoid_umap_plot <- ggplot(monster_umap_medoid_points) +
  aes(
    x = umap1,
    y = umap2,
    colour = Medoid,
    text = name,
    key = name
  ) +
  geom_point(alpha = 0.8, size = 2) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    title = "UMAP: PAM medoids highlighted"
  )

ggplotly(monster_medoid_pca_plot, tooltip = "text")
ggplotly(monster_medoid_umap_plot, tooltip = "text")
```

# 6. Cluster profiles

Cluster means are useful, but variables like HP and CR use different units. For profile plots, we rescale each variable to the range 0 to 1 before averaging within clusters.

## Voting worked example

```{r voting-profiles}
party_profile_long <- party_clustered |>
  select(Cluster, all_of(party_vars)) |>
  pivot_longer(
    cols = all_of(party_vars),
    names_to = "Party",
    values_to = "VoteShare"
  ) |>
  group_by(Cluster, Party) |>
  summarise(
    MeanVoteShare = mean(VoteShare, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  mutate(
    Cluster = paste0("Cluster ", Cluster, " (n = ", n, ")")
  )

party_profile_plot <- ggplot(party_profile_long) +
  aes(
    x = Cluster,
    y = MeanVoteShare,
    fill = Party,
    text = paste0(
      Cluster,
      "<br>Party: ", Party,
      "<br>Mean vote share: ", round(MeanVoteShare, 1)
    )
  ) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Mean vote share",
    fill = "Party",
    title = "Voting profile by cluster"
  )

ggplotly(party_profile_plot, tooltip = "text")
```

## D&D your turn

Use the hierarchical cluster labels from Section 3. Rescale each monster variable, average within cluster, then produce a faceted bar chart.

```{r monster-profiles}
monster_profile_long <- monster_clustered |>
  select(Cluster, all_of(monster_vars)) |>
  pivot_longer(
    cols = all_of(monster_vars),
    names_to = "Variable",
    values_to = "Value"
  ) |>
  group_by(Variable) |>
  mutate(
    ScaledValue = scales::rescale(Value)
  ) |>
  ungroup() |>
  group_by(Cluster, Variable) |>
  summarise(
    MeanScaledValue = mean(ScaledValue, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  mutate(
    Cluster = paste0("Cluster ", Cluster, " (n = ", n, ")")
  )

monster_profile_plot <- ggplot(monster_profile_long) +
  aes(
    x = Cluster,
    y = MeanScaledValue,
    fill = Variable,
    text = paste0(
      Cluster,
      "<br>Variable: ", Variable,
      "<br>Mean scaled value: ", round(MeanScaledValue, 2)
    )
  ) +
  geom_col(position = "fill") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Mean value rescaled within variable",
    fill = "Monster variable",
    title = "Monster profile by cluster"
  )

ggplotly(monster_profile_plot, tooltip = "text")
```

The profile bars make the split easy to explain: one cluster contains the higher-HP, higher-CR monsters, while the other covers the broader mid-range.

# 7. DBSCAN: density-based clustering

DBSCAN groups dense points and labels isolated points as noise. It has two parameters: `eps`, the neighbourhood radius, and `minPts`, the minimum number of points needed to define a dense region. Unlike K-means and PAM, DBSCAN does not require K in advance.

## Voting worked example

Choose `eps` with a k-nearest-neighbour distance plot.

```{r voting-dbscan-knn}
dbscan::kNNdistplot(as.matrix(party_scaled), k = 3)
abline(h = 1.7, lty = 2, col = 2)
```

Fit DBSCAN.

```{r voting-dbscan-fit}
party_db <- dbscan::dbscan(as.matrix(party_scaled), eps = 1.7, minPts = 3)
table(party_db$cluster)
```

Colour the PCA projection by DBSCAN cluster.

```{r voting-dbscan-plot}
party_db_points <- party_pca_points
party_db_points$Cluster <- factor(party_db$cluster)

party_db_plot <- ggplot(party_db_points) +
  aes(x = pc1, y = pc2, colour = Cluster, text = paste0(Electorate, "<br>Cluster: ", Cluster)) +
  geom_point(alpha = 0.8, size = 2) +
  labs(x = "PC1", y = "PC2", title = "Voting: DBSCAN clusters on PCA")

ggplotly(party_db_plot, tooltip = "text")
```

Try several DBSCAN settings. This code creates two matricies as tibbles and run DBSCAN across them.

```{r voting-dbscan-parameter-scan}
party_db_results <- tibble(
  eps = c(1.5, 1.0, 2.0, 1.8, 2.0),
  minPts = c(5, 3, 4, 3, 3)
) |>
  rowwise() |>
  mutate(
    dbscan_result = list(
      dbscan::dbscan(
        as.matrix(party_scaled),
        eps = eps,
        minPts = minPts
      )
    ),
    n_clusters = length(setdiff(unique(dbscan_result$cluster), 0)),
    n_noise = sum(dbscan_result$cluster == 0),
    n_total = length(dbscan_result$cluster)
  ) |>
  ungroup() |>
  select(eps, minPts, n_clusters, n_noise, n_total)

party_db_results
```

## D&D your turn

Apply DBSCAN to the monster data. Use `kNNdistplot` to choose `eps`, fit the model, tabulate clusters against monster type, and experiment with different parameter values.

```{r monster-dbscan-knn}
dbscan::kNNdistplot(as.matrix(monster_scaled), k = 4)
abline(h = 1.8, lty = 2, col = 2)
```

```{r monster-dbscan-fit}
monster_db <- dbscan::dbscan(as.matrix(monster_scaled), eps = 1.8, minPts = 4)
table(monster_db$cluster)
```

With these settings, most monsters fall into one dense cluster and a smaller set are treated as noise.

```{r monster-dbscan-results}
monster_db_results <- monster_data
monster_db_results$Cluster <- factor(monster_db$cluster)
monster_db_results$Status <- ifelse(monster_db_results$Cluster == "0", "noise", "clustered")

monster_db_results |>
  count(Cluster, type, sort = TRUE)

monster_db_results |>
  count(Status)
```

Try several DBSCAN settings.

```{r monster-dbscan-parameter-scan}
monster_db_parameter_results <- tibble(
  eps = c(1.2, 1.3, 1.4, 1.25, 1.25),
  minPts = c(3, 5, 5, 4, 5)
) |>
  rowwise() |>
  mutate(
    db = list(
      dbscan::dbscan(
        as.matrix(monster_scaled),
        eps = eps,
        minPts = minPts
      )
    ),
    n_clusters = length(setdiff(unique(db$cluster), 0)),
    n_noise = sum(db$cluster == 0),
    n_total = length(db$cluster)
  ) |>
  ungroup() |>
  select(eps, minPts, n_clusters, n_noise, n_total)

monster_db_parameter_results
```

# 8. Wrap-up

## Final questions

1. Which clustering method gave the most interpretable segmentation of the election data? Of the D&D monsters?

2. How important was scaling for the monsters data? Would your conclusions change if you used the raw values instead of z-scores?

3. How do party-vote clusters differ from the political narrative you expected? Which electorates surprised you?

4. For the monsters, did DBSCAN find a different number of clusters than your chosen K from the other methods? What does that tell you about the data structure?

5. If you had to explain your final clustering of D&D monsters to someone who has never played D&D, which output would you show: a dendrogram, cluster means, medoids, silhouette widths, the profile bars, or the DBSCAN plot?

## Guide notes

For the election data, the cluster summaries and profiles are the easiest way to explain the groupings. For the monsters, the K-means summaries and the profile bars give a quick overview, while PAM gives concrete representative monsters.

Scaling matters a lot for the monsters. Without z-scores, `hp_number`, `cr`, and `speed_base_number` dominate the distances and the ability scores matter much less.

The election data still has a broad mainstream block, but some electorates stand out because of distinctive Green or Maori Party support.

DBSCAN mostly finds one dense core plus noise points instead of reproducing the same K as the partitioning methods. That suggests the monster data has a strong central mass with a tail of more unusual cases.
