---
title: "Workshop 12 Guide: Association Rule Mining"
output: html_document
---

This guide follows the same structure as `lab12.Rmd` and fills in one
complete solution path. The exact rules you focus on can vary, but the
workflow below is the intended one.

Use the `tidytuesdayR` package to pull the recipe data directly from
TidyTuesday. If RStudio offers to install `tidytuesdayR`, allow it
before running the setup chunk.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(arules)

theme_set(theme_minimal())

load_tidytuesday_dataset <- function(tt_date, dataset_name, parquet_url) {
  if (requireNamespace("tidytuesdayR", quietly = TRUE)) {
    tt_data <- tryCatch(
      suppressMessages(tidytuesdayR::tt_load(tt_date)),
      error = function(e) NULL
    )

    if (!is.null(tt_data) &&
        dataset_name %in% names(tt_data) &&
        !is.null(tt_data[[dataset_name]])) {
      return(tt_data[[dataset_name]] |> tibble::as_tibble())
    }
  }

  if (!requireNamespace("arrow", quietly = TRUE)) {
    stop(
      "Install the 'tidytuesdayR' package (RStudio will usually prompt you), ",
      "or install 'arrow' to use the backup data file.",
      call. = FALSE
    )
  }

  parquet_file <- tempfile(fileext = ".parquet")
  on.exit(unlink(parquet_file), add = TRUE)
  download_status <- tryCatch(
    utils::download.file(parquet_url, parquet_file, mode = "wb", quiet = TRUE),
    error = function(e) 1L
  )

  if (!identical(download_status, 0L) || !file.exists(parquet_file)) {
    stop(
      "Could not load the dataset from TidyTuesday or the course backup file.",
      call. = FALSE
    )
  }

  arrow::read_parquet(parquet_file) |> tibble::as_tibble()
}

compose_ggplot <- function(plot, ...) {
  layers <- list(...)

  for (i in seq_along(layers)) {
    plot <- ggplot2:::add_ggplot(plot, layers[[i]], paste0("layer_", i))
  }

  plot
}
```

## 1. Load and clean the recipe data

```{r}
cuisines_raw <- load_tidytuesday_dataset(
  tt_date = "2025-09-16",
  dataset_name = "cuisines",
  parquet_url = "https://knowlton.co.nz/161324/data/cuisines.parquet"
)

cuisines_raw |> glimpse()
```

```{r}
not_ingredients <- c(
  "chopped", "minced", "diced", "sliced", "grated", "shredded", "crushed",
  "julienned", "cubed", "halved", "quartered", "peeled", "seeded", "stemmed",
  "trimmed", "cleaned", "rinsed", "drained", "thawed", "softened", "melted",
  "beaten", "divided", "separated", "sifted", "crumbled", "toasted",
  "mashed", "pureed", "blanched", "boiled", "roasted", "juiced",
  "thinly sliced", "finely chopped", "coarsely chopped",
  "finely", "coarsely", "thinly",
  "to taste", "as needed", "or to taste", "or as needed",
  "or more", "or more to taste", "or more as needed",
  "salt to taste", "salt and pepper to taste",
  "salt and ground black pepper to taste",
  "salt and ground black pepper",
  "salt and pepper", "if desired", "for garnish",
  "salt", "pepper", "kosher salt", "ground black pepper",
  "freshly ground black pepper",
  "skinless", "cooking spray", "room temperature", "needed", "taste"
)

unit_rx <- regex(
  paste0(
    "^(cups?|tablespoons?|tbsp|teaspoons?|tsp|pounds?|lbs?|ounces?|oz|",
    "cans?|packages?|pkgs?|envelopes?|pinch(es)?|cloves?|slices?|",
    "stalks?|heads?|bunche?s?|pieces?|strips?|sprigs?)\\s+(of\\s+)?"
  ),
  ignore_case = TRUE
)

recipe_items <- cuisines_raw |>
  select(name, country, ingredients) |>
  mutate(recipe_id = row_number()) |>
  separate_longer_delim(ingredients, delim = ", ") |>
  mutate(
    ingredient = ingredients |>
      str_to_lower() |>
      str_trim() |>
      str_remove_all("\\(.*?\\)") |>
      str_remove("^[0-9\\u00bd\\u00bc\\u00be\\u2153\\u2154\\u215b\\u215c\\u215d\\u215e/\\. ]+") |>
      str_remove(unit_rx) |>
      str_remove(regex("^(large|medium|small|extra-?large)\\s+", ignore_case = TRUE)) |>
      str_remove(regex("\\s+(to taste|as needed|if desired|or more to taste|or more)$",
                       ignore_case = TRUE)) |>
      str_remove(regex(",\\s*(divided|optional|drained|thawed|softened|melted|beaten)$",
                       ignore_case = TRUE)) |>
      str_trim() |>
      str_replace_all("^unsalted butter$", "butter") |>
      str_replace_all("^garlic powder$", "garlic") |>
      str_replace_all("^minced garlic$", "garlic") |>
      str_replace_all("^eggs$", "egg") |>
      str_replace_all("^onions$", "onion") |>
      str_replace_all("^potatoes$", "potato") |>
      str_replace_all("^tomatoes$", "tomato") |>
      str_replace_all("^carrots$", "carrot") |>
      str_replace_all("^lemons$", "lemon") |>
      str_replace_all("^limes$", "lime") |>
      str_replace_all("^mushrooms$", "mushroom") |>
      str_replace_all("^bananas$", "banana") |>
      str_replace_all("^green onions$", "green onion") |>
      str_replace_all("^cloves$", "clove")
  ) |>
  filter(
    nchar(ingredient) > 2,
    !ingredient %in% not_ingredients
  )

recipe_items |>
  count(ingredient, sort = TRUE) |>
  slice_head(n = 20)
```

Very common ingredients deserve caution later, because they can produce
high-confidence rules with lift near one.

## 2. Create the transaction object

```{r}
top_ingredients <- recipe_items |>
  count(ingredient, sort = TRUE) |>
  slice_head(n = 75) |>
  pull(ingredient)

recipe_items_filtered <- recipe_items |>
  filter(ingredient %in% top_ingredients)

recipe_trans <- recipe_items_filtered |>
  select(recipe_id, ingredient) |>
  distinct() |>
  mutate(present = TRUE) |>
  pivot_wider(
    names_from = ingredient,
    values_from = present,
    values_fill = FALSE
  ) |>
  select(-recipe_id) |>
  as.matrix() |>
  as("transactions")

recipe_trans
```

```{r}
recipe_trans |>
  head(n = 10) |>
  inspect()
```

`inspect()` is usually the better first tool for transactions because it
keeps the basket structure visible.

## 3. Item frequencies

```{r}
itemFrequencyPlot(recipe_trans, topN = 25, type = "relative")
```

```{r}
recipe_freq <- recipe_trans |>
  itemFrequency() |>
  enframe(name = "ingredient", value = "support") |>
  arrange(desc(support)) |>
  slice_head(n = 25)

compose_ggplot(
  ggplot(recipe_freq, aes(x = support, y = fct_reorder(ingredient, support))),
  geom_col()
)
```

The top-frequency items are exactly the ones that can make confidence
look large even when the rule tells us little that is new.

## 4. Mine rules with Apriori

```{r}
recipe_rules <- apriori(
  recipe_trans,
  parameter = list(support = 0.05, confidence = 0.40, minlen = 2, maxlen = 4),
  control = list(verbose = FALSE)
)

recipe_rules
summary(recipe_rules)
```

```{r}
recipe_rules |>
  sort(by = "support") |>
  head(n = 10) |>
  inspect()

recipe_rules |>
  sort(by = "confidence") |>
  head(n = 10) |>
  inspect()

recipe_rules |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```

For exploratory work, lift is usually the more informative ranking
because it discounts consequents that are common everywhere.

## 5. High confidence versus high lift

```{r}
recipe_rules |>
  subset(confidence >= 0.75 & lift <= 1.05) |>
  sort(by = "confidence") |>
  head(n = 10) |>
  inspect()
```

These are the classic misleading rules: the consequent is so common that
high confidence does not necessarily indicate a strong association.

## 6. Filter rules

```{r}
garlic_rules <- recipe_rules |>
  subset(lhs %in% "garlic" & lift > 1.1)

garlic_rules |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```

```{r}
recipe_rules |>
  subset(lhs %in% "olive oil") |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```

```{r}
recipe_rules |>
  subset(lhs %ain% c("garlic", "onion")) |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```

This is where `subset()` becomes useful: once the full rule set has been
generated, the main challenge is reducing it to the part that matches
your substantive question.

## 7. Optional interactive view

```{r, eval=FALSE}
library(arulesViz)

recipe_rules |>
  sort(by = "lift") |>
  head(n = 100) |>
  inspectDT()
```

## 8. Bank-service extension

```{r}
bank <- read_csv("../data/bank-rules.csv", show_col_types = FALSE) |>
  transactions(format = "long", cols = c("ACCOUNT", "SERVICE"))

bank_rules <- apriori(
  bank,
  parameter = list(support = 0.005, confidence = 0.4),
  control = list(verbose = FALSE)
)
```

```{r}
bank_rules |>
  subset(rhs %in% "CCRD") |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```

```{r}
bank_rules |>
  subset((lhs %in% "MTG" | rhs %in% "MTG") &
         (lhs %in% "HMEQLC" | rhs %in% "HMEQLC")) |>
  sort(by = "lift") |>
  head(n = 10) |>
  inspect()
```

Compared with the recipe data, the bank rules are often easier to read
as action items because the services already have a direct cross-selling
interpretation.

## Wrap-up

One defensible summary is:

- support tells us how common a pattern is
- confidence tells us how often the rhs appears once the lhs is present
- lift is usually the best first check for whether a rule is genuinely
  more informative than baseline item popularity
- `subset()` is essential once the rule set gets large
