---
title: "Workshop 10 Guide: Ensemble Learning with tidymodels"
output: html_document
---

This guide follows the same workflow as `lab10.Rmd`, but fills in one
complete solution path. The exact tuning results may differ slightly,
especially for the `xgboost` grid, but the sequence below is the
intended one.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(baguette)
library(stacks)
library(bonsai)
source(file.path(dirname(knitr::current_input(dir = TRUE)), "..", "common", "tidytuesday.R"))
source(file.path(dirname(knitr::current_input(dir = TRUE)), "..", "common", "longbeach.R"))

theme_set(theme_minimal())
tidymodels_prefer()

doc_dir <- dirname(knitr::current_input(dir = TRUE))
data_dir <- resolve_data_dir(doc_dir)
longbeach_url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-04/longbeach.csv"

longbeach <- read_tidytuesday_cached(
  longbeach_url,
  backup_file = "longbeach.parquet",
  data_dir = data_dir,
  csv_file = "longbeach.csv"
)
if (requireNamespace("arrow", quietly = TRUE)) {
  write_tidytuesday_parquet(longbeach, "longbeach.parquet", data_dir)
}

adoption_data <- prepare_longbeach_adoption_data(longbeach)
```

## 1. Split the data

```{r}
set.seed(161324)
adoption_split <- initial_split(adoption_data, prop = 0.8, strata = is_adopted)
adoption_train <- training(adoption_split)
adoption_test <- testing(adoption_split)

adoption_folds <- vfold_cv(adoption_train, v = 3, strata = is_adopted)
ensemble_metrics <- metric_set(roc_auc, pr_auc, accuracy)
```

## 2. Shared recipe

```{r}
adoption_recipe <- recipe(is_adopted ~ ., data = adoption_train) |>
  step_other(primary_color, intake_subtype, threshold = 0.03) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())

adoption_recipe
```

The key idea is to keep one preprocessing path for all models. That way,
the ensemble comparison is mostly about the learner rather than hidden
differences in preprocessing.

## 3. Model specifications

```{r}
tree_spec <- decision_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")

bag_spec <- bag_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart", times = 10) |>
  set_mode("classification")

xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")

tree_wf <- workflow() |> add_recipe(adoption_recipe) |> add_model(tree_spec)
bag_wf <- workflow() |> add_recipe(adoption_recipe) |> add_model(bag_spec)
xgb_wf <- workflow() |> add_recipe(adoption_recipe) |> add_model(xgb_spec)
```

## 4. Tune each workflow

```{r}
tree_grid <- grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 3)
bag_grid <- grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 2)
xgb_grid <- grid_latin_hypercube(
  trees(),
  tree_depth(),
  learn_rate(),
  min_n(),
  loss_reduction(),
  sample_prop(),
  size = 6
)
```

```{r}
set.seed(161324)
tree_res <- tune_grid(
  tree_wf,
  resamples = adoption_folds,
  grid = tree_grid,
  metrics = ensemble_metrics,
  control = control_stack_grid()
)

set.seed(161324)
bag_res <- tune_grid(
  bag_wf,
  resamples = adoption_folds,
  grid = bag_grid,
  metrics = ensemble_metrics,
  control = control_stack_grid()
)

set.seed(161324)
xgb_res <- tune_grid(
  xgb_wf,
  resamples = adoption_folds,
  grid = xgb_grid,
  metrics = ensemble_metrics,
  control = control_stack_grid()
)
```

```{r}
bind_rows(
  show_best(tree_res, metric = "roc_auc", n = 3) |> mutate(model = "tree"),
  show_best(bag_res, metric = "roc_auc", n = 3) |> mutate(model = "bagged"),
  show_best(xgb_res, metric = "roc_auc", n = 3) |> mutate(model = "xgboost")
) |>
  select(model, mean, everything())
```

In most runs, the single tree is the weakest, the bagged tree is more
stable, and the boosted tree is usually strongest on ROC AUC.

## 5. Finalise and test

```{r}
best_tree <- select_best(tree_res, metric = "roc_auc")
best_bag <- select_best(bag_res, metric = "roc_auc")
best_xgb <- select_best(xgb_res, metric = "roc_auc")

tree_fit <- finalize_workflow(tree_wf, best_tree) |> fit(adoption_train)
bag_fit <- finalize_workflow(bag_wf, best_bag) |> fit(adoption_train)
xgb_fit <- finalize_workflow(xgb_wf, best_xgb) |> fit(adoption_train)
```

```{r}
score_model <- function(fit, data) {
  bind_cols(
    data |> select(is_adopted),
    predict(fit, data),
    predict(fit, data, type = "prob")
  )
}

tree_pred <- score_model(tree_fit, adoption_test)
bag_pred <- score_model(bag_fit, adoption_test)
xgb_pred <- score_model(xgb_fit, adoption_test)

bind_rows(
  ensemble_metrics(tree_pred, truth = is_adopted, estimate = .pred_class, .pred_Adopted) |> mutate(model = "tree"),
  ensemble_metrics(bag_pred, truth = is_adopted, estimate = .pred_class, .pred_Adopted) |> mutate(model = "bagged"),
  ensemble_metrics(xgb_pred, truth = is_adopted, estimate = .pred_class, .pred_Adopted) |> mutate(model = "xgboost")
) |>
  select(model, .metric, .estimate)
```

That table is the key comparison. A good discussion point is whether the
extra performance from `xgboost` is worth the reduction in
interpretability relative to the simpler tree-based models.

## 6. Optional stack

```{r}
model_stack <- stacks() |>
  add_candidates(tree_res) |>
  add_candidates(bag_res) |>
  add_candidates(xgb_res) |>
  blend_predictions(metric = metric_set(roc_auc)) |>
  fit_members()
```

```{r}
stack_pred <- bind_cols(
  adoption_test |> select(is_adopted),
  predict(model_stack, adoption_test),
  predict(model_stack, adoption_test, type = "prob")
)

ensemble_metrics(stack_pred, truth = is_adopted, estimate = .pred_class, .pred_Adopted)
```

The stack will not always beat the best single model by much, but it is
a useful capstone because it shows the full path from single learners to
bagging, boosting, and finally model blending.
