---
always_allow_html: true
---
# Ensemble Learning {#sec:ensemble}

```{r, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(comment = "#>")
theme_set(theme_minimal())
options(pillar.sigfig = 5)
```

The methods covered in previous chapters each build a single predictive
model: a regression equation, a tree, a neural network, and so on. An
*ensemble* takes a different approach: it builds many models and
combines their predictions. The resulting ensemble often outperforms any
of its constituent learners, sometimes dramatically so.

**How this chapter connects.** This chapter builds directly on the
supervised modelling workflow from Chapter 5 and on the single-model
methods introduced earlier in prediction and classification. In the
comparison table in Section \@ref(sec:task-compare), ensembles are not a
separate problem type so much as a strategy for improving supervised
performance by combining several learners.

We will use the **Long Beach Animal Shelter** data introduced in
Lecture 10 to anchor the examples in this chapter. The modelling task is
to predict whether a cat or dog will be *adopted*, using only information
available at the time of intake.

```{r ensemble-setup, message=FALSE, warning=FALSE}
library(tidymodels)
library(baguette)
library(mirai)
daemons(10)
tidymodels_prefer()

source("../common/longbeach.R")
longbeach <- read_csv("../data/longbeach.csv", show_col_types = FALSE)
adoption_data <- prepare_longbeach_adoption_data(longbeach)

set.seed(2024)
adopt_split  <- initial_split(adoption_data, prop = 0.75, strata = is_adopted)
adopt_train  <- training(adopt_split)
adopt_test   <- testing(adopt_split)
adopt_folds  <- vfold_cv(adopt_train, v = 5, strata = is_adopted)
```

## Motivation: Why Combine Models? {#sec:ensemble-motivation}

### The variance reduction argument

To see why aggregation helps, suppose we have $B$ learners
$\hat{f}_1, \ldots, \hat{f}_B$, each with variance $\sigma^2$ for a
given test point. If the pairwise correlation between any two learners
is $\rho$, then the variance of their average is
\begin{equation}
\operatorname{Var}\!\left(\frac{1}{B}\sum_{b=1}^B \hat{f}_b(x)\right)
  = \rho\sigma^2 + \frac{1-\rho}{B}\,\sigma^2.
(\#eq:ensemble-var)
\end{equation}
The second term, $\frac{(1-\rho)\sigma^2}{B}$, shrinks to zero as $B$
increases, showing that averaging many learners reduces the uncertainty
from any one of them. The first term, $\rho\sigma^2$, remains no matter
how large $B$ becomes. This is the irreducible variance floor imposed by
the correlation between models.

Equation \@ref(eq:ensemble-var) has two immediate practical lessons.

1.  Averaging helps most when individual learners are highly variable
    (high $\sigma^2$). Trees are a classic example of such an unstable
    learner (see Section \@ref(sec:treeinstability)).

2.  Averaging helps most when learners are weakly correlated (low
    $\rho$). Building diverse learners -- by resampling, by restricting
    features, or by using entirely different model families -- is
    therefore central to ensemble design.

### Bias vs variance in ensembles

Recall from Section \@ref(sec:predictionbasics) that prediction error
decomposes as MSE $=$ Bias$^2$ $+$ Variance $+$ Irreducible noise. The
different ensemble strategies attack this decomposition in different
ways.

*Bagging* averages many learners trained on bootstrap samples. It
reduces variance (Equation \@ref(eq:ensemble-var)) but does not reduce
bias, because each individual learner is still fit on a sample from the
same distribution.

*Boosting* builds learners sequentially, with each learner targeting the
errors of the previous stage. It primarily reduces bias by progressively
improving the fit, at the cost of some increase in variance if too many
rounds are used.

*Stacking* combines different model families. Since each family has
different inductive biases, a well-chosen meta-learner can exploit
situations where each base learner makes different errors, potentially
reducing both bias and variance.

## Bagging {#sec:bagging}

### The bootstrap aggregating algorithm

Bagging (Breiman, 1996) applies the bootstrap idea from validation
(Section \@ref(sec:validation)) to model building. The full algorithm
for regression is:

1.  For $b = 1, \ldots, B$:
    a.  Draw a bootstrap sample $\mathcal{D}^{(b)}$ of size $n$ from
        the training data, sampling rows uniformly with replacement.
    b.  Fit a learner $\hat{f}_b$ to $\mathcal{D}^{(b)}$.
2.  For a new test point $x$, predict
    \begin{equation}
    \hat{f}_{\text{bag}}(x) = \frac{1}{B}\sum_{b=1}^B \hat{f}_b(x).
    (\#eq:bagging-pred)
    \end{equation}

For classification the individual learners produce class probabilities
$\hat{p}_{bc}(x)$ for class $c$. These are averaged across the $B$
learners,
$$\hat{p}_c(x) = \frac{1}{B}\sum_{b=1}^B \hat{p}_{bc}(x),$$
and the predicted class is $\arg\max_c \hat{p}_c(x)$.

Because each bootstrap sample is drawn independently, the $B$ learners
are not perfectly correlated, giving the variance reduction in
Equation \@ref(eq:ensemble-var). Bias is unchanged because each individual
tree is trained on a dataset that has the same distribution as the
original training set.

### Out-of-bag error {#sec:oob}

A useful property of bagging is that each observation is absent from
roughly $1/e \approx 36.8\%$ of the bootstrap samples[^oob]. The
observations not included in $\mathcal{D}^{(b)}$ are called
*out-of-bag* (OOB) for tree $b$. For each training observation $i$, we
can form an OOB prediction

$$\hat{f}_{\text{OOB}}(x_i) = \frac{1}{|\mathcal{B}_i|}\sum_{b \in \mathcal{B}_i} \hat{f}_b(x_i)$$

where $\mathcal{B}_i = \{b : i \notin \mathcal{D}^{(b)}\}$ is the set
of trees for which observation $i$ was out-of-bag. The mean square
error of these OOB predictions is a nearly unbiased estimate of the
test error, without needing a separate validation set.

[^oob]: For a sample of size $n$, the probability that a given
    observation is excluded from a bootstrap sample is
    $(1 - 1/n)^n \to e^{-1} \approx 0.368$ as $n \to \infty$.

### Implementation in R

Bagged trees are available in `tidymodels` via the `bag_tree()` function
from the `baguette` package.

::: {.example}
**Bagged trees for the Long Beach adoption data**

We fit a bagged tree model with 50 trees and compare it to a single
decision tree on the same data.

```{r, message=FALSE}
adopt_recipe <- recipe(is_adopted ~ ., data = adopt_train) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors())

single_tree_spec <- decision_tree(mode = "classification") |>
  set_engine("rpart")

bag_spec <- bag_tree(mode = "classification") |>
  set_engine("rpart", times = 50)

wf <- workflow() |> add_recipe(adopt_recipe)

single_tree_res <- wf |>
  add_model(single_tree_spec) |>
  fit_resamples(adopt_folds, metrics = metric_set(accuracy, roc_auc))

bag_res <- wf |>
  add_model(bag_spec) |>
  fit_resamples(adopt_folds, metrics = metric_set(accuracy, roc_auc))

bind_rows(
  collect_metrics(single_tree_res) |> mutate(model = "single tree"),
  collect_metrics(bag_res)         |> mutate(model = "bagged trees")
) |>
  select(model, .metric, mean)
```

Points to note:

-   `step_novel()` assigns unseen factor levels (in folds) to a new
    level, and `step_unknown()` handles `NA` values in nominal
    predictors, both of which are needed when working with many folds.

-   The 5-fold cross-validation AUC provides an honest comparison
    between the two models without touching the test set.

-   The bagged version typically shows reduced variance across folds,
    consistent with Equation \@ref(eq:ensemble-var).
:::

## Random Forests {#sec:ensemblerf}

Random forests (Section \@ref(sec:randomforest)) are bagged trees with
one important modification: at each candidate split, only a random
subset of $m$ predictors is considered rather than all $p$ predictors.
Choosing $m < p$ ensures that the individual trees are less correlated
with each other, thereby pushing $\rho$ in Equation \@ref(eq:ensemble-var)
closer to zero and further improving ensemble accuracy.

A common default is $m = \lfloor \sqrt{p} \rfloor$ for classification
and $m = \lfloor p/3 \rfloor$ for regression, corresponding to the
`mtry` argument in most R implementations.

### Variable importance {#sec:varimp}

Ensembles of trees lose the legibility of a single tree, but they gain
the ability to measure the *importance* of each predictor in a
systematic way.

**Impurity-based importance** (also called Gini importance) sums the
reduction in node impurity -- measured by the Gini index
(Equation \@ref(eq:gini)) for classification, or variance reduction for
regression -- attributed to splits on variable $j$, weighted by the
proportion of observations reaching each node and averaged over all $B$
trees:
$$\hat{\mathcal{I}}_j = \frac{1}{B}\sum_{b=1}^B \sum_{t:\, j_t = j} \frac{n_t}{n}\,\Delta\mathcal{G}(t)$$
where $j_t$ is the variable used at node $t$, $n_t$ is the number of
training observations reaching node $t$, $n$ is the total training
size, and $\Delta\mathcal{G}(t)$ is the impurity reduction at node $t$.

**Permutation importance** provides a more direct measure of predictive
contribution. For each tree $b$ and each variable $j$:
1.  Compute the OOB error for tree $b$ on the unperturbed OOB data.
2.  Randomly permute the values of variable $j$ in the OOB data and
    recompute the OOB error.
3.  The importance of $j$ is the average *increase* in error caused by
    the permutation.

A large permutation importance value indicates that the variable
carries genuine predictive information; permuting it hurts performance.
A near-zero value suggests the variable is irrelevant.

::: {.example}
**Random forest and variable importance for the Long Beach adoption data**

```{r, message=FALSE}
library(ranger)
rf_spec <- rand_forest(mode = "classification", trees = 500,
                       mtry = tune()) |>
  set_engine("ranger", importance = "impurity")

rf_wf <- wf |> add_model(rf_spec)

# Tune mtry over a small grid
rf_grid <- tibble(mtry = c(2L, 4L, 6L))
rf_tune <- tune_grid(rf_wf, resamples = adopt_folds,
                     grid = rf_grid,
                     metrics = metric_set(roc_auc))
show_best(rf_tune, metric = "roc_auc")
```

```{r, message=FALSE}
# Fit best model on full training set and inspect importance
best_mtry <- select_best(rf_tune, metric = "roc_auc")
final_rf  <- finalize_workflow(rf_wf, best_mtry) |> fit(adopt_train)

final_rf |>
  extract_fit_engine() |>
  (\(m) tibble(variable = names(m$variable.importance),
               importance = m$variable.importance))() |>
  arrange(desc(importance)) |>
  slice_head(n = 10) |>
  ggplot(aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = NULL, y = "Impurity importance", title = "Random forest: variable importance")
```

The plot shows which intake-time features drive the adoption prediction.
Variables near the top account for the largest total impurity reduction
across the 500 trees.
:::

## Boosting {#sec:boosting}

### Forward stagewise additive modelling

Boosting belongs to a class of methods known as *forward stagewise
additive modelling*. The goal is to construct an additive expansion

\begin{equation}
F_M(x) = \sum_{m=1}^M \beta_m f_m(x)
(\#eq:additive-expansion)
\end{equation}

where each $f_m$ is a weak learner (typically a shallow tree) and
$\beta_m > 0$ is its contribution weight. We want to minimise a loss
function

$$\sum_{i=1}^n L\bigl(y_i,\, F_M(x_i)\bigr)$$

over the choices of $\beta_m$ and $f_m$. Minimising this jointly over
all $M$ terms simultaneously is computationally intractable. The
*forward stagewise* approach instead adds one term at a time, keeping
all previous terms fixed:

At step $m$, with $F_{m-1}$ already constructed, we find
$$(\beta_m, f_m) = \arg\min_{\beta, f} \sum_{i=1}^n L\bigl(y_i,\, F_{m-1}(x_i) + \beta f(x_i)\bigr).$$

When the loss is squared error, $L(y, \hat y) = \tfrac{1}{2}(y - \hat y)^2$, the
optimal next step is simply to fit $f_m$ to the residuals
$r_i = y_i - F_{m-1}(x_i)$. For other loss functions the residuals are
replaced by *pseudo-residuals*.

### Gradient boosting {#sec:gradboosting}

Gradient boosting (Friedman, 2001) generalises the forward stagewise
approach to any differentiable loss $L$ by interpreting each new step as
a gradient descent move in function space.

Define the pseudo-residuals at step $m$ as the negative gradient of the
loss with respect to the current predictions:
\begin{equation}
r_i^{(m)} = - \left[\frac{\partial L(y_i,\, F(x_i))}{\partial F(x_i)}\right]_{F = F_{m-1}}.
(\#eq:pseudoresid)
\end{equation}

The algorithm is then:

1.  Initialise $F_0(x) = \arg\min_c \sum_i L(y_i, c)$
    (e.g. the mean response for squared error, or the log-odds of the
    majority class for log-loss).

2.  For $m = 1, \ldots, M$:
    a.  Compute pseudo-residuals $r_i^{(m)}$ from
        Equation \@ref(eq:pseudoresid).
    b.  Fit a weak learner $h_m$ to the pairs
        $\{(x_i, r_i^{(m)})\}_{i=1}^n$.
    c.  Update
        \begin{equation}
        F_m(x) = F_{m-1}(x) + \eta\, h_m(x)
        (\#eq:boost-update)
        \end{equation}
        where $\eta \in (0, 1]$ is the *learning rate* (shrinkage
        parameter).

The key connection is that the pseudo-residuals
$r_i^{(m)}$ are the negative gradient of the total loss with respect to
the current functional predictions $F_{m-1}(x_i)$. Fitting $h_m$ to
these residuals therefore moves the ensemble prediction in the steepest
descent direction of the loss functional.

**For squared error loss:**  $L(y, \hat y) = \tfrac{1}{2}(y-\hat y)^2$, so
$r_i^{(m)} = y_i - F_{m-1}(x_i)$; the pseudo-residuals are the ordinary
residuals.

**For log-loss (binary classification):**  Writing
$p(x) = \operatorname{sigmoid}(F(x))$ and
$L(y, F) = -y\log p - (1-y)\log(1-p)$, the pseudo-residuals simplify to
$r_i^{(m)} = y_i - p_{m-1}(x_i)$.  In both cases the pseudo-residuals
have the same interpretation: the part of the target that the current
model has not yet explained.

The learning rate $\eta$ controls how aggressively each new tree moves
the ensemble prediction. A small $\eta$ requires more trees $M$ to
achieve the same fit, but the combined effect of many small steps can be
more accurate and more robust to overfitting than a few large ones.

### XGBoost: regularised gradient boosting {#sec:xgboost}

XGBoost (Chen and Guestrin, 2016) extends gradient boosting with a
regularised objective function and an efficient implementation.

At round $m$, after a second-order Taylor expansion of the loss around
$F_{m-1}(x_i)$, the objective for the new tree $f_m$ becomes
\begin{equation}
\tilde{\mathcal{L}}^{(m)} =
  \sum_{i=1}^n \Bigl[ g_i\, w_{q(x_i)} + \tfrac{1}{2}(h_i + \lambda)\, w_{q(x_i)}^2 \Bigr]
  + \gamma T
(\#eq:xgboost-obj)
\end{equation}
where

-   $g_i = \partial_{F} L(y_i, F(x_i))\big|_{F_{m-1}}$ and
    $h_i = \partial^2_{F} L(y_i, F(x_i))\big|_{F_{m-1}}$ are the
    first and second derivatives of the loss for observation $i$;
-   $q : \mathbb{R}^p \to \{1, \ldots, T\}$ maps each observation to
    its leaf index in the current tree;
-   $w_t$ is the score (weight) assigned to leaf $t$;
-   $T$ is the number of leaves;
-   $\lambda \geq 0$ is an L2 regularisation penalty on leaf weights;
-   $\gamma \geq 0$ is a penalty on tree complexity (number of leaves).

For a fixed tree structure, the objective \@ref(eq:xgboost-obj) is
quadratic in the leaf weights. Minimising over $w_t$ for each leaf $t$ gives the
*optimal leaf weight*
\begin{equation}
w_t^* = -\frac{G_t}{H_t + \lambda},
\qquad
G_t = \sum_{i \in \text{leaf}\,t} g_i, \quad
H_t = \sum_{i \in \text{leaf}\,t} h_i.
(\#eq:xgboost-leafwt)
\end{equation}
Substituting back yields the *gain* from adding a split at node $t$,
which XGBoost uses to select the best split at each node in place of
the usual impurity criteria.

The practical implication of $\lambda$ and $\gamma$ is that they shrink
and prune the trees, acting as a form of regularisation that reduces
overfitting. In combination with the learning rate $\eta$, row and
column subsampling, and tree depth limits, XGBoost provides many axes
for controlling the bias-variance trade-off.

**Key hyperparameters and their interpretation:**

| `tidymodels` argument  | Symbol        | Effect                                          |
|:-----------------------|:--------------|:------------------------------------------------|
| `trees`                | $M$           | Number of boosting rounds; more is not always better |
| `tree_depth`           | max depth     | Deeper trees capture interactions but overfit more easily |
| `learn_rate`           | $\eta$        | Shrinks each tree's contribution; smaller requires more trees |
| `mtry`                 | subsample cols| Fraction of columns used per tree; adds diversity |
| `sample_size`          | subsample rows| Fraction of rows used per tree; reduces overfitting |
| `min_n`                | min child wt  | Minimum $H_t$ in a leaf; controls minimum leaf size |
| `loss_reduction`       | $\gamma$      | Minimum gain required to make a split           |

### Implementation in R {#sec:boost-r}

::: {.example}
**Gradient boosting for the Long Beach adoption data**

We compare a default boosted tree with a tuned XGBoost model. The
`bonsai` package is required to use the `lightgbm` engine; here we use
the more portable `xgboost` engine.

```{r boost-fit, message=FALSE}
library(xgboost)

boost_spec <- boost_tree(
  mode       = "classification",
  trees      = tune(),
  tree_depth = tune(),
  learn_rate = tune()
) |>
  set_engine("xgboost")

boost_wf   <- wf |> add_model(boost_spec)

# A small grid covering key hyperparameters
boost_grid <- grid_regular(
  trees(range = c(50L, 300L)),
  tree_depth(range = c(2L, 5L)),
  learn_rate(range = c(-2, -0.5), trans = log10_trans()),
  levels = 3
)

boost_tune <- tune_grid(
  boost_wf, resamples = adopt_folds,
  grid   = boost_grid,
  metrics = metric_set(roc_auc)
)

show_best(boost_tune, metric = "roc_auc", n = 5)
```

```{r boost-final, message=FALSE}
best_boost   <- select_best(boost_tune, metric = "roc_auc")
final_boost  <- finalize_workflow(boost_wf, best_boost) |>
  last_fit(adopt_split, metrics = metric_set(accuracy, roc_auc))

collect_metrics(final_boost)
```

Points to note:

-   `grid_regular()` produces a factorial grid over the three
    hyperparameters. In practice, `grid_latin_hypercube()` or Bayesian
    optimisation via `tune_bayes()` are more efficient for larger
    grids.

-   `last_fit()` refits the finalised workflow on the full training set
    and evaluates on the held-out test set in a single call, respecting
    the original split. This is the correct way to report test
    performance.
:::

## Stacking {#sec:stacking}

### Stacked generalisation

Stacking (Wolpert, 1992) -- also called *blending* or *stacked
generalisation* -- combines predictions from $K$ diverse *base learners*
using a *meta-learner*. The key insight is that different model families
typically make different errors, and a meta-learner trained on their
predictions can allocate weight to the model that is most reliable for
each region of the feature space.

The naïve approach -- train base learners on the full training set,
generate their predictions on the same training set, and train a
meta-learner on these predictions -- is flawed. The base learners will
have overfit to the training data, so their training-set predictions
will be unfairly optimistic, leading the meta-learner to over-weight
complex models.

### Cross-validated meta-features {#sec:cv-meta}

The remedy is to use *cross-validated* predictions as meta-features.
With $V$-fold cross-validation:

1.  Partition the training data $\mathcal{D}_{\text{train}}$ into $V$
    folds $\mathcal{D}_1, \ldots, \mathcal{D}_V$.

2.  For each fold $v = 1, \ldots, V$ and each base learner $k = 1,
    \ldots, K$:
    a.  Train learner $k$ on $\mathcal{D}_{\text{train}} \setminus
        \mathcal{D}_v$.
    b.  Predict fold $v$: $\hat{z}_{ik} = \hat{f}_k(x_i)$ for
        $i \in \mathcal{D}_v$.

3.  Collect the out-of-fold predictions into a meta-feature matrix
    $\hat{\mathbf{Z}} \in \mathbb{R}^{n \times K}$ with rows
    $(\hat{z}_{i1}, \ldots, \hat{z}_{iK})$.

4.  Train a meta-learner $g$ on
    $\{(\hat{\mathbf{z}}_i, y_i)\}_{i=1}^n$.

5.  To predict a new observation $x^*$:
    a.  Refit each base learner on the entire training set.
    b.  Compute base predictions $\hat{z}_k^* = \hat{f}_k(x^*)$.
    c.  Output $g(\hat{z}_1^*, \ldots, \hat{z}_K^*)$.

The meta-learner $g$ is often a regularised linear model (ridge or
lasso regression for regression tasks, logistic regression for
classification). Using out-of-fold predictions ensures that the
meta-learner sees predictions that were computed on held-out data,
removing the bias that would otherwise favour complex base learners.

### Implementation in R {#sec:stacking-r}

The `stacks` package implements cross-validated stacking within the
`tidymodels` framework. Base learners are added to a model *stack*, which
then blend their candidate predictions into a final ensemble.

::: {.example}
**A stacked ensemble for the Long Beach adoption data**

We assemble a stack from three base learners: a logistic regression, a
random forest, and a boosted tree.

```{r stacking-fit, message = FALSE, warning = FALSE}
library(stacks)

ctrl <- control_stack_resamples()

# Logistic regression candidate
lr_spec <- logistic_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet")
lr_res <- wf |>
  add_model(lr_spec) |>
  tune_grid(resamples = adopt_folds,
            grid      = 5,
            metrics   = metric_set(roc_auc),
            control   = ctrl)

# Random forest candidate (fewer trees for speed)
rf_fast_spec <- rand_forest(mode = "classification", trees = 200,
                            mtry = tune()) |>
  set_engine("ranger")
rf_fast_res <- wf |>
  add_model(rf_fast_spec) |>
  tune_grid(resamples = adopt_folds,
            grid      = tibble(mtry = c(2L, 4L)),
            metrics   = metric_set(roc_auc),
            control   = ctrl)

# Boosted tree candidate
bt_spec <- boost_tree(mode = "classification",
                      trees = 200, learn_rate = 0.05,
                      tree_depth = tune()) |>
  set_engine("xgboost")
bt_res <- wf |>
  add_model(bt_spec) |>
  tune_grid(resamples = adopt_folds,
            grid      = tibble(tree_depth = c(2L, 4L)),
            metrics   = metric_set(roc_auc),
            control   = ctrl)

# Build and blend
adopt_stack <- stacks() |>
  add_candidates(lr_res)      |>
  add_candidates(rf_fast_res) |>
  add_candidates(bt_res)      |>
  blend_predictions(penalty = 10^(-2:-1))

autoplot(adopt_stack, type = "weights")
```

```{r stacking-final, message = FALSE}
adopt_ensemble <- adopt_stack |> fit_members()

adopt_test_pred <- adopt_ensemble |>
  predict(new_data = adopt_test) |>
  bind_cols(predict(adopt_ensemble, adopt_test, type = "prob")) |>
  bind_cols(adopt_test |> select(is_adopted))

adopt_test_pred |> roc_auc(truth = is_adopted, .pred_Adopted)
```

Points to note:

-   `control_stack_resamples()` instructs each tuning call to save the
    out-of-fold predictions needed by `stacks`.

-   `blend_predictions()` fits a non-negative lasso regression on the
    stacked predictions. The `penalty` argument controls how many base
    model candidates receive non-zero weight.

-   `autoplot(type = "weights")` shows the blending coefficients, making
    it easy to see which model families contribute most to the stack.

-   `fit_members()` refits only the selected base learners on the full
    training set before deploying.
:::

## AutoML: Automated Model Selection {#sec:automl}

AutoML systems automate the search over model families, preprocessing
steps, and hyperparameters. Tools such as H2O AutoML, FLAML, AutoGluon,
and the `tidymodels` extension `agua` fit dozens of candidates in a
structured search and produce a leaderboard ordered by a chosen metric.

The appeal is clear: AutoML can surface competitive baselines quickly,
comparing methods that might otherwise require many separate modelling
pipelines. In competitions and initial exploration it is a useful
starting point.

However, AutoML carries important risks that are easy to overlook:

-   **Data leakage.** AutoML systems preprocess and transform data
    internally. If preprocessing uses information from left-out
    validation folds (e.g. target-encoding fitted on the whole dataset),
    the leaderboard scores will be optimistic.

-   **Opaque pipelines.** The best AutoML model may involve a
    transformation, an imputation strategy, or a feature interaction
    that is never surfaced to the analyst.

-   **Interpretability loss.** An ensemble of stacked and blended models
    may achieve a fraction of a percent better AUC but be impossible
    to explain to a stakeholder or to audit for fairness.

-   **Deployment complexity.** A full AutoML pipeline may depend on
    many libraries and have complex serialisation requirements.

AutoML is most useful as an *initial benchmark* to understand what
performance is achievable on the task, after which a simpler, more
interpretable model can be justified if it comes close to matching the
benchmark.

## Summary

Ensemble methods improve predictive performance by combining multiple
learners.

1.  **Bagging** reduces variance by averaging many bootstrap-trained
    learners. The variance of the ensemble is $\rho\sigma^2 +
    (1-\rho)\sigma^2/B$, so decorrelating the learners is as important
    as increasing $B$.

2.  **Random forests** add random feature subsets to bagging, further
    decorrelating the trees and enabling reliable variable importance
    measures.

3.  **Gradient boosting** reduces bias sequentially by fitting each new
    weak learner to the pseudo-residuals (negative gradient of the loss)
    of the current ensemble.

4.  **XGBoost** extends gradient boosting with a regularised
    second-order objective, enabling efficient split search and
    protecting against overfitting via leaf-weight shrinkage.

5.  **Stacking** combines diverse model families using a meta-learner
    trained on cross-validated out-of-fold predictions, avoiding the
    bias that would result from using training-set predictions directly.

6.  **AutoML** automates the above choices but does not replace
    careful thinking about leakage, interpretability, and deployment.

Looking ahead, the same workflow ideas from Chapter 5 can be used to
compare these ensembles against simpler baselines, helping us decide
when the extra complexity is genuinely worthwhile.
